---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# deli

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/r-causal/deli/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-causal/deli/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/r-causal/deli/graph/badge.svg)](https://app.codecov.io/gh/r-causal/deli)
<!-- badges: end -->

deli provides M-estimation and empirical sandwich variance estimation in R.

M-estimators express a wide range of statistical procedures as the solution to
a set of estimating equations, and the empirical sandwich estimator supplies
their variance without further derivation. deli offers a general interface for
both custom and pre-built estimating equations, covering basic statistics,
regression, causal inference, survival analysis, measurement error, and
pharmacokinetics.

deli is an R port of the Python
[delicatessen](https://github.com/pzivich/Delicatessen) library. The
[Translating from Python](https://r-causal.github.io/deli/articles/translating-from-python.html)
article maps the Python interface onto deli and documents where the R surface
differs by convention.

## Installation

Install the development version of deli from
[GitHub](https://github.com/r-causal/deli) with:

```{r eval = FALSE}
# install.packages("pak")
pak::pak("r-causal/deli")
```

## Example

The quickest way to fit a model is `m_estimate()`, which takes a formula, a
data frame, and a pre-built estimating equation, then constructs and solves the
estimator in a single call. Here is a linear regression on the `mtcars` data:

```{r example}
library(deli)

fit <- m_estimate(
  mpg ~ wt + hp,
  data = mtcars,
  .ee = ee_regression,
  model = "linear"
)

fit
```

Standard errors come from the sandwich variance estimator, so the usual
accessors report robust inference without additional arguments:

```{r summary}
summary(fit)
```

`vcov()` returns the sandwich variance-covariance matrix and `confint()`
returns Wald confidence intervals. deli also supplies
[broom](https://broom.tidymodels.org) tidiers, `tidy()`, `augment()`, and `glance()`.

## Learn more

The [package website](https://r-causal.github.io/deli/) collects the reference
documentation and articles. Start with
[Getting Started](https://r-causal.github.io/deli/articles/getting-started.html),
then see the articles on
[custom estimating equations](https://r-causal.github.io/deli/articles/custom-estimating-equations.html),
[regression models](https://r-causal.github.io/deli/articles/regression-models.html),
and [causal inference](https://r-causal.github.io/deli/articles/causal-inference.html)
for worked examples.
