---
title: "Causal Inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(deli)
```

## Overview

deli provides estimating equations for several causal inference methods.
A key advantage of the M-estimation framework is that all nuisance model
parameters (e.g., propensity scores, outcome models) are estimated
jointly, so the sandwich variance correctly propagates uncertainty from
all stages.

This vignette demonstrates:

- G-formula (g-computation / standardization)
- Inverse probability weighting (IPW)
- Augmented IPW (AIPW / doubly robust)

## Simulated data

We simulate data with a binary treatment `A`, a confounder `W`, and a
continuous outcome `Y`:

```{r}
set.seed(42)
n <- 1000
W1 <- rnorm(n)
W2 <- rbinom(n, 1, 0.4)
A <- rbinom(n, 1, inverse_logit(-0.5 + 0.5 * W1 + 0.3 * W2))
Y <- 2 + 1.5 * A + W1 - 0.5 * W2 + rnorm(n)

# True ATE = 1.5
```

## G-formula (g-computation)

The g-formula estimates the average causal effect by fitting an outcome
model and standardizing (averaging predictions over the covariate
distribution):

```{r}
X <- cbind(1, A, W1, W2)      # Observed design matrix
X1 <- cbind(1, 1, W1, W2)     # Counterfactual: all treated
X0 <- cbind(1, 0, W1, W2)     # Counterfactual: all untreated

psi <- function(theta) {
  ee_gformula(theta, y = Y, X = X, X1 = X1, X0 = X0)
}

# theta: ACE, E[Y(1)], E[Y(0)], beta0, beta1, beta2, beta3
m <- MEstimator(stacked_equations = psi,
                 init = c(0, 0, 0, 0, 0, 0, 0))
m <- estimate(m)

# ACE estimate (true = 1.5)
m@theta[1]

# With confidence interval
summary(m)
```

The first parameter is the average causal effect (ACE).

## Inverse probability weighting (IPW)

IPW reweights the observed data by the inverse of the treatment
probability (propensity score):

```{r}
W_ps <- cbind(1, W1, W2)  # Propensity score design matrix

psi <- function(theta) {
  ee_ipw(theta, y = Y, A = A, W = W_ps)
}

# theta: ACE, E[Y(1)], E[Y(0)], alpha0, alpha1, alpha2
m <- MEstimator(stacked_equations = psi,
                 init = c(0, 0, 0, 0, 0, 0))
m <- estimate(m)

# ACE
m@theta[1]
summary(m)
```

### Truncating propensity scores

Extreme propensity scores can inflate variance. Use `truncate` to clip
them:

```{r}
psi <- function(theta) {
  ee_ipw(theta, y = Y, A = A, W = W_ps,
         truncate = c(0.1, 0.9))
}

m <- MEstimator(stacked_equations = psi,
                 init = c(0, 0, 0, 0, 0, 0))
m <- estimate(m)
m@theta[1]
```

## Augmented IPW (doubly robust)

AIPW combines outcome modeling and IPW. It is "doubly robust": the
estimate is consistent if either the outcome model or the propensity
score model is correctly specified:

```{r}
X <- cbind(1, A, W1, W2)
X1 <- cbind(1, 1, W1, W2)
X0 <- cbind(1, 0, W1, W2)
W_ps <- cbind(1, W1, W2)

psi <- function(theta) {
  ee_aipw(theta, y = Y, A = A,
          W = W_ps,             # Propensity score model
          X = X, X1 = X1, X0 = X0)  # Outcome model
}

# theta: ACE, E[Y(1)], E[Y(0)], alpha (3), beta (4)
m <- MEstimator(stacked_equations = psi,
                 init = c(0, 0, 0, rep(0, 3), rep(0, 4)))
m <- estimate(m)

# ACE
m@theta[1]
summary(m)
```

## Comparing methods

```{r}
cat("True ATE: 1.5\n")
cat(sprintf("G-formula: %.3f\n", {
  psi <- function(theta) {
    ee_gformula(theta, y = Y, X = cbind(1, A, W1, W2),
                X1 = cbind(1, 1, W1, W2),
                X0 = cbind(1, 0, W1, W2))
  }
  m <- MEstimator(stacked_equations = psi, init = rep(0, 7))
  m <- estimate(m)
  m@theta[1]
}))
cat(sprintf("IPW:       %.3f\n", {
  psi <- function(theta) {
    ee_ipw(theta, y = Y, A = A, W = cbind(1, W1, W2))
  }
  m <- MEstimator(stacked_equations = psi, init = rep(0, 6))
  m <- estimate(m)
  m@theta[1]
}))
cat(sprintf("AIPW:      %.3f\n", {
  psi <- function(theta) {
    ee_aipw(theta, y = Y, A = A, W = cbind(1, W1, W2),
            X = cbind(1, A, W1, W2),
            X1 = cbind(1, 1, W1, W2),
            X0 = cbind(1, 0, W1, W2))
  }
  m <- MEstimator(stacked_equations = psi, init = rep(0, 10))
  m <- estimate(m)
  m@theta[1]
}))
```

## Further reading

- `?ee_ipw_msm`: Marginal structural models with IPW
- `?ee_gestimation_snmm`: G-estimation for structural nested mean models
- `?ee_iv_causal`: Instrumental variable estimation
- `?ee_2sls`: Two-stage least squares
- `?ee_mean_sensitivity_analysis`: Sensitivity analysis for unmeasured
  confounding
