---
title: "Getting Started with deli"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with deli}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

deli is an R package for M-estimation and empirical sandwich variance
estimation. It provides:

1. `m_estimate()`, a one-step fitting function with a formula interface for
   the pre-built estimating equations
2. Pre-built estimating equations for common models (means, regression, causal
   inference, survival analysis, and more)
3. Automatic sandwich variance estimation
4. The `MEstimator` and `GMMEstimator` classes, which solve arbitrary
   user-specified estimating equations

deli is an R port of the Python
[Delicatessen](https://github.com/pzivich/Delicatessen) library.

## Installation

```{r install, eval=FALSE}
# Install from GitHub (development version)
# pak::pak("r-causal/deli")
```

## Fitting your first model

The quickest way to fit a model is `m_estimate()`, which takes a formula, a
data frame, and a pre-built estimating equation. It constructs the estimator
and solves it in a single call, much as `lm()` fits a linear model in one
step. Here is a linear regression on the `mtcars` data:

```{r first-fit}
library(deli)

fit <- m_estimate(
  mpg ~ wt + hp,
  data = mtcars,
  .ee = ee_regression,
  model = "linear"
)

coef(fit)
```

The `.ee` argument names the estimating equation, and any further arguments
(such as `model = "linear"`) are passed on to it. The formula and data supply
the response and design matrix, so you do not need to build them by hand.

## Inspecting the fit

Fitted estimators support the standard R accessors. `coef()` returns the point
estimates, `vcov()` returns the sandwich variance-covariance matrix, and
`confint()` returns Wald confidence intervals:

```{r accessors}
vcov(fit)

confint(fit)
```

`summary()` collects the estimates, standard errors, test statistics, and
confidence intervals into a single table:

```{r fit-summary}
summary(fit)
```

deli also provides [broom](https://broom.tidymodels.org/) tidiers, so results
flow into tidyverse pipelines. `tidy()` returns one row per parameter, and
`glance()` returns a one-row model summary:

```{r tidiers}
generics::tidy(fit, conf.int = TRUE)

generics::glance(fit)
```

If you prefer deli's own inference helpers, `confidence_intervals()`,
`z_scores()`, `p_values()`, and `s_values()` each return the corresponding
quantities directly:

```{r inference-helpers}
z_scores(fit)
p_values(fit)
```

## Other regression models

The pre-built regression equation covers several model families. Switch from
linear to logistic regression by changing the `model` argument. Here `vs` is a
binary indicator of engine type:

```{r logistic-fit}
fit_logistic <- m_estimate(
  vs ~ mpg + wt,
  data = mtcars,
  .ee = ee_regression,
  model = "logistic"
)

summary(fit_logistic)
```

The sandwich variance estimator supplies robust standard errors automatically,
without any additional arguments. `ee_regression()` covers linear, logistic,
and Poisson models. See `?ee_glm` for the wider family of generalized linear
models, and the dedicated functions `?ee_robust_regression`,
`?ee_ridge_regression`, `?ee_lasso_regression`, `?ee_elasticnet_regression`,
`?ee_beta_regression`, and `?ee_tobit` for robust, ridge, LASSO, elastic net,
beta, and tobit regression.

## The estimator objects

`m_estimate()` is a convenience wrapper around the `MEstimator` class. When a
pre-built estimating equation and the formula interface do not fit your
problem, for example when you supply a custom estimating equation, you can work
with the estimator directly. The workflow has two steps:

1. Define a `psi` function that returns the estimating equation contributions
2. Construct an `MEstimator` and call `estimate()` to solve it

Here is a simple example that estimates a mean:

```{r mean-estimator}
y <- c(1, 2, 3, 1, 4, 5, 3, 2, 6, 7)

# Define the estimating equation
psi <- function(theta) {
  ee_mean(theta, y = y)
}

# Construct the estimator and solve
m <- MEstimator(stacked_equations = psi, init = 0) |> estimate()

coef(m)
```

The estimated mean matches `mean(y)`:

```{r mean-check}
mean(y)
```

The same accessors that worked on the formula fit work here. The point
estimate and variance are also available on the object as `m@theta` and
`m@variance`:

```{r estimator-slots}
m@theta       # Point estimate (the mean)
m@variance    # Sandwich variance estimate
```

`m_estimate()` accepts a `psi` function directly as well, which fits the same
model in one step:

```{r psi-one-step}
m <- m_estimate(psi, init = 0)
coef(m)
```

## Stacking estimating equations

A key strength of M-estimation is the ability to stack estimating equations.
When you stack equations, the sandwich variance correctly accounts for all
sources of uncertainty. Here is an example that estimates a mean and variance
simultaneously:

```{r mean-variance}
y <- c(1, 2, 3, 1, 4, 5, 3, 2, 6, 7)

psi <- function(theta) {
  ee_mean_variance(theta, y = y)
}

m <- m_estimate(stacked_equations = psi, init = c(0, 0))
m@theta  # c(mean, variance)
```

For more complex stacking with custom equations, see
`vignette("custom-estimating-equations")`.

## Causal inference: inverse probability weighting

deli includes estimating equations for causal inference. Because they combine
several components, they are a natural fit for the function interface. Here is
an example that uses inverse probability weighting (IPW) to estimate an average
treatment effect:

```{r ipw}
set.seed(42)
n <- 500
w <- rbinom(n, 1, 0.5)               # Binary confounder
A <- rbinom(n, 1, plogis(-0.5 + w))  # Treatment depends on w
Y <- 1 + 2 * A + w + rnorm(n)        # Outcome
W <- cbind(1, w)                     # Propensity score design matrix

psi <- function(theta) {
  ee_ipw(theta, y = Y, A = A, W = W)
}

# theta: ACE, E[Y(1)], E[Y(0)], beta0, beta1
m <- m_estimate(stacked_equations = psi, init = c(0, 0, 0, 0, 0))

# ACE (average causal effect) ~ 2
m@theta[1]
```

## Predictions

After fitting a model, use `regression_predictions()` for predicted values
with confidence intervals. It takes a design matrix rather than a data frame,
so build the prediction design with the same columns in the same order as the
fitted model. Here the formula `y ~ x` gives an intercept followed by `x`:

```{r predictions}
set.seed(42)
n <- 200
x <- rnorm(n)
y <- 1 + 2 * x + rnorm(n)
d <- data.frame(x, y)

m <- m_estimate(y ~ x, data = d, .ee = ee_regression, model = "linear")

# Predict at new values
x_new <- seq(-2, 2, by = 1)
X_new <- cbind(1, x_new)
preds <- regression_predictions(X_new, m@theta, m@variance)
preds
```

## Clustered data

For clustered or grouped data, use `aggregate_efuncs()` inside your `psi`
function to get cluster-robust variance estimates:

```{r clustered}
set.seed(42)
n <- 200
n_groups <- 50
group <- rep(1:n_groups, each = n / n_groups)
group_effect <- rnorm(n_groups, sd = 2)
y <- group_effect[group] + rnorm(n)

# Cluster-robust variance
psi <- function(theta) {
  ef <- ee_mean(theta, y = y)
  aggregate_efuncs(ef, group = group)
}

m <- m_estimate(stacked_equations = psi, init = mean(y))

m@theta
m@variance  # Accounts for within-cluster correlation
```

## Exact differentiation

The delta method needs the derivative (Jacobian) of a transform, and the
sandwich variance needs the derivative of the estimating equations. By default
deli computes these with central finite differences (`deriv_method =
"capprox"`), which introduces a small step-size approximation. Passing
`deriv_method = "exact"` switches to forward-mode automatic differentiation,
which evaluates the derivative in closed form. It returns exact derivatives with
no step size to tune. In Python Delicatessen exact differentiation is the
default only for `delta_method`; the estimators and the sandwich variance
default to a forward finite difference.

The `delta_method()` function accepts `deriv_method` directly, and the survival
prediction helpers `survival_predictions()` and `aft_predictions_function()`
forward it to the delta method. The following fits a logistic model and uses
exact differentiation to obtain the variance of the fitted probability at a
covariate pattern:

```{r exact-delta-method}
set.seed(42)
n <- 500
x <- rnorm(n)
pr <- plogis(0.5 + x)
y <- rbinom(n, 1, pr)
d <- data.frame(x, y)

m <- m_estimate(y ~ x, data = d, .ee = ee_regression, model = "logistic")

# Variance of the predicted probability at x = 1, via the delta method.
# The log-odds at the pattern (intercept = 1, x = 1) is theta[1] + theta[2].
transform <- function(theta) inverse_logit(theta[1] + theta[2])
delta_method(m, transform = transform, deriv_method = "exact")
```

That chunk computes the inverse logit twice with two different functions, and
the difference is deliberate. The simulation uses `stats::plogis()`, because it
is ordinary numeric work on plain doubles. The transform uses deli's
`inverse_logit()`, because `plogis()` cannot be differentiated exactly and would
stop the delta method with an error. The two functions return identical numbers;
only one of them survives `deriv_method = "exact"`. The rest of this section
explains where that line falls.

Automatic differentiation works by carrying a derivative alongside each value
through the computation, so the transform must be built from operations deli
knows how to differentiate. The supported surface covers:

- the arithmetic operators (`+`, `-`, `*`, `/`, `^`) and the comparisons
- the matrix product `%*%`, the transpose `t()`, indexing, and `length()`
- `sum()`, `prod()`, `max()`, `min()`, and `mean()`
- the exponential and logarithmic family (`exp`, `expm1`, `log`, `log1p`,
  `log2`, `log10`, `sqrt`)
- the trigonometric and hyperbolic families (`sin`, `cos`, `tan`, `asin`,
  `acos`, `atan`, `sinh`, `cosh`, `tanh`, `asinh`, `acosh`, `atanh`)
- `abs`, `floor`, `ceiling`, `sign`, `cumsum`, and the
  `lgamma` / `digamma` / `trigamma` family
- deli's own transforms `inverse_logit()`, `logit()`, `standard_normal_cdf()`,
  `standard_normal_pdf()`, `deli_polygamma()`, `deli_digamma()`, and
  `identity_transform()`

Each of the three operator groups has gaps. Under `deriv_method = "exact"` the
following error rather than returning a derivative:

- from the `Math` group, `gamma`, `round`, `signif`, `trunc`, `cospi`, `sinpi`,
  `tanpi`, `cumprod`, `cummax`, and `cummin`
- from the `Summary` group, `range`, `any`, and `all`
- from the `Ops` group, the modulo operators `%%` and `%/%`, the logical
  operators `&` and `|`, and unary `!`

Use the finite-difference default for a transform that needs any of them.

`median()`, `quantile()`, and a trimmed `mean(x, trim)` error too, for a reason
worth separating from the others. Each selects among the order statistics of
its argument, so its derivative follows whichever order statistics the current
values happen to select. That is not the derivative of the population quantity,
so a variance built from it would not be valid. It is the same reason
`ee_percentile()` has no usable bread under any derivative method. The
finite-difference methods do return a number for these, but the number inherits
the same problem.

### The base R functions that stop exact differentiation

Five base R functions that look like natural choices inside a transform do not
work under exact differentiation. Each hands its argument straight to compiled
code that requires a plain number. deli catches the resulting failure and
raises its own error, naming the function that stopped the computation and the
deli function to write instead. deli exports a counterpart for each one:

| Use this deli function   | in place of                    | because base R |
| ------------------------ | ------------------------------ | -------------- |
| `inverse_logit(x)`       | `stats::plogis(x)`             | errors         |
| `logit(p)`               | `stats::qlogis(p)`             | errors         |
| `standard_normal_cdf(x)` | `stats::pnorm(x)`              | errors         |
| `standard_normal_pdf(x)` | `stats::dnorm(x)`              | errors         |
| `deli_polygamma(n, x)`   | `base::psigamma(x, deriv = n)` | errors         |
| `deli_digamma(z)`        | `base::digamma(z)`             | differentiates |
| `identity_transform(v)`  | `base::identity(v)`            | differentiates |

The rule follows from where the code runs. Inside a `psi` function and inside a
transform passed to `delta_method()`, reach for the deli function in the first
column. Everywhere else, including data simulation, post-fit display, and plain
numeric work, the base R function is the better choice because it is the one
your readers already know. The last two rows are exempt: `digamma()` and
`identity()` differentiate exactly, so base R serves in any position. Note also
that the polygamma pair takes its arguments in opposite orders, so the two are
not interchangeable by position alone.

The table lists the five functions with a deli counterpart, but the rule reaches
further. Every distribution function in `stats` fails the same way, `qnorm()`
and `dbinom()` among them. The error names whichever one stopped the
computation, whether or not deli has a replacement for it.

One limitation applies to custom code. The reshaping and binding helpers
`matrix()`, `rbind()`, and `cbind()`, along with `ifelse()`, differentiate
exactly only when they are called from within the package. A user-defined
transform that calls them from the global environment reaches the base R
versions, which do not carry derivatives, so use `deriv_method = "exact"` only
when the transform avoids those helpers, and rely on the finite-difference
default otherwise. Where you would reach for `ifelse()`, write
`ind * yes + (1 - ind) * no`, which is ordinary arithmetic and differentiates
exactly.

Assigning row names does not run into that boundary. deli labels the parameters
from the row names of an estimating function that returns them, and a value that
carries derivatives records the labels as a plain matrix does, so an estimating
function can name its rows and then select from them by name under either
derivative method. The labels take no part in the variance computation, which
reads the derivative components of the labeled value and not its labels.
`rownames<-` hands its work to `dimnames<-`, which is a generic, so deli
registers the setter there and an estimating function can label its rows from
anywhere.

Arithmetic, `%*%`, `t()`, `[`, `c()`, and `mean()` are the operations that
always carry derivatives, because deli registers each as an S3 method and S3
dispatch does not depend on where the call is written. Build a p-by-n return
with `t(X * resid)`, the idiom `ee_regression()` itself uses. Use `c()` to
flatten a matrix-shaped result such as `X %*% theta` down to a vector;
`as.numeric()` cannot do that job under exact differentiation, because it has
to return a plain double and so cannot carry a derivative. It aborts on any
value that carries one rather than silently handing back a zero derivative,
including a single parameter such as `as.numeric(theta[1])`, where the primal
is already a plain double and returning it would have looked harmless.
`as.double()`, `as.integer()`, `as.character()`, `as.complex()`,
`as.matrix()`, and `as.vector()` asked for any atomic mode abort for the same
reason. `as.vector()` with its default `mode = "any"` hands its argument back
unchanged, so it keeps the derivatives. `as.logical()` is the one coercion that
answers, because a logical is a step function of its argument and so has a
derivative of zero almost everywhere; that is what lets the conditional idiom
`ind * yes + (1 - ind) * no` take a condition built from `theta`.

`c()` has one wrinkle worth knowing. R dispatches it on its first argument
alone, so a call that starts with a plain number never reaches deli's method:
write `c(theta[1], 0)` rather than `c(0, theta[1])`. The wrong order fails
loudly rather than returning a zero derivative, so it is not a silent trap.

## Next steps
- `vignette("custom-estimating-equations")`: Write your own estimating
  equations and stack them
- `?m_estimate`: the one-step formula and function interface
- `?ee_regression`: linear, logistic, and Poisson regression
- `?ee_glm`: generalized linear models, plus `?ee_ridge_regression`,
  `?ee_lasso_regression`, and the other penalized and robust regression
  functions
- `?ee_gformula`: G-computation / standardization
- `?ee_aipw`: Augmented inverse probability weighting
- `?ee_aft`: Accelerated failure time models
- `?ee_survival_model`: Parametric survival models
