---
title: "Regression Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(deli)
```

## Overview

deli provides estimating equations for a wide range of regression models.
All regression EEs follow the same pattern: define a `psi` function that
supplies the design matrix and response, then pass it to `m_estimate()`, which
builds the estimator and solves it in one call. The sandwich variance estimator
automatically provides robust standard errors.

## Linear regression

The most basic regression model uses `ee_regression()` with `model = "linear"`:

```{r}
set.seed(42)
n <- 300
x1 <- rnorm(n)
x2 <- rbinom(n, 1, 0.5)
y <- 1 + 2 * x1 - 0.5 * x2 + rnorm(n)
X <- cbind(1, x1, x2)

psi <- function(theta) {
  ee_regression(theta, X = X, y = y, model = "linear")
}

m <- m_estimate(stacked_equations = psi, init = c(0, 0, 0))
summary(m)
```

## Logistic regression

For binary outcomes, use `model = "logistic"`:

```{r}
set.seed(42)
n <- 500
x <- rnorm(n)
y <- rbinom(n, 1, inverse_logit(0.5 + x))
X <- cbind(1, x)

psi <- function(theta) {
  ee_regression(theta, X = X, y = y, model = "logistic")
}

m <- m_estimate(stacked_equations = psi, init = c(0, 0))
summary(m)
```

## Poisson regression

For count data, use `model = "poisson"`:

```{r}
set.seed(42)
n <- 500
x <- rnorm(n)
y <- rpois(n, lambda = exp(0.5 + 0.3 * x))
X <- cbind(1, x)

psi <- function(theta) {
  ee_regression(theta, X = X, y = y, model = "poisson")
}

m <- m_estimate(stacked_equations = psi, init = c(0, 0))
summary(m)
```

## GLM: generalized linear models

`ee_glm()` provides a more flexible interface where you specify the
distribution and link function separately:

```{r}
set.seed(42)
n <- 500
x <- rnorm(n)
y <- rpois(n, lambda = exp(0.5 + 0.3 * x))
X <- cbind(1, x)

psi <- function(theta) {
  ee_glm(theta, X = X, y = y, distribution = "poisson", link = "log")
}

m <- m_estimate(stacked_equations = psi, init = c(0, 0))
m@theta
```

Available distributions: `"normal"`, `"binomial"`, `"poisson"`, `"gamma"`,
`"negative_binomial"`, `"inverse_gaussian"`, and `"tweedie"`.
`"negative_binomial"` estimates an extra dispersion parameter and `"tweedie"`
takes a variance-power `hyperparameter`; see `?ee_glm` for the details.
Available links: `"identity"`, `"log"`, `"logit"`, `"probit"`, `"cauchy"`
(alias `"cauchit"`), `"loglog"`, `"cloglog"`, `"inverse"`, and `"sqrt"`.

## Penalized regression

deli supports several penalized regression methods. These add a penalty term
to the estimating equations.

### Ridge regression

L2 penalty shrinks coefficients toward zero:

```{r}
set.seed(42)
n <- 200
x1 <- rnorm(n)
x2 <- rnorm(n)
y <- 1 + 0.5 * x1 + 0.3 * x2 + rnorm(n)
X <- cbind(1, x1, x2)

psi <- function(theta) {
  ee_ridge_regression(theta, X = X, y = y, model = "linear",
                       penalty = 0.5)
}

m <- m_estimate(stacked_equations = psi, init = c(0, 0, 0))
m@theta
```

### LASSO regression

L1 penalty produces sparse solutions:

```{r}
psi <- function(theta) {
  ee_lasso_regression(theta, X = X, y = y, model = "linear",
                       penalty = 0.1)
}

# Not differentiable, so the sandwich variance should not be trusted here
m <- m_estimate(stacked_equations = psi, init = c(0, 0, 0))
m@theta
```

### Elastic net

Combines L1 and L2 penalties:

```{r}
psi <- function(theta) {
  ee_elasticnet_regression(theta, X = X, y = y, model = "linear",
                            penalty = 0.1, ratio = 0.5)
}

# The L1 half is not differentiable, so again distrust the sandwich variance
m <- m_estimate(stacked_equations = psi, init = c(0, 0, 0))
m@theta
```

## Robust regression

`ee_robust_regression()` replaces the squared loss with a robust loss
function, providing resistance to outliers:

```{r}
set.seed(42)
n <- 200
x <- rnorm(n)
y <- 1 + 2 * x + rnorm(n)
# Add some outliers
y[1:5] <- y[1:5] + 20
X <- cbind(1, x)

# Huber loss with k = 1.345
psi <- function(theta) {
  ee_robust_regression(theta, X = X, y = y, model = "linear",
                        loss = "huber", k = 1.345)
}

# Robust losses can have several roots, so seed the search from least squares
start <- coef(lm(y ~ x))
m <- m_estimate(stacked_equations = psi, init = start)

# Compare with OLS (affected by outliers)
psi_ols <- function(theta) {
  ee_regression(theta, X = X, y = y, model = "linear")
}
m_ols <- m_estimate(stacked_equations = psi_ols, init = c(0, 0))

rbind(robust = m@theta, ols = m_ols@theta)
```

Available loss functions: `"huber"`, `"tukey"`, `"andrew"`, `"hampel"`.

## Weighted regression

All regression EEs support observation weights via the `weights` argument:

```{r}
set.seed(42)
n <- 200
x <- rnorm(n)
y <- 1 + 2 * x + rnorm(n)
X <- cbind(1, x)
w <- runif(n, 0.5, 1.5)

psi <- function(theta) {
  ee_regression(theta, X = X, y = y, model = "linear", weights = w)
}

m <- m_estimate(stacked_equations = psi, init = c(0, 0))
m@theta
```

## Predictions after regression

After fitting any regression model, use `regression_predictions()` for
predicted values with confidence intervals:

```{r}
set.seed(42)
n <- 300
x <- rnorm(n)
y <- 1 + 2 * x + rnorm(n)
X <- cbind(1, x)

psi <- function(theta) {
  ee_regression(theta, X = X, y = y, model = "linear")
}
m <- m_estimate(stacked_equations = psi, init = c(0, 0))

# Predict at new covariate values
x_new <- seq(-2, 2, length.out = 5)
X_new <- cbind(1, x_new)
regression_predictions(X_new, m@theta, m@variance)
```
