---
title: "Regression Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(deli)
```

## Overview

deli provides estimating equations for a wide range of regression models.
All of them are fitted the same way: pass a formula, a data frame, and the
estimating equation to `m_estimate()`, which builds the design matrix,
constructs the estimator, and solves it in one call. Further arguments, such as
`model` or `penalty`, go to the estimating equation. The sandwich variance
estimator automatically provides robust standard errors, and the parameters are
named after the design matrix columns.

Every model shown here can also be fitted through the function interface, where
you write a `psi` function that supplies the design matrix and response
yourself. That form is what you need for a design the formula notation cannot
express, and for custom or stacked equations; see
`vignette("custom-estimating-equations")`.

## Linear regression

The most basic regression model uses `ee_regression()` with `model = "linear"`:

```{r}
set.seed(42)
n <- 300
x1 <- rnorm(n)
x2 <- rbinom(n, 1, 0.5)
y <- 1 + 2 * x1 - 0.5 * x2 + rnorm(n)
d <- data.frame(x1, x2, y)

m <- m_estimate(y ~ x1 + x2, data = d, .ee = ee_regression, model = "linear")
summary(m)
```

## Logistic regression

For binary outcomes, use `model = "logistic"`:

```{r}
set.seed(42)
n <- 500
x <- rnorm(n)
y <- rbinom(n, 1, inverse_logit(0.5 + x))
d <- data.frame(x, y)

m <- m_estimate(y ~ x, data = d, .ee = ee_regression, model = "logistic")
summary(m)
```

## Poisson regression

For count data, use `model = "poisson"`:

```{r}
set.seed(42)
n <- 500
x <- rnorm(n)
y <- rpois(n, lambda = exp(0.5 + 0.3 * x))
d <- data.frame(x, y)

m <- m_estimate(y ~ x, data = d, .ee = ee_regression, model = "poisson")
summary(m)
```

## GLM: generalized linear models

`ee_glm()` provides a more flexible interface where you specify the
distribution and link function separately:

```{r}
set.seed(42)
n <- 500
x <- rnorm(n)
y <- rpois(n, lambda = exp(0.5 + 0.3 * x))
d <- data.frame(x, y)

m <- m_estimate(
  y ~ x,
  data = d,
  .ee = ee_glm,
  distribution = "poisson",
  link = "log"
)
m@theta
```

Available distributions: `"normal"`, `"binomial"`, `"poisson"`, `"gamma"`,
`"negative_binomial"`, `"inverse_gaussian"`, and `"tweedie"`.
`"negative_binomial"` estimates an extra dispersion parameter and `"tweedie"`
takes a variance-power `hyperparameter`; see `?ee_glm` for the details.
Available links: `"identity"`, `"log"`, `"logit"`, `"probit"`, `"cauchy"`
(alias `"cauchit"`), `"loglog"`, `"cloglog"`, `"inverse"`, and `"sqrt"`.

## Penalized regression

deli supports several penalized regression methods. These add a penalty term
to the estimating equations.

### Ridge regression

L2 penalty shrinks coefficients toward zero:

```{r}
set.seed(42)
n <- 200
x1 <- rnorm(n)
x2 <- rnorm(n)
y <- 1 + 0.5 * x1 + 0.3 * x2 + rnorm(n)
d <- data.frame(x1, x2, y)

m <- m_estimate(
  y ~ x1 + x2,
  data = d,
  .ee = ee_ridge_regression,
  model = "linear",
  penalty = 0.5
)
m@theta
```

### LASSO regression

L1 penalty produces sparse solutions:

```{r}
# Not differentiable, so the sandwich variance should not be trusted here
m <- m_estimate(
  y ~ x1 + x2,
  data = d,
  .ee = ee_lasso_regression,
  model = "linear",
  penalty = 0.1
)
m@theta
```

### Elastic net

Combines L1 and L2 penalties:

```{r}
# The L1 half is not differentiable, so again distrust the sandwich variance
m <- m_estimate(
  y ~ x1 + x2,
  data = d,
  .ee = ee_elasticnet_regression,
  model = "linear",
  penalty = 0.1,
  ratio = 0.5
)
m@theta
```

## Robust regression

`ee_robust_regression()` replaces the squared loss with a robust loss
function, providing resistance to outliers:

```{r}
set.seed(42)
n <- 200
x <- rnorm(n)
y <- 1 + 2 * x + rnorm(n)
# Add some outliers
y[1:5] <- y[1:5] + 20
d <- data.frame(x, y)

# Robust losses can have several roots, so seed the search from least squares
start <- coef(lm(y ~ x, data = d))

# Huber loss with k = 1.345
m <- m_estimate(
  y ~ x,
  data = d,
  .ee = ee_robust_regression,
  model = "linear",
  loss = "huber",
  k = 1.345,
  init = start
)

# Compare with OLS (affected by outliers)
m_ols <- m_estimate(y ~ x, data = d, .ee = ee_regression, model = "linear")

rbind(robust = m@theta, ols = m_ols@theta)
```

Available loss functions: `"huber"`, `"tukey"`, `"andrew"`, `"hampel"`.

## Weighted regression

All regression EEs support observation weights via the `weights` argument.
Arguments passed through `m_estimate()` are evaluated in the data frame, so the
column name is enough:

```{r}
set.seed(42)
n <- 200
x <- rnorm(n)
y <- 1 + 2 * x + rnorm(n)
w <- runif(n, 0.5, 1.5)
d <- data.frame(x, y, w)

m <- m_estimate(
  y ~ x,
  data = d,
  .ee = ee_regression,
  model = "linear",
  weights = w
)
m@theta
```

## Predictions after regression

After fitting any regression model, use `regression_predictions()` for
predicted values with confidence intervals. It takes a design matrix rather
than a data frame, so build the prediction design with the same columns in the
same order as the fitted model. Here the formula `y ~ x` gives an intercept
followed by `x`:

```{r}
set.seed(42)
n <- 300
x <- rnorm(n)
y <- 1 + 2 * x + rnorm(n)
d <- data.frame(x, y)

m <- m_estimate(y ~ x, data = d, .ee = ee_regression, model = "linear")

# Predict at new covariate values
x_new <- seq(-2, 2, length.out = 5)
X_new <- cbind(1, x_new)
regression_predictions(X_new, m@theta, m@variance)
```
