---
title: "Agresti & Finlay (2009): Robust Regression"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[Agresti & Finlay (2009): Robust Regression example](https://deli.readthedocs.io/en/latest/Examples/Agresti-Crime.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

This example demonstrates robust regression using data from *Statistical Methods
for the Social Sciences*, 4th Edition by Alan Agresti and Barbara Finlay. Violent
crime is modeled as a function of poverty and the percentage of single parents,
at the US state level. The applied example follows the
[UCLA tutorial](https://stats.oarc.ucla.edu/r/dae/robust-regression/) with data
from the 2005 *Statistical Abstract of the United States*, available
[here](https://users.stat.ufl.edu/~aa/social/data.html).

## Data

The data contain 51 observations (50 US states plus Washington DC). We load
the data and standardize the crime outcome.

```{r}
#| label: data
d <- crime
d$intercept <- 1
d$crime_std <- (d$crime - mean(d$crime)) / sd(d$crime)

X <- as.matrix(d[, c("intercept", "poverty", "single")])
y <- d$crime_std
```

## Linear Regression

We start by modeling the standardized crime rate as a function of poverty and
single parent percentage using ordinary least squares. The estimating equation
for linear regression is

$$
\sum_{i=1}^{n}
\begin{bmatrix}
  (Y_i - \hat{Y}_i) \\
  (Y_i - \hat{Y}_i) P_i \\
  (Y_i - \hat{Y}_i) S_i
\end{bmatrix}
= 0
$$

where $Y$ is the crime rate, $P$ is poverty, $S$ the proportion of single
parents, and $\hat{Y}_i = \theta_0 + \theta_1 P_i + \theta_2 S_i$.

```{r}
#| label: ols
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_regression(theta, X = X, y = y, model = "linear")
  },
  init = c(0, 0, 0)
) |>
  estimate()

# Save OLS estimates for use as initial values in robust regression
ols_theta <- unname(estr@theta)

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

## Robust Regression

Robust regression limits the influence of outliers by applying a function $f$
to the residuals. The resulting estimating equation is

$$
\sum_{i=1}^{n} f_k(Y_i - X_i^T \theta)\, X_i = 0
$$

where $f_k$ is the chosen robust loss function with tuning parameter $k$.
Different choices of $f_k$ impose different constraints on how much any single
observation can shift the solution.

The redescending losses (Andrew's Sine, Ullah, and Hampel) drive the
contribution of large residuals to zero. These estimating equations are
non-convex and can have several roots, so we start each of these fits from
zero and use the Levenberg-Marquardt solver, matching the reference analysis.
For this data the point estimates are stable, but the interval estimates for
Andrew's Sine and Hampel are numerically delicate. Both losses set the score
to exactly zero beyond a hard cut-off, so the derivative that forms the bread
matrix is discontinuous at that cut-off. One observation (Washington DC) is a
pronounced outlier whose Andrew residual lands exactly on the cut-off, where
that derivative is undefined. Exact automatic differentiation and finite
differencing resolve the kink on opposite sides, so they return different but
individually defensible variances. The Andrew and Hampel confidence limits
should therefore be read as approximate and can shift with the derivative
method. Ullah is redescending but smooth, with no hard cut-off, so its bread
derivative is well defined and its interval is stable.

### Huber

The Huber function caps residual contributions at $\pm k$. While $k = 1.345$ is
commonly recommended, we use $k = 1$ here.

```{r}
#| label: huber
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "huber", k = 1)
  },
  init = c(0, 0, 0)
) |>
  estimate()

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Fair

Like Huber, the Fair function is monotonic but tapers continuously rather than
having a sharp hinge point. We set $k = 1$.

```{r}
#| label: fair
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "fair", k = 1)
  },
  init = c(0, 0, 0)
) |>
  estimate()

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Cauchy

The Cauchy function begins to reduce outlier influence beyond a certain point
but does not shrink contributions fully to zero. We set $k = 1$.

```{r}
#| label: cauchy
# Use OLS estimates as starting values. Redescending loss functions have
# multiple roots, so c(0,0,0) can converge to a spurious solution.
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "cauchy", k = 1)
  },
  init = ols_theta
) |>
  estimate()

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Andrew's Sine

Andrew's Sine allows the same maximum influence as Huber at $k$ but varies
smoothly and shrinks outlier contributions to zero at high levels. We set
$k = 1$.

```{r}
#| label: andrew
# The redescending losses are non-convex. Following the reference analysis,
# we start from zero and use the Levenberg-Marquardt solver ("lm"). Andrew's
# score is zero beyond a hard cut-off, so its bread derivative is discontinuous
# there; we take the derivative by exact automatic differentiation to get a
# deterministic variance at that discontinuity.
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "andrew", k = 1)
  },
  init = c(0, 0, 0)
) |>
  estimate(solver = "lm", deriv_method = "exact")

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Welsch

The Welsch function has a similar shape to Andrew's Sine but differs in
steepness at the same value of $k$. We set $k = 1$.

```{r}
#| label: welsch
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "welsch", k = 1)
  },
  init = c(0, 0, 0)
) |>
  estimate()

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Ullah

The Ullah function shares a similar shape to the previous redescending
functions. We set $k = 1$.

```{r}
#| label: ullah
# Start from zero with the Levenberg-Marquardt solver, matching the reference
# analysis, and take the derivative by exact automatic differentiation.
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "ullah", k = 1)
  },
  init = c(0, 0, 0)
) |>
  estimate(solver = "lm", deriv_method = "exact")

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Tukey's Biweight

Tukey's biweight substantially reduces the influence of outlying observations.
Small values of $k$ produce large standard errors, so we use $k = 5$.

```{r}
#| label: tukey
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "tukey", k = 5)
  },
  init = c(0, 0, 0)
) |>
  estimate()

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Hampel

The Hampel function mimics Huber until a set point, then linearly decreases
contributions to zero. Unlike the other functions, it takes three tuning
parameters. In R these are passed as `k = c(a, b, c)` where `a` is the first
hinge, `b` the plateau upper limit, and `c` the zero point.

```{r}
#| label: hampel
# Start from zero with the Levenberg-Marquardt solver, matching the reference
# analysis, and take the derivative by exact automatic differentiation. The
# tuning constants k = c(1/3, 4/3, 2) correspond to the reference parameters
# k = 2, lower = 1/3, upper = 4/3.
estr <- MEstimator(
  stacked_equations = function(theta) {
    # k = c(a, b, c) where a < b < c
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "hampel", k = c(1/3, 4/3, 2))
  },
  init = c(0, 0, 0)
) |>
  estimate(solver = "lm", deriv_method = "exact")

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

## References

Agresti A & Finlay B. (2009). *Statistical Methods for the Social Sciences*,
4th Edition. Pearson.

UCLA Advanced Research Computing. Robust regression.
<https://stats.oarc.ucla.edu/r/dae/robust-regression/>
