---
title: "Agresti & Finlay (2009): Robust Regression"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[Agresti & Finlay (2009): Robust Regression example](https://deli.readthedocs.io/en/latest/Examples/Agresti-Crime.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

This example demonstrates robust regression using data from *Statistical Methods
for the Social Sciences*, 4th Edition by Alan Agresti and Barbara Finlay. Violent
crime is modeled as a function of poverty and the percentage of single parents,
at the US state level. The applied example follows the
[UCLA tutorial](https://stats.oarc.ucla.edu/r/dae/robust-regression/) with data
from the 2005 *Statistical Abstract of the United States*, available
[here](https://users.stat.ufl.edu/~aa/social/data.html).

## Data

The data contain 51 observations (50 US states plus Washington DC). We load
the data and standardize the crime outcome.

```{r}
#| label: data
d <- crime
d$intercept <- 1
d$crime_std <- (d$crime - mean(d$crime)) / sd(d$crime)

X <- as.matrix(d[, c("intercept", "poverty", "single")])
y <- d$crime_std
```

## Linear Regression

We start by modeling the standardized crime rate as a function of poverty and
single parent percentage using ordinary least squares. The estimating equation
for linear regression is

$$
\sum_{i=1}^{n}
\begin{bmatrix}
  (Y_i - \hat{Y}_i) \\
  (Y_i - \hat{Y}_i) P_i \\
  (Y_i - \hat{Y}_i) S_i
\end{bmatrix}
= 0
$$

where $Y$ is the crime rate, $P$ is poverty, $S$ the proportion of single
parents, and $\hat{Y}_i = \theta_0 + \theta_1 P_i + \theta_2 S_i$.

```{r}
#| label: ols
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_regression(theta, X = X, y = y, model = "linear")
  },
  init = c(0, 0, 0)
)

# Save OLS estimates for use as initial values in robust regression
ols_theta <- unname(coef(estr))

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(coef(estr), 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

## Robust Regression

Robust regression limits the influence of outliers by applying a function $f$
to the residuals. The resulting estimating equation is

$$
\sum_{i=1}^{n} f_k(Y_i - X_i^T \theta)\, X_i = 0
$$

where $f_k$ is the chosen robust loss function with tuning parameter $k$.
Different choices of $f_k$ impose different constraints on how much any single
observation can shift the solution.

Huber and Fair are monotone: a large residual contributes at most a fixed
amount, but it never stops contributing. The other six losses are redescending,
driving the contribution of a large residual back toward zero. Andrew's Sine,
Tukey's biweight, and Hampel reach zero at a hard cut-off and hold it there,
while Cauchy, Ullah, and Welsch approach zero smoothly and never quite arrive.

A redescending estimating equation is non-convex and can have several roots, so
the starting values decide which root a solver reaches. Cauchy, Andrew's Sine,
and Hampel do not solve from a zero start on this data, and Ullah solves from
one to a root no other fit here is near, so those four fits begin at the
least-squares estimates above instead. Andrew's Sine, Ullah, and Hampel use the
Levenberg-Marquardt solver, matching the reference analysis, and take the bread
derivative by exact automatic differentiation rather than by finite
differencing.

### Huber

The Huber function caps residual contributions at $\pm k$. While $k = 1.345$ is
commonly recommended, we use $k = 1$ here.

```{r}
#| label: huber
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "huber", k = 1)
  },
  init = c(0, 0, 0)
)

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(coef(estr), 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Fair

Like Huber, the Fair function is monotonic but tapers continuously rather than
having a sharp hinge point. We set $k = 1$.

```{r}
#| label: fair
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "fair", k = 1)
  },
  init = c(0, 0, 0)
)

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(coef(estr), 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Cauchy

The Cauchy function begins to reduce outlier influence beyond a certain point
but does not shrink contributions fully to zero. We set $k = 1$.

```{r}
#| label: cauchy
# Use OLS estimates as starting values. Redescending loss functions have
# multiple roots, and from c(0,0,0) this one reaches none of them: the solver
# runs the intercept out past five million and warns that the estimating
# functions are not solved where it stopped.
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "cauchy", k = 1)
  },
  init = ols_theta
)

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(coef(estr), 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Andrew's Sine

Andrew's Sine allows the same maximum influence as Huber at $k$ but varies
smoothly and shrinks outlier contributions to zero at high levels. We set
$k = 1$.

```{r}
#| label: andrew
# The redescending losses are non-convex, and a zero start does not solve this
# one, so the fit begins at the OLS estimates. The Levenberg-Marquardt solver
# ("lm") follows the reference analysis. Andrew's score is zero beyond a hard
# cut-off, so its bread derivative is discontinuous there; we take the
# derivative by exact automatic differentiation to get a deterministic variance
# at that discontinuity.
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "andrew", k = 1)
  },
  init = ols_theta,
  solver = "lm",
  deriv_method = "exact"
)

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(coef(estr), 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Welsch

The Welsch function has a similar shape to Andrew's Sine but differs in
steepness at the same value of $k$. We set $k = 1$.

```{r}
#| label: welsch
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "welsch", k = 1)
  },
  init = c(0, 0, 0)
)

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(coef(estr), 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Ullah

The Ullah function shares a similar shape to the previous redescending
functions. We set $k = 1$.

```{r}
#| label: ullah
# A zero start does solve this one, but to a root that sits well away from every
# other fit here, so the fit begins at the OLS estimates for the same reason the
# three above do. The Levenberg-Marquardt solver matches the reference analysis,
# and the derivative is taken by exact automatic differentiation.
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "ullah", k = 1)
  },
  init = ols_theta,
  solver = "lm",
  deriv_method = "exact"
)

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(coef(estr), 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Tukey's Biweight

Tukey's biweight substantially reduces the influence of outlying observations.
Small values of $k$ produce large standard errors, so we use $k = 5$.

```{r}
#| label: tukey
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "tukey", k = 5)
  },
  init = c(0, 0, 0)
)

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(coef(estr), 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Hampel

The Hampel function mimics Huber until a set point, then linearly decreases
contributions to zero. Unlike the other functions, it takes three tuning
parameters. In R these are passed as `k = c(a, b, c)` where `a` is the first
hinge, `b` the plateau upper limit, and `c` the zero point.

```{r}
#| label: hampel
# A zero start does not solve this one either, so the fit begins at the OLS
# estimates, with the Levenberg-Marquardt solver matching the reference analysis
# and the derivative taken by exact automatic differentiation. The tuning
# constants k = c(1/3, 4/3, 2) correspond to the reference parameters
# k = 2, lower = 1/3, upper = 4/3.
estr <- m_estimate(
  stacked_equations = function(theta) {
    # k = c(a, b, c) where a < b < c
    ee_robust_regression(theta, X = X, y = y, model = "linear",
                         loss = "hampel", k = c(1/3, 4/3, 2))
  },
  init = ols_theta,
  solver = "lm",
  deriv_method = "exact"
)

data.frame(
  Param = c("Intercept", "Poverty", "Single"),
  Coef = round(coef(estr), 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

## References

Agresti A & Finlay B. (2009). *Statistical Methods for the Social Sciences*,
4th Edition. Pearson.

UCLA Advanced Research Computing. Robust regression.
<https://stats.oarc.ucla.edu/r/dae/robust-regression/>
