---
title: "Bonate (2011): Dose-Response Models"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[Bonate (2011): Pharmacokinetic-Pharmacodynamic Modeling example](https://deli.readthedocs.io/en/latest/Examples/Bonate-Pharmaco.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

Here, we replicate some of the examples described in Bonate (2011). The purpose of this document is to illustrate the versatility of `deli` by applying it to pharmacokinetic modeling. This can easily be done using the built-in estimating equations, as will be shown.

## Chapter 4: Variance Models, Weighting, and Transformations

The first example comes from Chapter 4. Data comes from Table 9 (pg 153) from a study by Byers et al. (1989) on XomaZyme-791 dose on percent change in albumin concentration among 17 patients. Note that this number of patients may be below what is considered sufficient for inference with the sandwich.

The E-max model is described by
$$ R_i = E_0 + (E_{m} - E_0) \frac{D_i}{ED_{50} + D_i} $$
where $D$ is the dose, $R$ is the response, $E_0$ is the response at a dose of zero, $E_m$ is the maximum response, and $ED_{50}$ is the halfway maximal dose. Here, $E_0 = 0$ so the E-max model reduces to
$$ R_i = E_{m} \frac{D_i}{ED_{50} + D_i} $$

The corresponding data is presented below.

```{r}
#| label: byers-data
d <- c(5, 5, 13, 11, 14.5, 6.8, 42.5, 37.5, 25, 38, 40, 26.5, 27.7, 27.4, 45, 61.4, 52.8)
r <- c(-12, -13, -28, -24, -45, -18, -26, -40, -45, -26, -29, -26, -22, -28, -36, -27, -48)
```

In the book, a two parameter E-max model is used. The first parameter is maximum response and the second is the 50% effective dose. The minimum dose will be set as zero (i.e., not estimated) since there is no data on a dose of zero.

### Two-parameter E-max model

Given this dose-response data, it appears that the dose-response begins to asymptote at some point beyond a dose of 10. Therefore, we should select a starting value of $ED_{50}$ below 10. Further, a maximum response value between -30 and -50 seems reasonable, so we will select -40 as the max effect starting value.

```{r}
#| label: emax-2param
# Define estimating equation for 2-parameter E-max (E0 fixed at 0)
psi <- function(theta) {
  lower <- 0
  vals <- c(lower, theta)
  # Drop the first row (E0 equation) since E0 is fixed
  ee_emax(theta = vals, dose = d, response = r)[2:3, , drop = FALSE]
}

# Estimate the parameters
estr <- m_estimate(stacked_equations = psi, init = c(-40, 5))

# View results
coef(estr)
sqrt(diag(vcov(estr)))
confint(estr)
```

These results are close to what is reported in the book (note that a different variance estimator is used).

### Outlier-robust E-max model

While not discussed in the chapter, one might be concerned about outlying response values. Particularly, there are some large responses at low doses. These few response values might be overly influential on the overall E-max parameter estimates. To go beyond the book, we can consider the use of outlier-robust estimating functions. These apply an additional constraint that limits the maximum influence an observation can have to the estimating function. The following code uses the Huber function to limit the influence.

```{r}
#| label: emax-robust
# Define robust estimating equation using Huber loss
psi_robust <- function(theta) {
  lower <- 0
  vals <- c(lower, theta)
  ee_emax(theta = vals, dose = d, response = r,
          loss = "huber", k = 5)[2:3, , drop = FALSE]
}

# Estimate the parameters
estr <- m_estimate(
  stacked_equations = psi_robust,
  init = c(-40, 5),
  solver = "nleqslv"
)

# View results
coef(estr)
sqrt(diag(vcov(estr)))
confint(estr)
```

Here, we observe a minor change in the estimated coefficients. The dose-response curve estimated using the outlier-robust approach did not drop as low as the non-robust method. This is expected given the apparent outliers around doses of 15, 25, and 52.

## Chapter 9: Nonlinear Mixed Effects Models: Case Studies

This chapter presents some case studies of nonlinear mixed effects models. Here, we will review some of the case studies but modify them to use estimating equations instead.

### Pharmacodynamic Modeling of Acetylcholinesterase Inhibition

Data comes from Cutler et al. 1995 and is provided in Table 1 of this chapter (pg 360). Data comes from 12 participants with Zifrosilone dose on inhibition of acetylcholinesterase. Here, multiple measurements were obtained from the same participants over different timings. We will explore the use of E-max models again but with some caveats. First, we load the data set.

```{r}
#| label: cutler-data
d <- cutler1995
```

In this data, participants have multiple measures over time.

#### Ignoring clustering

Let's first ignore the fact that a single unit contributes multiple times to estimation and treat all the observations as if they were independent.

```{r}
#| label: emax-independent
# Define estimating equation for 3-parameter E-max
psi <- function(theta) {
  ee_emax(theta = theta, dose = d$zifro, response = d$acetyl)
}

# Estimate the parameters
estr <- m_estimate(
  stacked_equations = psi,
  init = c(0, 60, 0.5),
  solver = "nleqslv"
)

# View results
coef(estr)
sqrt(diag(vcov(estr)))
confint(estr)
```

#### Accounting for clustering

To account for the fact that observations within a participant are not independent, we use `aggregate_efuncs` to collapse the contributions by unique subject IDs. The dimension of the new estimating functions will be 3-by-12, to reflect the fact that there are only 12 independent observations.

```{r}
#| label: emax-clustered
# Define estimating equation with aggregation by subject
psi_c <- function(theta) {
  ef_emax_i <- ee_emax(theta = theta, dose = d$zifro, response = d$acetyl)
  aggregate_efuncs(ef_emax_i, group = d$subject)
}

# Estimate the parameters
estr <- m_estimate(
  stacked_equations = psi_c,
  init = c(0, 60, 0.5),
  solver = "nleqslv"
)

# View results
coef(estr)
sqrt(diag(vcov(estr)))
confint(estr)
```

As seen here, the point estimates are similar but the variance changes.

#### Finite-sample correction

However, 12 observations is still relatively few so we can further apply a finite-sample correction using `finite_correction`.

```{r}
#| label: emax-finite
# Apply HC1 finite-sample correction
estr <- m_estimate(
  stacked_equations = psi_c,
  init = c(0, 60, 0.5),
  finite_correction = "HC1",
  solver = "nleqslv"
)

# View results
coef(estr)
sqrt(diag(vcov(estr)))
confint(estr)
```

Here, we see the variance increase a bit further (due to the additional correction). This concludes the case study showing how to account for non-independent observations and small sample sizes.

## Chapter 11: Generalized Linear Models and Its Extensions

### Adverse Events Case Study

The first example from Chapter 11 is the *Case Study: Assessing the Relationship Between Drug Concentrations and Adverse Events Using Logistic Regression*. Data comes from Table 2 of the book. In the book, a variety of different models for different adverse events are considered. Here, we only consider nausea (and vomiting) by AUC. For the one observation with a missing AUC value, it is dropped from the data set (same as the book). For ease of examining the coefficients, we will also divide the AUC value by 1000.

First, we load the data set and transform the columns.

```{r}
#| label: bonate-data
d <- bonate_adverse
d <- d[complete.cases(d), ]         # Drop rows with missing values
d$intercept <- 1                    # Adding intercept to data
d$auc <- d$auc / 1000              # Rescaling AUC
d$c_max <- d$c_max / 1000          # Rescaling C_max
```

#### Null model

To begin, we fit a null (intercept-only) logistic regression model using the built-in `ee_glm` estimating equation. For the logistic model, we specify a binomial distribution with the logit link.

```{r}
#| label: logit-null
# Estimating equation for null model
psi <- function(theta) {
  ee_glm(theta = theta,
         X = as.matrix(d[, "intercept", drop = FALSE]),
         y = d$nausea,
         distribution = "binomial",
         link = "logit")
}

# Estimate the parameters
estr_null <- m_estimate(stacked_equations = psi, init = 0)
coef(estr_null)
```

#### Full model

Next we fit a logistic regression model that includes linear terms for all the independent variables in the data set.

```{r}
#| label: logit-full
# Estimating equation for full model
psi <- function(theta) {
  ee_glm(theta = theta,
         X = as.matrix(d[, c("intercept", "auc", "sex", "age", "ps")]),
         y = d$nausea,
         distribution = "binomial",
         link = "logit")
}

# Estimate the parameters
estr_full <- m_estimate(stacked_equations = psi, init = c(0, 0, 0, 0, 0))
coef(estr_full)
```

#### Reduced model

In the book, Bonate performs some variable selection. In general, we would not recommend use of backwards-selection procedures (like those done in the book). Such procedures complicate inference (P-values and confidence intervals after these procedures are no longer valid). For comparison purposes, we estimate the reduced model reported in the book.

```{r}
#| label: logit-reduced
# Estimating equation for reduced model
psi <- function(theta) {
  ee_glm(theta = theta,
         X = as.matrix(d[, c("intercept", "auc", "sex")]),
         y = d$nausea,
         distribution = "binomial",
         link = "logit")
}

# Estimate the parameters
estr_redu <- m_estimate(stacked_equations = psi, init = c(0, 0, 0))
coef(estr_redu)
```

#### Probit model

Two alternative models are considered: a probit regression model and a complementary log-log model. For the probit model, we set the link equal to `probit`.

```{r}
#| label: probit
# Estimating equation for reduced probit model
psi <- function(theta) {
  ee_glm(theta = theta,
         X = as.matrix(d[, c("intercept", "auc", "sex")]),
         y = d$nausea,
         distribution = "binomial",
         link = "probit")
}

# Estimate the parameters
estr_prob <- m_estimate(stacked_equations = psi, init = c(0, 0, 0))
coef(estr_prob)
```

#### Complementary log-log model

Similarly, the complementary log-log model only requires setting the link to `cloglog`.

```{r}
#| label: cloglog
# Estimating equation for reduced C-log-log model
psi <- function(theta) {
  ee_glm(theta = theta,
         X = as.matrix(d[, c("intercept", "auc", "sex")]),
         y = d$nausea,
         distribution = "binomial",
         link = "cloglog")
}

# Estimate the parameters
estr_clog <- m_estimate(stacked_equations = psi, init = c(0, 0, 0))
coef(estr_clog)
```

These point estimates match those reported in Tables 3 and 4 on pages 469 and 470 in the book (note the null model differs slightly, since we dropped the one observation with the missing AUC value to fit this model, but the book does not). These results highlight how `deli` allows one to easily fit a variety of different models.

## References

Bonate PL. (2011). *Pharmacokinetic-Pharmacodynamic Modeling and Simulations*. 2nd Edition. Springer, New York, NY.
