---
title: "Boos & Stefanski (2013): M-Estimation (Estimating Equations)"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[Boos & Stefanski (2013): M-Estimation (Estimating Equations) example](https://deli.readthedocs.io/en/latest/Examples/Boos-Stefanski-Ch7.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

Selected examples from Chapter 7 of Boos & Stefanski (2013), *Essential
Statistical Inference*, demonstrating M-estimation with the empirical sandwich
variance estimator.

## 7.2.2 Mean and Variance

Simultaneous estimation of the mean and variance using stacked estimating
equations.

### Data

```{r}
#| label: data-7.2.2
set.seed(80950841)
n <- 200
y_mv <- rnorm(n, mean = 10, sd = sqrt(2))
x_mv <- rnorm(n, mean = 5, sd = 1)
```

### By-hand estimating equations

```{r}
#| label: mean-var-hand
# By-hand mean and variance EE
psi <- function(theta) {
  rbind(
    matrix(y_mv - theta[1], nrow = 1),            # EE for the mean
    matrix((y_mv - theta[1])^2 - theta[2], nrow = 1) # EE for the variance
  )
}

estr <- MEstimator(stacked_equations = psi, init = c(0, 0)) |> estimate()
estr@theta
confint(estr)
```

### Built-in `ee_mean_variance`

```{r}
#| label: mean-var-builtin
psi <- function(theta) {
  ee_mean_variance(theta, y = y_mv)
}

estr <- MEstimator(stacked_equations = psi, init = c(0, 0)) |> estimate()
estr@theta
confint(estr)
```

## 7.2.3 Ratio Estimator

Estimating the ratio $\theta = E[Y] / E[X]$ via the single estimating equation
$\psi_i(\theta) = Y_i - \theta X_i$.

### Single-equation version

```{r}
#| label: ratio-single
psi <- function(theta) {
  matrix(y_mv - theta[1] * x_mv, nrow = 1)
}

estr <- MEstimator(stacked_equations = psi, init = c(1)) |> estimate()
estr@theta
confint(estr)
```

### Three-equation version

Stacking mean of $Y$, mean of $X$, and the ratio:

```{r}
#| label: ratio-three
psi <- function(theta) {
  rbind(
    matrix(y_mv - theta[1], nrow = 1),             # EE for E[Y]
    matrix(x_mv - theta[2], nrow = 1),             # EE for E[X]
    matrix(rep(theta[1] - theta[3] * theta[2], n), nrow = 1) # ratio constraint
  )
}

estr <- MEstimator(stacked_equations = psi, init = c(10, 5, 2)) |> estimate()
estr@theta
confint(estr)
```

## 7.2.4 Delta Method via Stacking

Rather than applying the delta method after estimation, additional rows can be
stacked to propagate uncertainty through transformations. Here we add
$\sqrt{\sigma^2}$ and $\log(\sigma^2)$ as additional parameters.

```{r}
#| label: delta
psi <- function(theta) {
  rbind(
    matrix(y_mv - theta[1], nrow = 1),                      # mean
    matrix((y_mv - theta[1])^2 - theta[2], nrow = 1),       # variance
    matrix(rep(theta[3] - sqrt(theta[2]), n), nrow = 1),     # sqrt(var)
    matrix(rep(theta[4] - log(theta[2]), n), nrow = 1)       # log(var)
  )
}

estr <- MEstimator(stacked_equations = psi, init = c(10, 2, 1, 1)) |>
  estimate()
estr@theta
confint(estr)
```

## 7.2.6 Instrumental Variable

Correcting for measurement error using an instrumental variable. Consider the
model $Y = \alpha + \beta X^* + \epsilon$ where $X^*$ is measured with error as
$X = X^* + U$. An instrument $W$ satisfies $\text{Cov}(W, \epsilon) = 0$.

### Data

```{r}
#| label: data-7.2.6
set.seed(12345)
n_iv <- 500
xstar <- rnorm(n_iv, mean = 0, sd = 1)           # true X
e <- rnorm(n_iv, mean = 0, sd = 1)                # outcome error
u <- rnorm(n_iv, mean = 0, sd = 0.5)              # measurement error
w <- xstar + rnorm(n_iv, mean = 0, sd = 0.5)      # instrument
x_iv <- xstar + u                                  # mismeasured X
y_iv <- 1 + 0.5 * xstar + e                        # outcome
```

### IV estimator: two-equation formulation

Using $(1, W)$ as instruments:

```{r}
#| label: iv-two
psi <- function(theta) {
  resid <- y_iv - theta[1] - theta[2] * x_iv
  rbind(
    matrix(resid, nrow = 1),           # instrument: constant
    matrix(resid * w, nrow = 1)        # instrument: W
  )
}

estr <- MEstimator(stacked_equations = psi, init = c(0, 0)) |> estimate()
estr@theta
confint(estr)
```

### IV estimator: four-equation formulation

Including the mean of $X$ and $W$ to allow further inference:

```{r}
#| label: iv-four
psi <- function(theta) {
  resid <- y_iv - theta[1] - theta[2] * x_iv
  rbind(
    matrix(resid, nrow = 1),              # instrument: constant
    matrix(resid * w, nrow = 1),          # instrument: W
    matrix(x_iv - theta[3], nrow = 1),    # mean of X
    matrix(w - theta[4], nrow = 1)        # mean of W
  )
}

estr <- MEstimator(stacked_equations = psi, init = c(0, 0, 0, 0)) |>
  estimate()
estr@theta
confint(estr)
```

## 7.4.1 Robust Location (Huber)

The Huber loss function down-weights large residuals using a tuning constant $k$.
The estimating equation is $\psi_i(\theta) = h_k(Y_i - \theta)$ where $h_k$ is
the Huber function.

### Data

```{r}
#| label: data-7.4.1
set.seed(42)
y_rob <- rnorm(250)
```

### By-hand Huber robust mean

```{r}
#| label: robust-mean-hand
k <- 3

psi <- function(theta) {
  resid <- y_rob - theta[1]
  # Huber loss: clip residuals to [-k, k]
  clipped <- pmax(pmin(resid, k), -k)
  matrix(clipped, nrow = 1)
}

estr <- MEstimator(stacked_equations = psi, init = c(0)) |> estimate()
estr@theta
confint(estr)
```

### Built-in `ee_mean_robust`

```{r}
#| label: robust-mean-builtin
psi <- function(theta) {
  ee_mean_robust(theta, y = y_rob, k = 3, loss = "huber")
}

estr <- MEstimator(stacked_equations = psi, init = c(0)) |> estimate()
estr@theta
confint(estr)
```

## 7.5.1 Linear Regression

Linear regression $Y_i = \beta_0 + \beta_1 X_i + \beta_2 Z_i + \epsilon_i$
with the estimating equation $\psi_i(\beta) = (Y_i - X_i^T \beta) X_i$.

### Data

```{r}
#| label: data-7.5.1
set.seed(99)
n_lm <- 500
x_lm <- rnorm(n_lm)
z_lm <- rbinom(n_lm, 1, 0.5)
y_lm <- 0.5 + 2 * x_lm - z_lm + rnorm(n_lm)
X_lm <- cbind(1, x_lm, z_lm)
```

### By-hand linear regression

```{r}
#| label: lm-hand
psi <- function(theta) {
  yhat <- X_lm %*% theta                 # predicted values (n-by-1)
  resid <- y_lm - yhat                   # residuals (n-by-1)
  t(X_lm * as.numeric(resid))            # score: p-by-n
}

estr <- MEstimator(stacked_equations = psi, init = c(0, 0, 0)) |> estimate()
estr@theta
confint(estr)
```

### Built-in `ee_regression`

```{r}
#| label: lm-builtin
psi <- function(theta) {
  ee_regression(theta, X = X_lm, y = y_lm, model = "linear")
}

estr <- MEstimator(stacked_equations = psi, init = c(0, 0, 0)) |> estimate()
estr@theta
confint(estr)
```

## 7.5.4 Robust Regression

Huber robust regression with tuning constant $k = 1.345$. The estimating
equation is $\psi_i(\beta) = h_k(Y_i - X_i^T \beta) X_i$.

### By-hand robust regression

```{r}
#| label: robust-reg-hand
k_rr <- 1.345

psi <- function(theta) {
  yhat <- X_lm %*% theta
  resid <- y_lm - as.numeric(yhat)
  # Huber loss applied to residuals
  clipped <- pmax(pmin(resid, k_rr), -k_rr)
  t(X_lm * clipped)
}

estr <- MEstimator(stacked_equations = psi, init = c(0, 0, 0)) |>
  estimate(solver = "nleqslv")
estr@theta
confint(estr)
```

### Built-in `ee_robust_regression`

```{r}
#| label: robust-reg-builtin
psi <- function(theta) {
  ee_robust_regression(theta, X = X_lm, y = y_lm,
                       model = "linear", k = 1.345, loss = "huber")
}

estr <- MEstimator(stacked_equations = psi, init = c(0, 0, 0)) |>
  estimate(solver = "nleqslv")
estr@theta
confint(estr)
```

## 7.5.5 Generalized Linear Models

Using `ee_glm` with different distribution/link combinations to estimate common
epidemiologic effect measures from binary data.

### Data

```{r}
#| label: data-7.5.5
set.seed(321)
n_glm <- 500
x_glm <- rnorm(n_glm)
pr <- inverse_logit(-0.5 + x_glm)
y_glm <- rbinom(n_glm, 1, pr)
X_glm <- cbind(1, x_glm)
```

### Logistic regression (odds ratio)

```{r}
#| label: glm-logit
psi <- function(theta) {
  ee_glm(theta, X = X_glm, y = y_glm,
         distribution = "binomial", link = "logit")
}

estr <- MEstimator(stacked_equations = psi, init = c(0, 0)) |> estimate()
estr@theta
confint(estr)
```

### Log-binomial regression (risk ratio)

```{r}
#| label: glm-log
psi <- function(theta) {
  ee_glm(theta, X = X_glm, y = y_glm,
         distribution = "binomial", link = "log")
}

estr <- MEstimator(stacked_equations = psi, init = c(-1, 0)) |> estimate()
estr@theta
confint(estr)
```

### Identity-binomial regression (risk difference)

```{r}
#| label: glm-identity
psi <- function(theta) {
  ee_glm(theta, X = X_glm, y = y_glm,
         distribution = "binomial", link = "identity")
}

estr <- MEstimator(stacked_equations = psi, init = c(0.3, 0)) |> estimate()
estr@theta
confint(estr)
```
