---
title: "High School and Beyond (1982): Clustered Data"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[High School and Beyond (1982): Clustered Data example](https://deli.readthedocs.io/en/latest/Examples/ClusteredData.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

This example demonstrates how to handle clustered data with M-estimation using
the High School and Beyond (1982) dataset. The HSB82 data contains math
achievement scores for 7,185 students nested within 160 schools. Because
students within the same school are not independent, standard variance estimates
from the empirical sandwich will understate uncertainty. The `aggregate_efuncs()`
utility addresses this by collapsing individual-level estimating function
contributions to the group (school) level before computing the sandwich variance.

## Data

The dataset records math achievement (`MathAch`), sex (`Sex`), socioeconomic
status (`SES`), the school mean of socioeconomic status (`MEANSES`), and school
ID (`School`) for students surveyed in 1982. We create a binary `female`
indicator, a within-school centered socioeconomic status (`cses`), and an
intercept column for use in the design matrix.

```{r}
#| label: data
# Load from the nlme package (the High School and Beyond math achievement data)
d <- as.data.frame(nlme::MathAchieve)

# Prepare columns
d$intercept <- 1
d$female <- ifelse(d$Sex == "Female", 0, 1)
d$cses <- d$SES - d$MEANSES

# Extract arrays for the estimating equations
y <- d$MathAch
X <- as.matrix(d[, c("intercept", "female", "cses")])
g <- d$School

c(students = nrow(d), schools = length(unique(g)))
```

## Example 1: Ignoring Clustering

We start by fitting a simple linear regression of math achievement on sex and
centered SES, ignoring the school-level clustering entirely. The model is:

$$\text{MathAch}_i = \beta_0 + \beta_1 \cdot \text{female}_i + \beta_2 \cdot \text{cses}_i + \epsilon_i$$

The standard empirical sandwich variance treats all 7,185 observations as
independent units. This will generally produce standard errors that are too
small when the data are clustered.

```{r}
#| label: ex1
# Fit linear regression ignoring clustering
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_regression(theta, X = X, y = y, model = "linear")
  },
  init = c(10, 0, 0)
) |>
  estimate()

# Results
data.frame(
  Param = c("Intercept", "Female", "Centered SES"),
  Coef = round(estr@theta, 4),
  LCL = round(confint(estr)[, 1], 4),
  UCL = round(confint(estr)[, 2], 4)
)
```

## Example 2: Accounting for Clustering by School

When observations are clustered, the empirical sandwich variance estimator
should sum estimating function contributions within clusters rather than across
individual observations. The `aggregate_efuncs()` function does exactly this: it
takes the $p \times n$ matrix of individual contributions and collapses it into
a $p \times m$ matrix by summing within each group, where $m$ is the number of
clusters.

This changes the effective sample size from $n = 7{,}185$ students to
$m = 160$ schools. The point estimates (theta) remain the same, but the
standard errors and confidence intervals will be wider, correctly reflecting
the within-school correlation.

The key idea is straightforward: wrap the output of `ee_regression()` with
`aggregate_efuncs()` before returning from the estimating equation function.

```{r}
#| label: ex2
# Fit linear regression accounting for school-level clustering
estr_cl <- MEstimator(
  stacked_equations = function(theta) {
    # Compute individual-level estimating functions
    ef <- ee_regression(theta, X = X, y = y, model = "linear")
    # Aggregate to school level
    aggregate_efuncs(ef, group = g)
  },
  init = c(10, 0, 0)
) |>
  estimate()

# Results
data.frame(
  Param = c("Intercept", "Female", "Centered SES"),
  Coef = round(estr_cl@theta, 4),
  LCL = round(confint(estr_cl)[, 1], 4),
  UCL = round(confint(estr_cl)[, 2], 4)
)
```

## Comparison

The point estimates are identical between the two approaches; clustering
affects only the variance estimation, not the parameter estimates themselves.
However, the confidence intervals from the clustered analysis are wider,
reflecting the loss of information due to within-school correlation. In
particular, the intercept (which varies substantially across schools) shows the
largest increase in standard error.

This pattern is typical: ignoring clustering leads to overconfidence in the
estimates. The `aggregate_efuncs()` utility provides a simple, general-purpose
correction that works with any estimating equation, not just regression.

## References

Raudenbush SW & Bryk AS. (2002). *Hierarchical Linear Models: Applications and
Data Analysis Methods* (2nd ed.). Sage Publications.

Ross RK, Zivich PN, & Cole SR. (2024). M-estimation for common
epidemiological measures: introduction and applied examples. *International
Journal of Epidemiology*, 53(2), dyae030.
