---
title: "Coffman (2021): Mediation Analysis"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[Coffman et al. (2021): Causal Mediation Analysis example](https://deli.readthedocs.io/en/latest/Examples/Coffman-2021.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

For the `twangMediation` package in R, Coffman et al. provided a vignette on
causal mediation analysis in the context of health disparity research.
Specifically, the vignette examined possible mediating pathways for substance
use disparities among sexual minority women. Data for this analysis comes from
the National Survey of Drug Use and Health (NSDUH) and is available from the R
package.

Drawing from this vignette, we show how `deli` can be used to conduct causal
mediation analyses. Note that the provided data consists of more than 40,000
observations, so fitting these models might take a second.

## Data

In the context of the vignette, interest was in the disparities of adult
smoking status (`cigmon`) by sexual minority (e.g., lesbian, gay, bisexual;
`lgb_flag`) as mediated by early smoking initiation (prior to age 15; `cig15`).
The adjustment set is age (`age`), race (`race`), education (`educ`), income
(`income`), and employment (`employ`).

```{r}
#| label: data
d <- nsduh
```

## Average Causal Effect

To start, we estimate the average causal effect of sexual minority status on
adult smoking status using an inverse probability of treatment weighting
estimator. The weights are defined as

$$ IPTW = \frac{A_i}{\Pr(A_i = 1 \mid W_i)} + \frac{1 - A_i}{1 - \Pr(A_i = 1 \mid W_i)} $$

and then can be used to compute the weighted mean. Here, we use the Hajek
estimator due to its improved finite-sample performance relative to the
Horvitz-Thompson estimator.

```{r}
#| label: ace-data
# Design matrix with categorical variables
W_matrix <- model.matrix(
  ~ factor(age) + factor(race) + factor(educ) + factor(income) + factor(employ),
  data = d
)

a <- d$lgb_flag
y <- d$cigmon
```

```{r}
#| label: ace
psi_ace <- function(theta) {
  ace <- theta[1]
  mu1 <- theta[2]
  mu0 <- theta[3]
  alpha <- theta[4:length(theta)]

  # Propensity score model
  ee_ps <- ee_regression(theta = alpha, X = W_matrix, y = a, model = "logistic")
  ps <- inverse_logit(as.numeric(W_matrix %*% alpha))
  iptw <- a / ps + (1 - a) / (1 - ps)

  # Hajek estimator for the causal risks
  ee_r1 <- matrix(iptw * a * (y - mu1), nrow = 1)
  ee_r0 <- matrix(iptw * (1 - a) * (y - mu0), nrow = 1)
  ee_ace <- matrix(rep(mu1 - mu0 - ace, length(y)), nrow = 1)

  rbind(ee_ace, ee_r1, ee_r0, ee_ps)
}

init_vals <- c(0, 0.5, 0.5, rep(0, ncol(W_matrix)))
estr <- MEstimator(stacked_equations = psi_ace, init = init_vals) |>
  estimate()

c(ace = coef(estr)[[1]], confint(estr)[1, ])
```

These results suggest a disparity by sexual minority status in adult smoking
behaviors that is not explained by the adjustment set. These results are
similar to `twangMediation` but do differ as a result of how the propensity
scores are being estimated. In the vignette, XGBoost (an ensemble of decision
trees) is used. Here, a simple parametric model is used instead. It would be
better practice to specify a more flexible model for the propensity score
(e.g., interactions).

While it may seem like `twangMediation` has an advantage here, the
corresponding statistical inference is invalid as XGBoost is not generally
valid without the use of cross-fitting and doubly robust estimators. See Zivich
& Breskin (2021) for details on why the `twangMediation` approach is not
statistically valid. The approach based on parametric models used here with
`deli` is statistically valid.

## Controlled Direct Effect

The controlled direct effect is one parameter we can consider in mediation
analysis. Letting $Y(a,m)$ denote the potential outcome under action $a$ and
mediator $m$, the controlled direct effect is

$$ E[Y(a,m)] - E[Y(a',m)] $$

so we are considering the effect of changing $a$ when $m$ is held at a fixed
value.

For the controlled direct effect, we need to estimate a pair of weights
(Coffman & Zhong 2012). First, we estimate the inverse probability of treatment
weights as before. The second weight needed is the inverse probability of
mediator weights, defined as the following

$$ IPMW = \frac{M_i}{\Pr(M_i = 1 \mid A_i, W_i)} + \frac{1 - M_i}{1 - \Pr(M_i = 1 \mid A_i, W_i)} $$

which we can jointly estimate all parameters using a stacked extension of the
previous estimating functions with an additional logistic regression model.
Here, the same adjustment set is used for both nuisance models. However, it can
be the case that there are additional confounding variables for mediators.
These can be dealt with by extending the conditional probability for $IPMW$.

For estimation, we again use the Hajek estimator with the *product* of the two
weights.

```{r}
#| label: cde-data
# W_matrix is defined above (adjustment set only)
# X_matrix includes treatment in addition to adjustment set
X_matrix <- model.matrix(
  ~ lgb_flag + factor(age) + factor(race) + factor(educ) + factor(income) + factor(employ),
  data = d
)

a <- d$lgb_flag
m <- d$cig15
y <- d$cigmon
```

```{r}
#| label: cde
psi_cde <- function(theta) {
  ndim_W <- ncol(W_matrix)
  ace1 <- theta[1]
  ace0 <- theta[2]
  mu11 <- theta[3]
  mu01 <- theta[4]
  mu10 <- theta[5]
  mu00 <- theta[6]
  alpha <- theta[7:(6 + ndim_W)]
  gamma <- theta[(6 + ndim_W + 1):length(theta)]

  # Propensity score model for A
  ee_ps <- ee_regression(theta = alpha, X = W_matrix, y = a, model = "logistic")
  ps <- inverse_logit(as.numeric(W_matrix %*% alpha))
  iptw <- a / ps + (1 - a) / (1 - ps)

  # Mediation score model for M
  ee_ms <- ee_regression(theta = gamma, X = X_matrix, y = m, model = "logistic")
  ms <- inverse_logit(as.numeric(X_matrix %*% gamma))
  ipmw <- m / ms + (1 - m) / (1 - ms)

  # Hajek estimator for the causal risks
  ipw <- iptw * ipmw
  n <- length(y)
  ee_r11 <- matrix(ipw * a * m * (y - mu11), nrow = 1)
  ee_r01 <- matrix(ipw * (1 - a) * m * (y - mu01), nrow = 1)
  ee_r10 <- matrix(ipw * a * (1 - m) * (y - mu10), nrow = 1)
  ee_r00 <- matrix(ipw * (1 - a) * (1 - m) * (y - mu00), nrow = 1)
  ee_ace1 <- matrix(rep(mu11 - mu01 - ace1, n), nrow = 1)
  ee_ace0 <- matrix(rep(mu10 - mu00 - ace0, n), nrow = 1)

  rbind(ee_ace1, ee_ace0, ee_r11, ee_r01, ee_r10, ee_r00, ee_ps, ee_ms)
}

init_vals <- c(0, 0, 0.5, 0.5, 0.5, 0.5,
               rep(0, ncol(W_matrix)),
               rep(0, ncol(X_matrix)))
estr <- MEstimator(stacked_equations = psi_cde, init = init_vals) |>
  estimate()

cde <- cbind(estimate = coef(estr), confint(estr))[1:2, ]
rownames(cde) <- c("CDE(1)", "CDE(0)")
cde
```

Note that the `twangMediation` documentation does not list the controlled
direct effects, so no comparison can be made here. However, these numbers do
happen to be similar to the *pure* direct effect estimates, which we will
examine in the next section.

As we have estimated a pair of parameters (two controlled direct effects),
confidence bands are more appropriate to report here (since they provide
*simultaneous* coverage for both parameters). sup-t confidence bands can be
easily computed with the following functionality.

```{r}
#| label: cde-bands
cb <- confidence_bands(estr, subset = c(1, 2), seed = 4281014)
rownames(cb) <- c("CDE(1)", "CDE(0)")
cb
```

As shown here, these are slightly wider than the confidence intervals. This is
to be expected, as we are interested in multiple parameters but still would
like to have simultaneous coverage. However, the sup-t correction is not overly
conservative like the Bonferroni correction.

## Pure Direct Effect

The next parameter we could be interested in is the pure direct effect (also
known as the natural direct effect). Using the potential outcomes from before,
the pure direct effect is

$$ E[Y(a, M(a'))] -  E[Y(a', M(a'))] $$

where $M(a)$ is the potential mediator under $a$. Note that there is the
seemingly odd combination of $Y$ under $a$ but the mediator is the one under
$a'$. Such a pattern does not occur in nature. Identification of the pure
direct effect thus relies on a 'cross-world' assumption for identification.
This assumption is what makes these effects 'unnatural' (to me).

For the right quantity, note that $E[Y(a')]$. Thus, that part can be estimated
using the IPW estimator for the never-act part of the average causal effect
estimator. For the left quantity, we use 'strategy 3' of Tchetgen Tchetgen &
Shpitser (2012). The described IPW estimator is

$$ n^{-1} \sum_{i=1}^{n} Y_i \frac{I(A_i = 1)}{\Pr(A=1 \mid W_i)} \times \frac{\Pr(M_i = m \mid A = 0, W_i)}{\Pr(M_i = m \mid A_i, W_i)} $$

where the last probability fractions are the probability of the *observed*
value of $M$ for that individual.

```{r}
#| label: pde-data
# W_matrix, X_matrix defined above
# Create counterfactual design matrix with lgb_flag set to 0
d0 <- d
d0$lgb_flag <- 0
X0_matrix <- model.matrix(
  ~ lgb_flag + factor(age) + factor(race) + factor(educ) + factor(income) + factor(employ),
  data = d0
)

a <- d$lgb_flag
m <- d$cig15
y <- d$cigmon
```

```{r}
#| label: pde
psi_pde <- function(theta) {
  ndim_W <- ncol(W_matrix)
  pde <- theta[1]
  mu1m <- theta[2]
  mu0 <- theta[3]
  alpha <- theta[4:(3 + ndim_W)]
  gamma <- theta[(3 + ndim_W + 1):length(theta)]

  # Propensity score model for A
  ee_ps <- ee_regression(theta = alpha, X = W_matrix, y = a, model = "logistic")
  ps <- inverse_logit(as.numeric(W_matrix %*% alpha))
  iptw <- a / ps + (1 - a) / (1 - ps)

  # Mediation score model for M
  ee_ms <- ee_regression(theta = gamma, X = X_matrix, y = m, model = "logistic")
  ms <- inverse_logit(as.numeric(X_matrix %*% gamma))
  m0s <- inverse_logit(as.numeric(X0_matrix %*% gamma))
  ipmw <- m * (m0s / ms) + (1 - m) * (1 - m0s) / (1 - ms)

  # Hajek estimator for the causal risks
  n <- length(y)
  ee_r1m <- matrix(iptw * a * ipmw * (y - mu1m), nrow = 1)
  ee_r0 <- matrix(iptw * (1 - a) * (y - mu0), nrow = 1)
  ee_pde <- matrix(rep(mu1m - mu0 - pde, n), nrow = 1)

  rbind(ee_pde, ee_r1m, ee_r0, ee_ps, ee_ms)
}

init_vals <- c(0, 0.5, 0.5,
               rep(0, ncol(W_matrix)),
               rep(0, ncol(X_matrix)))
estr <- MEstimator(stacked_equations = psi_pde, init = init_vals) |>
  estimate()

c(pde = coef(estr)[[1]], confint(estr)[1, ])
```

These results are fairly similar to those provided in `twangMediation` (but the
results here are not stratified and we are using a different nuisance model).

## Pure Indirect Effect

The pure (natural) indirect effect can be defined as

$$ E[Y(a, M(a))] -  E[Y(a, M(a'))] $$

which again involves a cross-world assumption. Note the second part here is the
same as the left side of the pure direct effect. This shows how to decompose
everything. The following code implements the estimator for the pure indirect
effect.

```{r}
#| label: pie
psi_pie <- function(theta) {
  ndim_W <- ncol(W_matrix)
  pie <- theta[1]
  mu1 <- theta[2]
  mu1m <- theta[3]
  alpha <- theta[4:(3 + ndim_W)]
  gamma <- theta[(3 + ndim_W + 1):length(theta)]

  # Propensity score model for A
  ee_ps <- ee_regression(theta = alpha, X = W_matrix, y = a, model = "logistic")
  ps <- inverse_logit(as.numeric(W_matrix %*% alpha))
  iptw <- a / ps + (1 - a) / (1 - ps)

  # Mediation score model for M
  ee_ms <- ee_regression(theta = gamma, X = X_matrix, y = m, model = "logistic")
  ms <- inverse_logit(as.numeric(X_matrix %*% gamma))
  m0s <- inverse_logit(as.numeric(X0_matrix %*% gamma))
  ipmw <- m * (m0s / ms) + (1 - m) * (1 - m0s) / (1 - ms)

  # Hajek estimator for the causal risks
  n <- length(y)
  ee_r1 <- matrix(iptw * a * (y - mu1), nrow = 1)
  ee_r1m <- matrix(iptw * a * ipmw * (y - mu1m), nrow = 1)
  ee_pie <- matrix(rep(mu1 - mu1m - pie, n), nrow = 1)

  rbind(ee_pie, ee_r1, ee_r1m, ee_ps, ee_ms)
}

init_vals <- c(0, 0.5, 0.5,
               rep(0, ncol(W_matrix)),
               rep(0, ncol(X_matrix)))
estr <- MEstimator(stacked_equations = psi_pie, init = init_vals) |>
  estimate()

c(pie = coef(estr)[[1]], confint(estr)[1, ])
```

That provides the pure indirect effect estimate. This is again similar to
`twangMediation` and if we add the pure effects together, we get back to the
average causal effect.

## Pure Effects Altogether

As a final step, we now package the pure direct and indirect effect estimators
together into a single function. As with the controlled direct effect, interest
is in a *pair* of parameters, so confidence bands are more appropriate to
report here.

```{r}
#| label: full
psi_full <- function(theta) {
  ndim_W <- ncol(W_matrix)
  pde <- theta[1]
  pie <- theta[2]
  mu1 <- theta[3]
  mu0 <- theta[4]
  mu1m <- theta[5]
  alpha <- theta[6:(5 + ndim_W)]
  gamma <- theta[(5 + ndim_W + 1):length(theta)]

  # Propensity score model for A
  ee_ps <- ee_regression(theta = alpha, X = W_matrix, y = a, model = "logistic")
  ps <- inverse_logit(as.numeric(W_matrix %*% alpha))
  iptw <- a / ps + (1 - a) / (1 - ps)

  # Mediation score model for M
  ee_ms <- ee_regression(theta = gamma, X = X_matrix, y = m, model = "logistic")
  ms <- inverse_logit(as.numeric(X_matrix %*% gamma))
  m0s <- inverse_logit(as.numeric(X0_matrix %*% gamma))
  ipmw <- m * (m0s / ms) + (1 - m) * (1 - m0s) / (1 - ms)

  # Hajek estimator for the causal risks
  n <- length(y)
  ee_r1 <- matrix(iptw * a * (y - mu1), nrow = 1)
  ee_r0 <- matrix(iptw * (1 - a) * (y - mu0), nrow = 1)
  ee_r1m <- matrix(iptw * a * ipmw * (y - mu1m), nrow = 1)
  ee_pde <- matrix(rep(mu1m - mu0 - pde, n), nrow = 1)
  ee_pie <- matrix(rep(mu1 - mu1m - pie, n), nrow = 1)

  rbind(ee_pde, ee_pie, ee_r1, ee_r0, ee_r1m, ee_ps, ee_ms)
}

init_vals <- c(0, 0, 0.5, 0.5, 0.5,
               rep(0, ncol(W_matrix)),
               rep(0, ncol(X_matrix)))
estr <- MEstimator(stacked_equations = psi_full, init = init_vals) |>
  estimate()
cb <- confidence_bands(estr, subset = c(1, 2), seed = 921453)

pure <- cbind(estimate = coef(estr)[1:2], cb)
rownames(pure) <- c("PDE", "PIE")
pure
```

This provides our results collected from before into a single estimating
function. As a result, we are able to construct sup-t confidence bands for
appropriate inference on the decomposition of the total effect.

We can also check that our decomposition sums back to the average causal
effect.

```{r}
#| label: check-decomposition
estr@theta[1] + estr@theta[2]
```

This matches the average causal effect for many of the decimals (we expect some
minor differences due to floating point precision and the root-finding
procedure tolerance not being infinite).

This example provides another illustration of how to use `deli` to construct
custom estimating equations and use them for statistically valid inference.

## References

Coffman, D. L., & Zhong, W. (2012). Assessing mediation using marginal
structural models in the presence of confounding and moderation. *Psychological
Methods*, 17(4), 642.

Coffman, D. L., Schuler, M. S., McCaffrey, D. F., Castellano, K. E., Zhou, H.,
Vegetabile, B., & Griffin, B. A. (2021). A tutorial for conducting causal
mediation analysis with the twangMediation package.

Tchetgen Tchetgen, E. J., & Shpitser, I. (2012). Semiparametric theory for
causal mediation analysis: efficiency bounds, multiple robustness, and
sensitivity analysis. *Annals of Statistics*, 40(3), 1816.

Zivich, P. N., & Breskin, A. (2021). Machine learning for causal inference: on
the use of cross-fit estimators. *Epidemiology*, 32(3), 393-401.
