---
title: "Cole (2023): Missing Data Methods"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(deli)
```

Cole et al. (2023) reviewed a sensitivity analysis method for missing data
proposed in Robins et al. (1999). This example was illustrated using simulated
missing data on CD4 T cell counts among persons with HIV (an immune marker for
disease progression). In the paper, the estimator and corresponding estimating
functions are reviewed.

Here, we replicate this example using `deli`.

## Load Data

```{r}
#| label: load-data
d <- lau_wihs
d$cd41 <- as.numeric(d$cd41)
d$cd42 <- as.numeric(d$cd42)
d$cd43 <- as.numeric(d$cd43)
d$cd44 <- as.numeric(d$cd44)
```

Note that the last 3 columns are all variations on the extent of the missing
data.

## Complete-Case Analysis

As a starting point, consider a complete-case analysis for the second missing
data scenario.

```{r}
#| label: complete-case-data
# Extract outcome and missingness indicator
y <- d$cd42
r <- ifelse(is.na(y), 0, 1)
```

```{r}
#| label: complete-case
# Define the complete-case mean estimating equation
psi_mean <- function(theta) {
  # Replace NA with arbitrary value; zeroed out by r anyway
  y_filled <- ifelse(is.na(y), -999, y)
  # Complete-case mean EE
  matrix(r * (y_filled - theta), nrow = 1)
}

estr <- MEstimator(stacked_equations = psi_mean, init = c(300)) |> estimate()

c(mu = coef(estr)[[1]], confint(estr)[1, ])
```

However, this complete-case analysis assumes missing data is non-informative
(or "missing completely at random"). We might be suspicious that this is the
case. This is where the sensitivity analysis comes in.

## Sensitivity Analysis

This approach is based on the following expression:
$$ \mu = E \left[ \frac{R_i Y_i}{H[\beta + q(Y; \alpha)]} \right] $$
where $Y$ is CD4, $R$ indicates whether $Y$ was observed, $H$ is a
user-specified monotonic increasing function with outputs bounded between zero
and one, and $q$ is a user-specified bias function (that can depend on $Y$
itself). Here, $\alpha$ is not estimable (the problem of an unknown missing
data mechanism). However, we can plug-in different values for $\alpha$ and see
how that changes our results. Further, we might even have a best guess at a
range for $\alpha$. Here, $\beta$ is a bounding parameter that we will
estimate.

For the following example, $H$ is specified to be the logistic function. We
will also use the following $q(Y; \alpha)$ function:
$$ q(Y; \alpha) = \alpha Y $$

### By-hand estimating equations

First, we define the bias function $q$ and the sensitivity analysis estimating
equations by hand.

```{r}
#| label: q-function
q_function <- function(y_vals, alpha) {
  alpha * ifelse(is.na(y_vals), 0, y_vals)
}
```

```{r}
#| label: sens-byhand
psi_sens <- function(theta) {
  mu_y <- theta[1]
  beta <- theta[2]

  # Solving for the sensitivity analysis mean
  numerator <- r * y                                    # Numerator
  denominator <- inverse_logit(beta + qy)               # Denominator
  ratio <- numerator / denominator                       # Ratio
  ratio <- ifelse(is.na(ratio), 0, ratio)                # Set missing to zero
  mean_ee <- matrix(ratio - mu_y, nrow = 1)              # Mean EE


  # Solving for intercept of model
  h_params <- matrix(r / inverse_logit(beta + qy) - 1,  # Bounding parameter EE
                     nrow = 1)

  # Returning stacked estimating equations
  rbind(mean_ee, h_params)
}

# Set alpha and compute q
alpha <- 0.01
qy <- q_function(y_vals = y, alpha = alpha)

# Estimate
estr <- MEstimator(stacked_equations = psi_sens, init = c(300, 0)) |>
  estimate()

c(mu = coef(estr)[[1]], confint(estr)[1, ])
```

This is notably lower than the complete-case analysis.

### Using the built-in function

We can also implement this approach using the built-in
`ee_mean_sensitivity_analysis` function.

```{r}
#| label: sens-builtin
alpha <- 0.01
qy <- q_function(y_vals = y, alpha = alpha)

psi_robins <- function(theta) {
  ee_mean_sensitivity_analysis(
    theta = theta,
    y = y,
    delta = r,
    X = 1,
    q_eval = qy,
    H_function = inverse_logit
  )
}

estr <- MEstimator(stacked_equations = psi_robins, init = c(300, 0)) |>
  estimate()

c(mu = coef(estr)[[1]], confint(estr)[1, ])
```

This gives the same result as the by-hand version, as expected.

### Grid of alpha values

The prior results are only for a single value of $\alpha$. There is no strong
reason to believe that the true $\alpha$ is actually $0.01$. Instead, we should
look over a range of values for $\alpha$. This is also where M-estimation
shines: without it, we would need to bootstrap to appropriately estimate the
variance, which can be computationally intensive when exploring a large grid of
values.

The following code explores the sensitivity analysis over a grid of values for
$\alpha$ and then plots them. Note that we start at $\alpha := 0$ and build
outward, using the previous solution as the initial values for the next
iteration.

```{r}
#| label: sens-grid
alpha_l <- seq(0.0, -0.04, length.out = 100)
alpha_u <- seq(0.0, 0.04, length.out = 100)
point <- c()
lcl <- c()
ucl <- c()

for (alphas in list(alpha_l, alpha_u)) {
  init_vals <- c(400, 0)
  for (alpha in alphas) {
    # Updating q-function values
    qy <- q_function(y_vals = y, alpha = alpha)

    # Applying M-estimator
    estr <- MEstimator(stacked_equations = psi_robins, init = init_vals) |>
      estimate()
    ci <- confint(estr)

    # Extracting output
    point <- c(point, estr@theta[1])
    lcl <- c(lcl, ci[1, 1])
    ucl <- c(ucl, ci[1, 2])

    # Update init to speed up root-finding in next iteration
    init_vals <- estr@theta
  }
}

# Combine and sort
p <- data.frame(
  alpha = c(alpha_l, alpha_u),
  point = point,
  lcl = lcl,
  ucl = ucl
)
p <- p[order(p$alpha), ]
```

```{r}
#| label: sens-grid-plot
plot(p$alpha, p$point, type = "l",
     xlab = expression(alpha), ylab = expression(hat(mu)),
     ylim = c(300, 900))
polygon(c(p$alpha, rev(p$alpha)),
        c(p$lcl, rev(p$ucl)),
        col = rgb(0, 0, 0, 0.2), border = NA)
legend("topright", legend = "95% CI",
       fill = rgb(0, 0, 0, 0.2), border = NA)
```

Estimating $\mu$ across this grid of $\alpha$ takes a few seconds at most.
This is dramatically faster than bootstrapping. This again highlights the power
of M-estimation and the sandwich variance.

As we can see in the plot, the parameters asymptote at certain values. This is
indeed the case: this sensitivity analysis returns the nonparametric bounds in
the limit.

## Bounds

For the mean of a continuous variable, the bounds make the extreme assumption
that everyone has the highest (lowest) observed $Y$ value to compute the upper
(lower) bound on $\mu$. This is relatively easy to program by adapting the
complete-case analysis code.

```{r}
#| label: bounds
psi_bounds <- function(theta) {
  # Lower bound: replace missing with observed minimum
  ee_lower <- matrix(
    ifelse(is.na(y), min(y, na.rm = TRUE), y) - theta[1],
    nrow = 1
  )
  # Upper bound: replace missing with observed maximum
  ee_upper <- matrix(
    ifelse(is.na(y), max(y, na.rm = TRUE), y) - theta[2],
    nrow = 1
  )
  rbind(ee_lower, ee_upper)
}

estr <- MEstimator(stacked_equations = psi_bounds, init = c(400, 700)) |>
  estimate()
ci <- confint(estr)

# Bounds, with the interval running from the lower limit of the lower bound
# to the upper limit of the upper bound
c(
  lower_bound = coef(estr)[[1]],
  upper_bound = coef(estr)[[2]],
  lower = ci[1, 1],
  upper = ci[2, 2]
)
```

These are the nonparametric bounds for missingness of the mean. Note that these
bounds assume the maximum possible value of $Y$ for the missing observations is
the maximum in the observed data. We can avoid this assumption by replacing the
maximum (minimum) with larger (smaller) values if we are concerned.

This example shows how `deli` can be used for sensitivity
analyses. It also highlights the advantage of the sandwich variance for
computational purposes, allowing us to explore a range of sensitivity analyses
while remaining computationally quick.

## References

Cole, S. R., Zivich, P. N., Edwards, J. K., Shook-Sa, B. E., & Hudgens,
M. G. (2023). Sensitivity analyses for means or proportions with missing
outcome data. *Epidemiology*, 34(5), 645-651.

Robins, J. M., Rotnitzky, A., & Scharfstein, D. O. (2000). Sensitivity
analysis for selection bias and unmeasured confounding in missing data and
causal inference models. In *Statistical models in epidemiology, the
environment, and clinical trials* (pp. 1-94). New York, NY: Springer New York.
