---
title: "Cole et al. (2023): Fusion Designs"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[Cole et al. (2023): Fusion Designs example](https://deli.readthedocs.io/en/latest/Examples/Cole-AJE-2023.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

The following is a replication of the cases described in Cole et al. (2023).
Fusion designs combine information from multiple data sources to estimate
quantities that cannot be identified from any single source alone. These
designs are useful when the target population lacks the outcome of interest
but shares covariates with a study population where the outcome is measured.

The three cases below illustrate: (1) transporting a proportion to a target
population using inverse odds of sampling weights, (2) correcting for
outcome misclassification using the Rogan-Gladen estimator, and (3)
combining transport and misclassification correction in a single analysis.

## Case 1: Transporting the proportion

In this scenario, we have two populations: a study sample ($S=0$) where the
outcome $Y$ is observed, and a target population ($S=1$) where $Y$ is
missing. Both populations share a covariate $W$. The goal is to transport the
estimated proportion from the study to the target population using inverse
odds of sampling weights derived from a logistic model for $S \mid W$.

```{r}
#| label: case1-data
# Data from Cole et al. (2023), Table 1
d1 <- data.frame(
  Y = c(0, 1, 0, 1, 0, 0),
  W = c(0, 0, 1, 1, 0, 1),
  n = c(266, 67, 400, 267, 333, 167),
  S = c(0, 0, 0, 0, 1, 1)
)

# Expand rows by count
d1 <- d1[rep(seq_len(nrow(d1)), d1$n), ]
d1$n <- NULL
rownames(d1) <- NULL

# Replace Y with 0 where S=1 (target population has no outcome)
d1$Y_fill <- ifelse(d1$S == 1, 0, d1$Y)

# Design matrix for sampling model: intercept + W
d1$intercept <- 1
```

```{r}
#| label: case1
# Extract arrays
y <- d1$Y_fill
s <- d1$S
W <- as.matrix(d1[, c("intercept", "W")])

estr <- m_estimate(
  stacked_equations = function(theta) {
    mu_w <- theta[1]    # Weighted (transported) proportion
    mu_n <- theta[2]    # Naive (study-only) proportion
    beta <- theta[3:4]  # Sampling model coefficients

    # Sampling model: logistic regression of S on W
    ee_sm <- ee_regression(beta, X = W, y = s, model = "logistic")

    # Predicted probability of being in sample
    pi_s <- inverse_logit(as.numeric(W %*% beta))

    # Inverse odds of sampling weights
    iosw <- (1 - s) * pi_s / (1 - pi_s)

    # Weighted proportion (transported to target)
    ee_wprop <- matrix(iosw * (y - mu_w), nrow = 1)

    # Naive proportion (study sample only)
    ee_nprop <- matrix((1 - s) * (y - mu_n), nrow = 1)

    rbind(ee_wprop, ee_nprop, ee_sm)
  },
  init = c(0.5, 0.5, 0., 0.)
)

data.frame(
  Param = c("mu_weighted", "mu_naive", "beta_0", "beta_1"),
  Coef = round(estr@theta, 4),
  LCL = round(confint(estr)[, 1], 4),
  UCL = round(confint(estr)[, 2], 4)
)
```

The weighted (transported) proportion adjusts for the difference in the
distribution of $W$ between the study and target populations. The naive
proportion reflects the unadjusted estimate from the study sample alone.

## Case 2: Misclassified proportion (Rogan-Gladen)

When the outcome is subject to misclassification, the Rogan-Gladen estimator
uses external validation data to estimate sensitivity and specificity, then
corrects the observed proportion. Here, $R=1$ denotes the main study (where
only the mismeasured $Y^*$ is observed), and $R=0$ denotes the external
validation sample (where both $Y$ and $Y^*$ are observed).

```{r}
#| label: case2-data
# Misclassification data from Cole et al. (2023)
d2 <- data.frame(
  R     = c(1, 1, 0, 0, 0, 0),
  Y     = c(0, 0, 1, 1, 0, 0),
  Y_star = c(1, 0, 1, 0, 1, 0),
  n     = c(680, 270, 204, 38, 18, 71)
)

# Expand rows by count
d2 <- d2[rep(seq_len(nrow(d2)), d2$n), ]
d2$n <- NULL
rownames(d2) <- NULL
```

```{r}
#| label: case2
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_rogan_gladen(theta, y = d2$Y, y_star = d2$Y_star, r = d2$R)
  },
  init = c(0.5, 0.5, 0.75, 0.75)
)

data.frame(
  Param = c("Corrected", "Mismeasured", "Sensitivity", "Specificity"),
  Coef = round(estr@theta, 4),
  LCL = round(confint(estr)[, 1], 4),
  UCL = round(confint(estr)[, 2], 4)
)
```

The corrected proportion adjusts the mismeasured proportion using the
estimated sensitivity and specificity from the validation sample. This
demonstrates the Rogan-Gladen identity:
$\mu = (\mu^* + \text{specificity} - 1) / (\text{sensitivity} + \text{specificity} - 1)$.

## Case 3: Combined transport and misclassification

In this case, we combine both ideas: the outcome is misclassified in the
study population and must be transported to a target population. We have
three data sources:

- $S=1$: Main study with mismeasured outcome $Y^*$
- $S=2$: Target population with no outcome
- $S=3$: Validation sample with both $Y$ and $Y^*$

The stacked estimating equations include: (1) the Rogan-Gladen correction
applied with inverse odds of sampling weights to transport the corrected
proportion, and (2) a logistic sampling model to estimate the weights. The
main study ($S=1$) contributes the mismeasured outcome and the covariate
$W$, the target population ($S=2$) contributes only $W$, and the validation
sample ($S=3$) contributes both $Y$ and $Y^*$. Missing entries in each
source are filled with a placeholder value and excluded from the relevant
estimating functions by construction.

```{r}
#| label: case3-data
# Combined fusion data from Cole et al. (2023) rejoinder
d3 <- data.frame(
  Y_star = c(0, 1, 0, 1, NA, NA, 1, 0, 1, 0),
  Y      = c(NA, NA, NA, NA, NA, NA, 1, 1, 0, 0),
  W      = c(0, 0, 1, 1, 0, 1, NA, NA, NA, NA),
  n      = c(266, 67, 400, 267, 333, 167, 180, 20, 60, 240),
  S      = c(1, 1, 1, 1, 2, 2, 3, 3, 3, 3)
)

# Expand rows by count
d3 <- d3[rep(seq_len(nrow(d3)), d3$n), ]
d3$n <- NULL
rownames(d3) <- NULL
d3$intercept <- 1
```

```{r}
#| label: case3
# Replace missing values with a placeholder; these rows are zeroed out of the
# estimating functions where they do not apply
y_no_nan <- ifelse(is.na(d3$Y), -1, d3$Y)
ystar_no_nan <- ifelse(is.na(d3$Y_star), -1, d3$Y_star)
W_no_nan <- as.matrix(
  data.frame(intercept = d3$intercept, W = ifelse(is.na(d3$W), -1, d3$W))
)

# Population indicators
s1 <- as.numeric(d3$S == 1)   # Main study (mismeasured outcome)
s2 <- as.numeric(d3$S == 2)   # Target population
s3 <- as.numeric(d3$S == 3)   # Validation sample

estr <- m_estimate(
  stacked_equations = function(theta) {
    param <- theta[1:4]  # Measurement error parameters
    beta <- theta[5:6]   # Sampling model coefficients

    # Sampling model: logistic regression of the target indicator on W,
    # restricted to the main study and target populations (S=1 or S=2)
    ee_sm <- ee_regression(beta, X = W_no_nan, y = s2, model = "logistic")
    ee_sm <- ee_sm * matrix(1 - s3, nrow = nrow(ee_sm),
                            ncol = ncol(ee_sm), byrow = TRUE)

    # Inverse odds of sampling weights: odds weight for the main study,
    # one for the validation sample, zero for the target population
    pi_s <- inverse_logit(as.numeric(W_no_nan %*% beta))
    iosw <- s1 * pi_s / (1 - pi_s) + s3

    # Weighted Rogan-Gladen correction, restricted to S=1 or S=3
    ee_rg <- ee_rogan_gladen(param, y = y_no_nan, y_star = ystar_no_nan,
                             r = s1, weights = iosw)
    ee_rg <- ee_rg * matrix(1 - s2, nrow = nrow(ee_rg),
                            ncol = ncol(ee_rg), byrow = TRUE)

    rbind(ee_rg, ee_sm)
  },
  init = c(0.5, 0.5, 0.75, 0.75, 0., 0.),
  solver = "lm"
)

se <- sqrt(diag(vcov(estr)))
data.frame(
  Param = c("Corrected_transported", "Mismeasured_weighted",
            "Sensitivity", "Specificity", "beta_0", "beta_1"),
  Coef = round(estr@theta, 4),
  SE = round(se, 4),
  LCL = round(confint(estr)[, 1], 4),
  UCL = round(confint(estr)[, 2], 4)
)
```

This combined estimator simultaneously corrects for misclassification and
transports the result to the target population. The sandwich variance
estimator properly accounts for the uncertainty in all nuisance parameters
(sensitivity, specificity, and sampling weights), yielding valid confidence
intervals for the transported, corrected proportion.

## References

Cole SR, Edwards JK, Breskin A, Rosin S, Zivich PN, Shook-Sa BE, & Hudgens
MG. (2023). Illustration of two fusion designs and estimators. *American
Journal of Epidemiology*, 192(3), 467-474.
