---
title: "Cole et al. (2022): Addressing Missing Outcome Data"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[Cole et al. (2022): Addressing Missing Outcome Data example](https://deli.readthedocs.io/en/latest/Examples/Cole-AJE-2022.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

Cole et al. (2022) reviewed approaches to addressing missing outcome data
using direct imputation (g-computation). This was done in the context of a
randomized trial on 17P injection for the prevention of preterm birth. All
data was described in a table in the paper, so there is no need to track down
any external data set.

Here, we review selected examples and how g-computation for missing data can
be implemented using M-estimation.

## Data

The trial data is reconstructed from the table in the paper. Four treatment
by cervix-length strata define the observed outcomes. Several missing-data
scenarios (`preterm1` through `preterm4`) induce different missingness
patterns.

```{r}
#| label: data
d <- data.frame(
  short_cervix = c(rep(0, 215), rep(0, 222), rep(1, 186), rep(1, 177)),
  p17 = c(rep(0, 215), rep(1, 222), rep(0, 186), rep(1, 177))
)
d$placebo <- 1 - d$p17

d$preterm0 <- c(
  rep(1, 15), rep(0, 215 - 15),
  rep(1, 13), rep(0, 222 - 13),
  rep(1, 21), rep(0, 186 - 21),
  rep(1, 23), rep(0, 177 - 23)
)

d$preterm1 <- c(
  rep(1, 11), rep(0, 161 - 11), rep(NA, 54),
  rep(1, 10), rep(0, 167 - 10), rep(NA, 55),
  rep(1, 16), rep(0, 140 - 16), rep(NA, 46),
  rep(1, 17), rep(0, 133 - 17), rep(NA, 44)
)

d$preterm2 <- c(
  rep(1, 8), rep(0, 108 - 8), rep(NA, 107),
  rep(1, 13), rep(0, 222 - 13), rep(NA, 0),
  rep(1, 21), rep(0, 186 - 21), rep(NA, 0),
  rep(1, 12), rep(0, 89 - 12), rep(NA, 88)
)

d$preterm3 <- c(
  rep(1, 15), rep(0, 215 - 15), rep(NA, 0),
  rep(1, 13), rep(0, 222 - 13), rep(NA, 0),
  rep(1, 21), rep(0, 186 - 21), rep(NA, 0),
  rep(1, 0), rep(0, 0), rep(NA, 177)
)

d$preterm4 <- c(
  rep(1, 15), rep(0, 100 - 15), rep(NA, 115),
  rep(1, 7), rep(0, 216 - 7), rep(NA, 6),
  rep(1, 21), rep(0, 186 - 21), rep(NA, 0),
  rep(1, 12), rep(0, 81 - 12), rep(NA, 96)
)

d$C <- 1
n <- nrow(d)
```

## Full-data risk ratio

To start, we estimate the risk ratio using the complete outcome data
(`preterm0`). The estimating function computes each risk separately and then
takes the ratio (equivalently, the difference on the log-scale).

```{r}
#| label: full-data-rr
estr <- m_estimate(

  stacked_equations = function(theta) {
    a <- d$placebo
    y <- d$preterm0

    # Risk among placebo arm
    ee_r1 <- matrix(a * (y - theta[2]), nrow = 1)
    # Risk among 17P arm
    ee_r0 <- matrix((1 - a) * (y - theta[3]), nrow = 1)
    # Log risk ratio
    ee_rr <- matrix(rep(log(theta[2]) - log(theta[3]) - theta[1], n), nrow = 1)

    rbind(ee_rr, ee_r1, ee_r0)
  },
  init = c(0, 0.5, 0.5)
)

exp(c(rr = coef(estr)[[1]], confint(estr)[1, ]))
```

### Log-binomial model

In the paper, the risk ratio is estimated using a log-binomial model. We can
also do this via the `ee_glm` built-in estimating function. One caution is
that the log-binomial can be finicky with starting values. In particular, we
want to provide a good starting value for the intercept (using the mean of
$Y$).

```{r}
#| label: log-binomial
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_glm(
      theta = theta,
      X = as.matrix(d[, c("C", "placebo")]),
      y = d$preterm0,
      distribution = "binomial",
      link = "log"
    )
  },
  init = c(log(0.08), 0)
)

exp(c(rr = coef(estr)[[2]], confint(estr)[2, ]))
```

Both approaches give the same risk ratio to many decimal places.

## Scenario A

The first scenario induces non-informative missing outcome data (i.e.,
missing completely at random). Here, we expect a complete-case analysis and
g-computation to both be unbiased. The g-computation results should also be
more efficient (narrower confidence intervals).

### Complete-case analysis

```{r}
#| label: scenario-a-setup
a <- d$placebo
r <- ifelse(is.na(d$preterm1), 0, 1)
y_no_nan <- ifelse(is.na(d$preterm1), -999, d$preterm1)
```

```{r}
#| label: scenario-a-cc
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_glm(
      theta = theta,
      X = as.matrix(d[, c("C", "placebo")]),
      y = y_no_nan,
      distribution = "binomial",
      link = "log",
      weights = r
    )
  },
  init = c(log(0.09), 0)
)

exp(c(rr = coef(estr)[[2]], confint(estr)[2, ]))
```

As expected in this scenario, the complete-case results are similar to the
full-data results. The confidence intervals are wider due to smaller
effective sample size.

### G-computation

```{r}
#| label: scenario-a-gcomp
estr <- m_estimate(
  stacked_equations = function(theta) {
    beta1 <- theta[4:5]
    beta0 <- theta[6:7]
    X <- as.matrix(d[, c("C", "short_cervix")])

    # Outcome model for 17P=1 (placebo)
    ee_reg1 <- ee_regression(beta1, X = X, y = y_no_nan, model = "logistic")
    ee_reg1 <- ee_reg1 * matrix(r * a, nrow = nrow(ee_reg1), ncol = n, byrow = TRUE)

    # Outcome model for 17P=0
    ee_reg0 <- ee_regression(beta0, X = X, y = y_no_nan, model = "logistic")
    ee_reg0 <- ee_reg0 * matrix(r * (1 - a), nrow = nrow(ee_reg0), ncol = n, byrow = TRUE)

    # Predicted risks under each treatment
    y1hat <- inverse_logit(as.numeric(X %*% beta1))
    y0hat <- inverse_logit(as.numeric(X %*% beta0))

    # Risk in each arm and log risk ratio
    ee_r1 <- matrix(y1hat - theta[2], nrow = 1)
    ee_r0 <- matrix(y0hat - theta[3], nrow = 1)
    ee_rr <- matrix(rep(log(theta[2]) - log(theta[3]) - theta[1], n), nrow = 1)

    rbind(ee_rr, ee_r1, ee_r0, ee_reg1, ee_reg0)
  },
  init = c(0, 0.5, 0.5, 0, 0, 0, 0)
)

exp(c(rr = coef(estr)[[1]], confint(estr)[1, ]))
```

The g-computation results are similar to the complete-case results. Both are
unbiased under MCAR.

## Scenario B

In scenario B, we consider informative missing data. Specifically, cervix
length is related to missingness. Here, a complete-case analysis will be
biased, but g-computation should correct for this bias.

### Complete-case analysis

```{r}
#| label: scenario-b-setup
a <- d$placebo
r <- ifelse(is.na(d$preterm2), 0, 1)
y_no_nan <- ifelse(is.na(d$preterm2), -999, d$preterm2)
```

```{r}
#| label: scenario-b-cc
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_glm(
      theta = theta,
      X = as.matrix(d[, c("C", "placebo")]),
      y = y_no_nan,
      distribution = "binomial",
      link = "log",
      weights = r
    )
  },
  init = c(log(0.09), 0)
)

exp(c(rr = coef(estr)[[2]], confint(estr)[2, ]))
```

### G-computation

```{r}
#| label: scenario-b-gcomp
estr <- m_estimate(
  stacked_equations = function(theta) {
    beta1 <- theta[4:5]
    beta0 <- theta[6:7]
    X <- as.matrix(d[, c("C", "short_cervix")])

    # Outcome model for 17P=1 (placebo)
    ee_reg1 <- ee_regression(beta1, X = X, y = y_no_nan, model = "logistic")
    ee_reg1 <- ee_reg1 * matrix(r * a, nrow = nrow(ee_reg1), ncol = n, byrow = TRUE)

    # Outcome model for 17P=0
    ee_reg0 <- ee_regression(beta0, X = X, y = y_no_nan, model = "logistic")
    ee_reg0 <- ee_reg0 * matrix(r * (1 - a), nrow = nrow(ee_reg0), ncol = n, byrow = TRUE)

    # Predicted risks under each treatment
    y1hat <- inverse_logit(as.numeric(X %*% beta1))
    y0hat <- inverse_logit(as.numeric(X %*% beta0))

    # Risk in each arm and log risk ratio
    ee_r1 <- matrix(y1hat - theta[2], nrow = 1)
    ee_r0 <- matrix(y0hat - theta[3], nrow = 1)
    ee_rr <- matrix(rep(log(theta[2]) - log(theta[3]) - theta[1], n), nrow = 1)

    rbind(ee_rr, ee_r1, ee_r0, ee_reg1, ee_reg0)
  },
  init = c(0, 0.5, 0.5, 0, 0, 0, 0)
)

exp(c(rr = coef(estr)[[1]], confint(estr)[1, ]))
```

Here, the complete-case analysis indicates a spurious relationship between
17P and preterm birth. However, g-computation correctly removes this bias by
adjusting for cervix length as the driver of missingness.

Unlike the original paper, we avoid bootstrapping for variance estimation by
using the sandwich variance estimator instead, which is considerably more
efficient computationally. This advantage of M-estimation was noted in the
Discussion of Cole et al. (2022) but had not been demonstrated elsewhere.

Additional scenarios are available in the data as `preterm3` and `preterm4`.

## References

Cole, S. R., Zivich, P. N., Edwards, J. K., Ross, R. K., Shook-Sa, B. E.,
Price, J. T., & Stringer, J. S. (2023). Missing outcome data in
epidemiologic studies. *American Journal of Epidemiology*, 192(1), 6-10.
