---
title: "Silverman (1985): Generalized Additive Model"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(deli)
```

The following is a brief illustrative example of how generalized additive
models (GAMs) can be implemented using M-estimation. For demonstration, we use
data on motorcycle impacts from Silverman (1985).

## Loading Data

The motorcycle data contains measurements of head acceleration (in g) over
time (in ms) during simulated motorcycle impacts. The relationship between
time and acceleration is clearly non-linear.

```{r}
#| label: data
# Load the motorcycle data from the MASS package
d <- MASS::mcycle

# Format data for deli
X <- cbind(1, d$times)
y <- d$accel

# Create prediction grid
p_times <- seq(min(d$times), max(d$times), length.out = 200)
Xp <- cbind(1, p_times)
```

## Linear Regression

As easily seen in the scatterplot, the relationship between time and
acceleration is non-linear. We can show this further by fitting a regression
model with a linear relationship between the variables.

```{r}
#| label: linear
# Define estimating equation for linear regression
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_regression(theta, X = X, y = y, model = "linear")
  },
  init = c(0, 0)
)
estr <- estimate(estr)

# Predicted values and confidence intervals
pred_y <- regression_predictions(Xp, estr@theta, estr@variance)

# Plot
plot(d$times, d$accel, pch = 16, cex = 0.5,
     xlab = "Time (ms)", ylab = "Acceleration (g)")
lines(p_times, pred_y$predicted, col = "orange", lwd = 2)
polygon(
  c(p_times, rev(p_times)),
  c(pred_y$lower, rev(pred_y$upper)),
  col = adjustcolor("orange", alpha.f = 0.3), border = NA
)
```

As we can clearly see, a linear relationship is very different from the
observed data.

## Implementing a GAM

GAMs are implemented through the use of $L_2$-penalized splines. The user
needs to provide the knot locations (or number of knots). Other optional
arguments include the power of the splines, whether to restrict, and the
strength of the penalty. Stronger penalties move the model closer to the fit
of a linear model.

For the intercept term, we do not want any splines. For times, we specify
12 knots and a penalty of 2000. This penalty prevents the GAM from being too
"wiggly".

```{r}
#| label: gam
# Define knot locations at percentiles of times
knot_locs <- as.numeric(quantile(d$times, probs = seq(0.025, 0.975, length.out = 11)))

# Specifications for the additive design matrix
specs <- list(
  NULL,                                                    # No spline for intercept
  list(knots = knot_locs, penalty = 2000, power = 3)       # Spline for times
)

# Define estimating equation for additive regression
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_additive_regression(
      theta, X = X, y = y,
      model = "linear",
      specifications = specs
    )
  },
  init = rep(0, 12)
)
estr <- estimate(estr)

# Build the additive design matrix for predictions
Xpa <- additive_design_matrix(Xp, specifications = specs)

# Generate predicted values
pred_y <- regression_predictions(Xpa, estr@theta, estr@variance)

# Plot
plot(d$times, d$accel, pch = 16, cex = 0.5,
     xlab = "Time (ms)", ylab = "Acceleration (g)")
lines(p_times, pred_y$predicted, col = "orange", lwd = 2)
polygon(
  c(p_times, rev(p_times)),
  c(pred_y$lower, rev(pred_y$upper)),
  col = adjustcolor("orange", alpha.f = 0.3), border = NA
)
```

The GAM does a much better job of following the shape of the scatterplot.

### Impact of the Penalty

To show how the penalty impacts the wiggliness of the GAM, let's increase the
penalty by a large amount. As this penalty is only applied to the spline terms,
we would expect to see the regression line become smoother.

#### Strong Penalty (1,000,000)

```{r}
#| label: gam-strong-penalty
# Increase the penalty
specs_strong <- list(
  NULL,
  list(knots = knot_locs, penalty = 1000000, power = 3)
)

estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_additive_regression(
      theta, X = X, y = y,
      model = "linear",
      specifications = specs_strong
    )
  },
  init = rep(0, 12)
)
estr <- estimate(estr)

# Build additive design matrix and predict
Xpa <- additive_design_matrix(Xp, specifications = specs_strong)
pred_y <- regression_predictions(Xpa, estr@theta, estr@variance)

# Plot
plot(d$times, d$accel, pch = 16, cex = 0.5,
     xlab = "Time (ms)", ylab = "Acceleration (g)")
lines(p_times, pred_y$predicted, col = "orange", lwd = 2)
polygon(
  c(p_times, rev(p_times)),
  c(pred_y$lower, rev(pred_y$upper)),
  col = adjustcolor("orange", alpha.f = 0.3), border = NA
)
```

The estimated function from the GAM is smoother than the previous GAM. Let's
crank up the penalty even further.

#### Very Strong Penalty (100,000,000,000)

```{r}
#| label: gam-very-strong-penalty
# Very large penalty
specs_vstrong <- list(
  NULL,
  list(knots = knot_locs, penalty = 1e11, power = 3)
)

estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_additive_regression(
      theta, X = X, y = y,
      model = "linear",
      specifications = specs_vstrong
    )
  },
  init = rep(0, 12)
)
estr <- estimate(estr)

# Build additive design matrix and predict
Xpa <- additive_design_matrix(Xp, specifications = specs_vstrong)
pred_y <- regression_predictions(Xpa, estr@theta, estr@variance)

# Plot
plot(d$times, d$accel, pch = 16, cex = 0.5,
     xlab = "Time (ms)", ylab = "Acceleration (g)")
lines(p_times, pred_y$predicted, col = "orange", lwd = 2)
polygon(
  c(p_times, rev(p_times)),
  c(pred_y$lower, rev(pred_y$upper)),
  col = adjustcolor("orange", alpha.f = 0.3), border = NA
)
```

Here, the GAM is very similar to the linear regression we started with. This
demonstrates the general relationship: as the penalty goes to infinity, the GAM
will coincide with the linear model.

To summarize, we want a penalty that prevents overfitting to the data but does
not prevent the GAM from approximating the data. This selection of penalization
is commonly done via cross-validation. However, use of cross-validation
invalidates standard variance estimation approaches (including with the sandwich
variance estimator). Therefore, we recommend sacrificing the optimal penalty
parameter (as determined by cross-validation) in favor of simpler variance
estimation.

GAMs are a flexible method that place fewer constraints on the functional forms
of relationships via splines. These splines do not require us to know the true
underlying functional form. Instead, the higher-order splines (with well-chosen
knots) can serve as close approximations.

## References

Silverman BW. (1985). Some aspects of the spline smoothing approach to
non-parametric regression curve fitting. *Journal of the Royal Statistical
Society: Series B (Methodological)*, 47(1), 1-21.
