---
title: "GMM for Over-Identified Parameters"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[GMM for Over-Identified Parameters example](https://deli.readthedocs.io/en/latest/Examples/GMM-OverID.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

When there are more estimating equations than parameters, the system is
*over-identified*. The standard `MEstimator` requires a square system (number of
equations equals number of parameters). `GMMEstimator` handles over-identified
systems by minimizing a quadratic form of the estimating equations, producing
valid point estimates and sandwich variance estimates.

This document walks through instrumental variable (IV) examples where combining
multiple instruments creates over-identification.

## IV Example 1: Separate instruments (just-identified)

Consider estimating the causal effect of treatment $A$ on outcome $Y$ using two
instrumental variables $Z_1$ and $Z_2$. When each instrument gets its own
parameter, the system is just-identified: 2 equations, 2 parameters.

### Data

```{r}
#| label: data1
# Generate data
set.seed(777)
n <- 500

# Covariate
W <- rbinom(n, size = 1, prob = 0.25)

# Two instruments
Z1 <- rnorm(n, mean = 0, sd = 0.5)
Z2 <- rnorm(n, mean = 0, sd = 0.5)

# Treatment depends on both instruments
A <- Z1 + Z2 + rnorm(n)

# Outcome with effect modification by W
Y <- 2 * A - W * A + rnorm(n)
```

### Estimation with separate parameters

Each instrument defines its own estimating equation with its own parameter. This
is a square system (2 equations, 2 parameters), so `MEstimator` works.

```{r}
#| label: iv-separate
# Two instruments, two parameters (just-identified)
estr <- MEstimator(
  stacked_equations = function(theta) {
    # Z1 moment condition with beta_1
    ee1 <- matrix(Z1 * (Y - theta[1] * A), nrow = 1)
    # Z2 moment condition with beta_2
    ee2 <- matrix(Z2 * (Y - theta[2] * A), nrow = 1)
    rbind(ee1, ee2)
  },
  init = c(0, 0)
) |>
  estimate()

data.frame(
  Param = c("beta_1 (Z1)", "beta_2 (Z2)"),
  Coef = round(estr@theta, 4),
  LCL = round(confint(estr)[, 1], 4),
  UCL = round(confint(estr)[, 2], 4)
)
```

Each instrument gives a separate estimate of the effect. Both are valid, but
they use the data inefficiently since each equation only leverages one
instrument.

## IV Example 1: Combined instruments (over-identified)

Now we combine both instruments to estimate a single parameter $\beta$. This
creates an over-identified system: 2 equations, 1 parameter. Both moment
conditions $E[Z_1 (Y - \beta A)] = 0$ and $E[Z_2 (Y - \beta A)] = 0$ must hold
simultaneously for the same $\beta$.

### MEstimator fails

`MEstimator` requires a square system (number of equations = number of
parameters). With 2 equations and 1 parameter, it cannot solve the system.

```{r}
#| label: mestimator-fail
#| error: true
# Over-identified: 2 equations, 1 parameter
estr_fail <- MEstimator(
  stacked_equations = function(theta) {
    # Both instruments, single beta
    ee1 <- matrix(Z1 * (Y - theta[1] * A), nrow = 1)
    ee2 <- matrix(Z2 * (Y - theta[1] * A), nrow = 1)
    rbind(ee1, ee2)
  },
  init = c(0)
) |>
  estimate()
```

### GMMEstimator works

`GMMEstimator` handles the over-identified case by minimizing a quadratic form
$Q(\theta) = \bar{g}(\theta)^T W \bar{g}(\theta)$, where $\bar{g}(\theta)$ is
the vector of average estimating equations and $W$ is a weight matrix. It
iterates to find the optimal weight matrix.

```{r}
#| label: gmm-combined
# Over-identified: 2 equations, 1 parameter
estr_gmm <- GMMEstimator(
  stacked_equations = function(theta) {
    # Both instruments, single beta
    ee1 <- matrix(Z1 * (Y - theta[1] * A), nrow = 1)
    ee2 <- matrix(Z2 * (Y - theta[1] * A), nrow = 1)
    rbind(ee1, ee2)
  },
  init = c(0)
) |>
  estimate()

data.frame(
  Param = "beta (combined)",
  Coef = round(estr_gmm@theta, 4),
  LCL = round(confint(estr_gmm)[, 1], 4),
  UCL = round(confint(estr_gmm)[, 2], 4)
)
```

By combining both instruments into a single estimating equation system,
`GMMEstimator` efficiently pools information from both $Z_1$ and $Z_2$. This
typically produces narrower confidence intervals compared to using either
instrument alone.

## IV Example 2: With transportability

Now consider a setting with two populations: an external study population
($S = 0$) where instrumental variable analyses can be conducted, and a target
population ($S = 1$) where we want to transport the results. We use inverse odds
of sampling weights to transport the IV estimates, combined with over-identified
moment conditions.

### Data

```{r}
#| label: data2
set.seed(777)

# External population (S=0)
n0 <- 400
W0 <- rbinom(n0, size = 1, prob = 0.25)
Z1_0 <- rnorm(n0, mean = 0, sd = 0.5)
Z2_0 <- rnorm(n0, mean = 0, sd = 0.5)
A0 <- Z1_0 + Z2_0 + rnorm(n0)
Y0 <- 2 * A0 - W0 * A0 + rnorm(n0)
S0 <- rep(0, n0)

# Target population (S=1)
n1 <- 100
W1 <- rbinom(n1, size = 1, prob = 0.5)
Z1_1 <- rep(0, n1)
Z2_1 <- rep(0, n1)
A1 <- rnorm(n1)
Y1 <- rep(0, n1)
S1 <- rep(1, n1)

# Combine into single data frame
d <- data.frame(
  W = c(W0, W1),
  Z1 = c(Z1_0, Z1_1),
  Z2 = c(Z2_0, Z2_1),
  A = c(A0, A1),
  Y = c(Y0, Y1),
  S = c(S0, S1)
)
```

### Estimation with transportability

The estimating equations have three parameters:

- $\beta$: the IV effect estimate, transported to the target population
- $\alpha_0, \alpha_1$: logistic regression coefficients for the sampling model $P(S = 1 \mid W)$

The IV moment conditions are weighted by inverse odds of sampling weights
$\frac{P(S=1 \mid W)}{P(S=0 \mid W)}$ to transport the estimates from the
external study to the target population.

```{r}
#| label: iv-transport
# Extract vectors for use in estimating equations
S <- d$S
W_vec <- d$W
Z1_all <- d$Z1
Z2_all <- d$Z2
A_all <- d$A
Y_all <- d$Y
X_s <- cbind(1, W_vec)  # Design matrix for sampling model

estr_transport <- GMMEstimator(
  stacked_equations = function(theta) {
    beta <- theta[1]
    alpha <- theta[2:3]
    n <- length(Y_all)

    # Sampling model: P(S=1 | W)
    ee_s <- ee_regression(alpha, X = X_s, y = S, model = "logistic")

    # Inverse odds of sampling weights
    pr_s <- inverse_logit(as.numeric(X_s %*% alpha))
    iosw <- pr_s / (1 - pr_s)

    # Weighted IV moment conditions (only external population contributes)
    ee1 <- matrix((1 - S) * iosw * Z1_all * (Y_all - beta * A_all), nrow = 1)
    ee2 <- matrix((1 - S) * iosw * Z2_all * (Y_all - beta * A_all), nrow = 1)

    rbind(ee1, ee2, ee_s)
  },
  init = c(0, 0, 0)
) |>
  estimate()

data.frame(
  Param = c("beta (transported)", "alpha_0", "alpha_1"),
  Coef = round(estr_transport@theta, 4),
  LCL = round(confint(estr_transport)[, 1], 4),
  UCL = round(confint(estr_transport)[, 2], 4)
)
```

This example has 4 estimating equations (2 IV moment conditions + 2 sampling
model equations) but only 3 parameters ($\beta$, $\alpha_0$, $\alpha_1$),
making it over-identified. The `GMMEstimator` efficiently combines both
instruments while correctly propagating uncertainty from the sampling model
through the sandwich variance estimator.

## References

Hansen LP. (1982). Large sample properties of generalized method of moments
estimators. *Econometrica*, 50(4), 1029-1054.
