---
title: "Hernan & Robins (2023): Causal Inference Examples"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[Hernan & Robins (2023): Causal Inference with Models example](https://deli.readthedocs.io/en/latest/Examples/Hernan-Robins-2023.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

The following section replicates selected examples from the textbook
["Causal Inference: What If"](https://www.routledge.com/Causal-Inference-What-If/Hernan-Robins/p/book/9781420076165)
by Hernan and Robins. These replications focus on Part II of the textbook.
I recommend reading Part I prior to looking through the following code. It is
a great, approachable, and freely available resource on causal inference.

Here, we demonstrate application of `deli` (R)
for causal inference. Throughout, we use the empirical sandwich variance
estimator. This is not described in the textbook, but is an alternative to
the bootstrap. Importantly, it is a consistent estimator of the variance
that is computationally simpler (in terms of the computer's computational
effort). The package automates the whole procedure.

Broadly, interest will be in estimating the average causal effect of stopping
smoking (variable name: `qsmk`) on 10-year weight change (variable name:
`wt82_71`). If we let $Y^a$ indicate the potential weight change under
smoking status $a$, then the average causal effect can be written as
$$E[Y^1] - E[Y^0]$$
Hereafter, we assume that the interest parameter is identified (see the book
for details on what this means). Our focus will be on estimators described in
the book for this quantity (or related ones).

## Loading Data

The data set used in the book and this tutorial is a subset of the full
NHEFS. First, we will load the data and run some basic variable
manipulations.

```{r}
#| label: data
# Load the NHEFS data from causaldata package
df <- as.data.frame(causaldata::nhefs)

# Convert factors to numeric (matching original CSV format)
for (col in c("sex", "race", "education", "active", "exercise")) {
  df[[col]] <- as.numeric(as.character(df[[col]]))
}

df <- df[complete.cases(df[, c("sex", "age", "race", "ht",
                                "school", "alcoholpy", "smokeintensity")]), ]

# Recoding some variables
df$inactive <- ifelse(df$active == 2, 1, 0)
df$no_exercise <- ifelse(df$exercise == 2, 1, 0)
df$university <- ifelse(df$education == 5, 1, 0)

# Subsetting only variables of interest
df <- df[, c("wt82_71", "qsmk", "sex", "age", "race", "wt71", "wt82", "ht",
             "school", "alcoholpy", "smokeintensity", "smokeyrs",
             "smkintensity82_71", "education", "exercise", "active", "death")]

# Creating quadratic terms
for (col in c("age", "wt71", "smokeintensity", "smokeyrs")) {
  df[[paste0(col, "_sq")]] <- df[[col]] * df[[col]]
}

df$I <- 1

# Indicator terms
df$educ_2 <- ifelse(df$education == 2, 1, 0)
df$educ_3 <- ifelse(df$education == 3, 1, 0)
df$educ_4 <- ifelse(df$education == 4, 1, 0)
df$educ_5 <- ifelse(df$education == 5, 1, 0)
df$exer_1 <- ifelse(df$exercise == 1, 1, 0)
df$exer_2 <- ifelse(df$exercise == 2, 1, 0)
df$active_1 <- ifelse(df$active == 1, 1, 0)
df$active_2 <- ifelse(df$active == 2, 1, 0)

# Interaction terms
df$qsmk_smkint <- df$qsmk * df$smokeintensity

# Complete-case data
dc <- df[complete.cases(df[, "wt82_71"]), ]
```

## Chapter 12: IP weighting and marginal structural models

The first estimation approach is inverse probability weighting. Inverse
probability weights are defined as
$$ \frac{1}{\Pr(A=a | W)} $$
where $W$ is the set of confounders. We will estimate these weights and
then use them to estimate the parameters of a marginal structural model.
An example of a marginal structural model is
$$ E[Y^a] = \alpha_0 + \alpha_1 a $$
where $\alpha$ are the parameters to estimate. Here, $\alpha_1$ represents
the average causal effect. We will estimate the marginal structural model
using the observed data and a regression model weighted by the inverse
probability weights (see the book for details).

### 12.1: The causal question

Chapter 12 starts out with estimating the crude association between `qsmk`
and `wt82_71` (12.1). We will do this by fitting a linear regression model.

```{r}
#| label: ch12-1
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_regression(theta = theta,
                  X = as.matrix(dc[, c("I", "qsmk")]),
                  y = dc$wt82_71,
                  model = "linear")
  },
  init = c(0, 0)
)

# The crude estimate of qsmk
crude <- cbind(estimate = coef(estr), confint(estr))
rownames(crude) <- c("intercept", "qsmk")
round(crude, 2)
```

The book reports an unadjusted estimate of 2.5 (95% CI: 1.7, 3.4).
The confidence interval may differ slightly because we are using the
sandwich variance (the book uses a different approach). While the variance
estimators used here and in the book are expected to be asymptotically
equal (i.e., equal as $n$ goes to $\infty$), they can produce different
results in finite samples.

### 12.2: Estimating inverse probability weights via modeling

Now we will estimate the parameters of the marginal structural model using
unstabilized inverse probability weights.

To do this, we will define the corresponding design matrices. Then we will
define the stacked estimating equations. Then we will estimate the
parameters and covariance and present the output.

```{r}
#| label: ch12-2-setup
# Design matrix for the propensity score model
W_cols <- c("I", "sex", "race", "age", "age_sq",
            "educ_2", "educ_3", "educ_4", "educ_5",
            "smokeintensity", "smokeintensity_sq",
            "smokeyrs", "smokeyrs_sq",
            "exer_1", "exer_2", "active_1", "active_2",
            "wt71", "wt71_sq")
W <- as.matrix(dc[, W_cols])

# Design matrix for the marginal structural model
msm <- as.matrix(dc[, c("I", "qsmk")])

# Treatment variable
a <- dc$qsmk

# Outcome variable
y <- dc$wt82_71

# Good practice: start the solver from the standalone propensity score fit.
# Keeping the coefficient names labels those parameters in later fits.
init_ps <- coef(glm(a ~ W - 1, family = binomial()))
```

```{r}
#| label: ch12-2-byhand
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters into corresponding estimation equations
    alpha <- theta[1:2]
    beta <- theta[3:length(theta)]

    # Estimating the propensity scores
    ee_ps <- ee_regression(theta = beta,        # Estimate propensity scores
                           X = W, y = a,        # ... given observed A,W
                           model = "logistic")  # ... with logit model
    pi_val <- inverse_logit(as.numeric(W %*% beta))  # Get Pr(A = 1 | W)
    ipw <- 1 / ifelse(a == 1, pi_val, 1 - pi_val)   # Convert to IPW

    # Estimating the MSM using a weighted linear model
    ee_msm <- ee_regression(theta = alpha,      # MSM parameters
                            X = msm, y = y,     # ... observed data
                            model = "linear",   # ... with linear model
                            weights = ipw)      # ... but weighted by IPW

    # Stacking the estimating equations and returning
    rbind(ee_msm, ee_ps)
  },
  init = c(0, 0, init_ps),
  solver = "nleqslv",
  maxiter = 5000
)

msm_est <- cbind(estimate = coef(estr), confint(estr))[1:2, ]
rownames(msm_est) <- c("alpha_0", "alpha_1 (ACE)")
round(msm_est, 2)
```

The book provides the point estimate for `qsmk` ($\hat{\alpha}_1$) as 3.4
(95% CI: 2.4, 4.5).

Instead of coding this by-hand, we can also use the built-in `ee_ipw_msm`
function. This function estimates a marginal structural model using inverse
probability weights, as done above.

```{r}
#| label: ch12-2-builtin
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Built-in estimating equation
    ee_ipw_msm(theta, y = y, A = a, W = W, V = msm,
               distribution = "normal",
               link = "identity")
  },
  init = c(0, 0, init_ps),
  solver = "nleqslv",
  maxiter = 5000
)

msm_est <- cbind(estimate = coef(estr), confint(estr))[1:2, ]
rownames(msm_est) <- c("alpha_0", "alpha_1 (ACE)")
round(msm_est, 2)
```

As expected, this built-in functionality produces the same results as the
by-hand version.

### 12.3: Stabilized inverse probability weights

Next, we are going to use stabilized weights. The stabilized weights will
require us to estimate an additional parameter. We will accomplish this by
stacking an estimating equation for that parameter. This extra estimating
equation is for an intercept-only model for the probability of `qsmk`.

```{r}
#| label: ch12-3
# Likewise, start the numerator model from its own intercept-only fit
init_num <- unname(coef(glm(a ~ 1, family = binomial())))

estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters into corresponding estimation equations
    alpha <- theta[1:2]                    # MSM parameters
    gamma <- theta[3]                      # Numerator parameter
    beta <- theta[4:length(theta)]         # Propensity score parameters

    # Estimating the propensity scores using a logit model
    ee_ps <- ee_regression(theta = beta,         # Propensity score model
                           X = W, y = a,         # ... with observed data
                           model = "logistic")   # ... and logit model
    pi_val <- inverse_logit(as.numeric(W %*% beta))  # Predicted prob of A=1

    # Estimating intercept-only for numerator
    ee_num <- ee_regression(theta = gamma,                       # Numerator model
                            X = as.matrix(dc[, "I", drop = FALSE]), y = a,  # ... intercept-only
                            model = "logistic")                  # ... logit model
    num_val <- inverse_logit(as.numeric(as.matrix(dc[, "I", drop = FALSE]) %*% gamma))

    # Construct stabilized weights
    ipw <- ifelse(a == 1, num_val / pi_val, (1 - num_val) / (1 - pi_val))

    # Estimating the MSM using a weighted linear model
    ee_msm <- ee_regression(theta = alpha,      # MSM
                            X = msm, y = y,     # ... with observed data
                            model = "linear",   # ... linear
                            weights = ipw)      # ... weighted by stabilized

    # Stacking the estimating equations and returning
    rbind(ee_msm, ee_num, ee_ps)
  },
  init = c(0, 0, init_num, init_ps),
  solver = "nleqslv",
  maxiter = 5000
)

msm_est <- cbind(estimate = coef(estr), confint(estr))[1:2, ]
rownames(msm_est) <- c("alpha_0", "alpha_1 (ACE)")
round(msm_est, 2)
```

The estimate is the same as in the previous section. This is expected
because as long as the marginal structural model is saturated, the
unstabilized and stabilized IPTW should produce the same answer.

### 12.4: Marginal structural models

Now we will consider the IPW estimator for a continuous action. We will
look at `smokeintensity` on `wt82_71`.

```{r}
#| label: ch12-4-setup
# Restricting data by smoking intensity
ds <- dc[dc$smokeintensity <= 25, ]

# Design matrix for the propensity score model
W_ds <- as.matrix(ds[, W_cols])

# Design matrix for the marginal structural model
ds$smkint_sq <- ds$smkintensity82_71^2
msm_ds <- as.matrix(ds[, c("I", "smkintensity82_71", "smkint_sq")])

# Treatment array
a_ds <- ds$smkintensity82_71

# Outcome array
y_ds <- ds$wt82_71
```

```{r}
#| label: ch12-4
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters into corresponding estimation equations
    alpha <- theta[1:3]                  # Marginal structural model
    gamma <- theta[4]                    # Numerator
    beta <- theta[5:length(theta)]       # Propensity score model
    n_obs <- nrow(W_ds)
    div_ps <- n_obs - length(beta)       # Divisor for PS SD
    div_nm <- n_obs - length(gamma)      # Divisor for Num SD

    # Estimating the propensity scores using a linear model
    ee_ps <- ee_regression(theta = beta,         # Generalized PS model
                           X = W_ds, y = a_ds,   # ... for observed data
                           model = "linear")     # ... linear regression
    mu <- as.numeric(W_ds %*% beta)              # Predicted values
    mu_resid <- sum((a_ds - mu)^2) / div_ps      # Standard deviation
    fAL <- dnorm(a_ds, mu, sqrt(mu_resid))       # PDF from normal

    # Estimating intercept-only for numerator
    I_ds <- as.matrix(ds[, "I", drop = FALSE])
    ee_num <- ee_regression(theta = gamma,           # Numerator for stabilized
                            X = I_ds, y = a_ds,      # ... for observed data
                            model = "linear")        # ... linear regression
    num_val <- as.numeric(I_ds %*% gamma)            # Predicted values
    num_resid <- sum((a_ds - num_val)^2) / div_nm    # Standard deviation
    fA <- dnorm(a_ds, num_val, sqrt(num_resid))      # PDF from normal

    # Stabilized weights
    ipw <- fA / fAL

    # Estimating the MSM using a weighted linear model
    ee_msm <- ee_regression(theta = alpha,       # Marginal structural model
                            X = msm_ds, y = y_ds, # ... observed data
                            model = "linear",    # ... linear model
                            weights = ipw)       # ... weighted by IPW

    # Stacking the estimating equations and returning
    rbind(ee_msm, ee_num, ee_ps)
  },
  init = c(0, 0, 0, 0, rep(0, ncol(W_ds))),
  solver = "nleqslv",
  tolerance = 1e-12,
  maxiter = 5000
)

# MSM coefficients
msm_est <- cbind(estimate = coef(estr), confint(estr))[1:3, ]
rownames(msm_est) <- c("alpha_0", "alpha_1", "alpha_2")
round(msm_est, 3)
```

The book reports coefficients of: 2.005, -0.109, 0.003. These match the
output shown above.

As done in the book, we want to know the weight change for no change in
smoking intensity and a +20 in smoking intensity.

```{r}
#| label: ch12-4-predictions
# Creating design matrix for combinations to predict
vals <- matrix(c(1, 0, 0,
                 1, 20, 400), nrow = 2, byrow = TRUE)

# Getting predicted values and variance for combinations
pred_y <- regression_predictions(vals,
                                 estr@theta[1:3],
                                 estr@variance[1:3, 1:3])
rownames(pred_y) <- c("no change in smoking", "+20 smoking intensity")

pred_y[, c("predicted", "lower", "upper")]
```

Again, we get similar results to those reported in the book: 2.0
(95% CI: 1.4, 2.6) and 0.9 (95% CI: -1.7, 3.5). However, our confidence
intervals are slightly more narrow since we are using a variance estimator
that is not overly conservative.

#### Binary outcome

We now repeat the process, but for a binary outcome and treatment. We will
use a GLM with the binomial distribution and logistic link (this will
estimate the causal odds ratio). We will also be using `qsmk` again.

```{r}
#| label: ch12-4-binary-setup
# Reset design matrices for qsmk analysis
W <- as.matrix(dc[, W_cols])
msm <- as.matrix(dc[, c("I", "qsmk")])
a <- dc$qsmk
y_death <- dc$death
```

```{r}
#| label: ch12-4-binary
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters into corresponding estimation equations
    alpha <- theta[1:2]              # Marginal structural model
    gamma <- theta[3]                # Numerator model
    beta <- theta[4:length(theta)]   # Propensity score model

    # Estimating the propensity scores using a logit model
    ee_ps <- ee_regression(theta = beta,         # Propensity score
                           X = W, y = a,         # ... observed data
                           model = "logistic")   # ... logistic model
    pi_val <- inverse_logit(as.numeric(W %*% beta))

    # Estimating intercept-only for numerator
    I_dc <- as.matrix(dc[, "I", drop = FALSE])
    ee_num <- ee_regression(theta = gamma,           # Numerator model
                            X = I_dc, y = a,         # ... observed data
                            model = "logistic")      # ... logit model
    num_val <- inverse_logit(as.numeric(I_dc %*% gamma))

    # Stabilized inverse probability weights
    ipw <- ifelse(a == 1, num_val / pi_val, (1 - num_val) / (1 - pi_val))

    # Estimating the MSM using a weighted GLM
    ee_msm <- ee_glm(theta = alpha,                # MSM
                     X = msm, y = y_death,         # ... observed data
                     link = "logit",               # ... logit link
                     distribution = "binomial",    # ... binomial dist
                     weights = ipw)                # ... weighted

    # Stacking the estimating equations and returning
    rbind(ee_msm, ee_num, ee_ps)
  },
  init = c(0, 0, 0, rep(0, ncol(W))),
  solver = "nleqslv",
  maxiter = 5000
)

# Causal odds ratio for qsmk
causal_or <- exp(cbind(estimate = coef(estr), confint(estr))[2, , drop = FALSE])
rownames(causal_or) <- "qsmk"
round(causal_or, 4)
```

The previous results are for the causal odds ratio. They are similar to
the book with slight differences in the confidence intervals (i.e., 1.0;
95% CI: 0.8, 1.4).

We can replicate this approach using the built-in `ee_ipw_msm`.

```{r}
#| label: ch12-4-binary-builtin
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Built-in estimating equation
    ee_ipw_msm(theta, y = y_death, A = a, W = W, V = msm,
               distribution = "binomial",
               link = "logit")
  },
  init = c(0, 0, init_ps),
  solver = "nleqslv",
  maxiter = 5000
)

# Causal odds ratio for qsmk
causal_or <- exp(cbind(estimate = coef(estr), confint(estr))[2, , drop = FALSE])
rownames(causal_or) <- "qsmk"
round(causal_or, 4)
```

### 12.5: Effect modification and marginal structural models

We will now use marginal structural models to study effect measure
modification. We will look at effect modification by sex (`sex`) of
quitting smoking (`qsmk`) on 10-year weight change (`wt82_71`).

```{r}
#| label: ch12-5
# Design matrix for propensity scores (already set)
# Design matrix for marginal structural model
dc$qsmk_sex <- dc$qsmk * dc$sex
msm_em <- as.matrix(dc[, c("I", "qsmk", "sex", "qsmk_sex")])

# Treatment and outcome (already set)
a <- dc$qsmk
y <- dc$wt82_71

estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters into corresponding estimation equations
    alpha <- theta[1:4]              # Marginal structural model
    gamma <- theta[5]                # Numerator parameter
    beta <- theta[6:length(theta)]   # Propensity score

    # Estimating the propensity scores using a logit model
    ee_ps <- ee_regression(theta = beta,         # Propensity score
                           X = W, y = a,         # ... observed data
                           model = "logistic")   # ... logit model
    pi_val <- inverse_logit(as.numeric(W %*% beta))

    # Estimating intercept-only for numerator
    I_dc <- as.matrix(dc[, "I", drop = FALSE])
    ee_num <- ee_regression(theta = gamma,            # Numerator model
                            X = I_dc, y = a,          # ... observed data
                            model = "logistic")       # ... logit model
    num_val <- inverse_logit(as.numeric(I_dc %*% gamma))

    # Stabilized inverse probability weights
    ipw <- ifelse(a == 1, num_val / pi_val, (1 - num_val) / (1 - pi_val))

    # Estimating the MSM using a weighted linear model
    ee_msm <- ee_regression(theta = alpha,        # Marginal structural model
                            X = msm_em, y = y,    # ... observed data
                            model = "linear",     # ... linear model
                            weights = ipw)        # ... weighted

    # Stacking the estimating equations and returning
    rbind(ee_msm, ee_num, ee_ps)
  },
  init = c(0, 0, 0, 0, init_num, init_ps),
  solver = "nleqslv",
  maxiter = 5000
)

# MSM with effect modification by sex
msm_est <- cbind(estimate = coef(estr), confint(estr))[1:4, ]
rownames(msm_est) <- c("alpha_0", "alpha_1", "alpha_2", "alpha_3")
round(msm_est, 2)
```

While not reported in the book, other online references report the
following coefficients: 1.7844, 3.5220, -0.0087, -0.1595.

### 12.6: Censoring and missing data

To conclude, we will now consider the missing outcomes that were ignored
earlier. To do this, we will use stabilized inverse probability of
missingness weights (IPCW in the book).

```{r}
#| label: ch12-6-setup
# Design matrix for propensity score model (using full df)
W_full <- as.matrix(df[, W_cols])

# Design matrix for missing model
X_cols <- c("I", "qsmk", "sex", "race", "age", "age_sq",
            "educ_2", "educ_3", "educ_4", "educ_5",
            "smokeintensity", "smokeintensity_sq",
            "smokeyrs", "smokeyrs_sq",
            "exer_1", "exer_2", "active_1", "active_2",
            "wt71", "wt71_sq")
X_miss <- as.matrix(df[, X_cols])

# Design matrix for marginal structural model
msm_full <- as.matrix(df[, c("I", "qsmk")])

# Treatment, outcome, and missing indicator
a_full <- df$qsmk
y_full <- df$wt82_71
r <- ifelse(is.na(df$wt82_71), 0, 1)
```

```{r}
#| label: ch12-6
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters up for their corresponding estimation equations
    alpha <- theta[1:2]                              # MSM
    gamma_n <- theta[3]                              # Numerator PS
    beta_n <- theta[4]                               # Numerator MW
    gamma_d <- theta[5:(4 + ncol(W_full))]           # Propensity score
    beta_d <- theta[(5 + ncol(W_full)):length(theta)] # Missing model

    I_full <- as.matrix(df[, "I", drop = FALSE])

    # Estimating the propensity scores using a logit model
    ee_ps <- ee_regression(theta = gamma_d,      # Propensity score
                           X = W_full, y = a_full, # ... observed data
                           model = "logistic")   # ... logit model
    pi_a <- inverse_logit(as.numeric(W_full %*% gamma_d))

    # Estimating intercept-only for numerator of IPTW
    ee_num <- ee_regression(theta = gamma_n,     # Numerator
                            X = I_full, y = a_full, # ... observed data
                            model = "logistic")  # ... logit model
    num_a <- inverse_logit(as.numeric(I_full %*% gamma_n))

    # Estimating the missing scores using a logit model
    ee_ms <- ee_regression(theta = beta_d,       # Missing score
                           X = X_miss, y = r,    # ... observed data
                           model = "logistic")   # ... logit model
    pi_m <- inverse_logit(as.numeric(X_miss %*% beta_d))

    # Estimating intercept-only for numerator of IPMW
    ee_sms <- ee_regression(theta = beta_n,      # Numerator
                            X = I_full, y = r,   # ... observed data
                            model = "logistic")  # ... logit model
    num_m <- inverse_logit(as.numeric(I_full %*% beta_n))

    # Stabilized inverse probability weights
    iptw <- ifelse(a_full == 1, num_a / pi_a, (1 - num_a) / (1 - pi_a))
    ipmw <- ifelse(r == 1, num_m / pi_m, 0)
    ipw <- iptw * ipmw

    # Estimating the MSM using a weighted linear model
    ee_msm <- ee_regression(theta = alpha,           # MSM
                            X = msm_full, y = y_full, # ... observed data
                            model = "linear",        # ... linear model
                            weights = ipw)           # ... weighted
    # Setting rows with missing Y's as zero (no contribution)
    ee_msm[is.na(ee_msm)] <- 0

    # Stacking the estimating equations and returning
    rbind(ee_msm, ee_num, ee_sms, ee_ps, ee_ms)
  },
  init = c(0, 0, 0, 0, rep(0, ncol(W_full)), rep(0, ncol(X_miss))),
  maxiter = 5000
)

msm_est <- cbind(estimate = coef(estr), confint(estr))[1:2, ]
rownames(msm_est) <- c("alpha_0", "alpha_1 (ACE)")
round(msm_est, 2)
```

Here, the book reports 3.5 (95% CI: 2.5, 4.5). This concludes chapter 12.

## Chapter 13: Standardization and the Parametric G-Formula

For Chapter 13, the book reviews the g-formula. Unlike IPW, the g-formula
relies on modeling the outcome process. The g-computation algorithm
estimator is
$$ \hat{E}[Y^a] = n^{-1} \sum_{i=1}^n m(a, W_i; \beta) $$
where $m$ is a statistical model for $E[Y | A, W]$ and $\beta$ are the
parameters defining the model. This version is slightly different from the
standardization form given in the book, but it is equivalent. Broadly, we
apply the g-computation algorithm via (1) estimate an outcome model, (2)
predict the outcomes had everyone been assigned $a$, and (3) compute the
mean of those predictions.

### 13.2: Estimating the mean outcome via modeling

First, we will fit a linear model for `wt82_71` conditional on `qsmk` and
the set of confounding variables. To begin, we will ignore the missing
outcomes.

```{r}
#| label: ch13-2-setup
# Design matrix for outcome model
X_out_cols <- c("I", "qsmk", "qsmk_smkint",
                "sex", "race", "age", "age_sq",
                "educ_2", "educ_3", "educ_4", "educ_5",
                "smokeintensity", "smokeintensity_sq",
                "smokeyrs", "smokeyrs_sq",
                "exer_1", "exer_2", "active_1", "active_2",
                "wt71", "wt71_sq")
X_out <- as.matrix(dc[, X_out_cols])
y <- dc$wt82_71
```

```{r}
#| label: ch13-2
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_regression(theta = theta, X = X_out, y = y, model = "linear")
  },
  # Naming the starting values names the estimates
  init = setNames(rep(0, ncol(X_out)), X_out_cols),
  solver = "nleqslv",
  maxiter = 5000
)

# Save starting values, with their names, for later use
init_reg <- estr@theta
```

### 13.3: Standardizing the mean outcome to the confounder distribution

Now we can apply the g-computation algorithm. To do this, we are going to
create a copy of our data set and set `qsmk=1` for all observations, then
repeat for `qsmk=0`.

```{r}
#| label: ch13-3-setup
# Copy of the data that we will update qsmk in
dca <- dc

# Setting qsmk to 1
dca$qsmk <- 1
dca$qsmk_smkint <- dca$qsmk * dca$smokeintensity
# Design matrix from qsmk=1 data
X1 <- as.matrix(dca[, X_out_cols])

# Setting qsmk to 0
dca$qsmk <- 0
dca$qsmk_smkint <- dca$qsmk * dca$smokeintensity
# Design matrix from qsmk=0 data
X0 <- as.matrix(dca[, X_out_cols])
```

```{r}
#| label: ch13-3
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters into corresponding estimation equations
    rd <- theta[1]
    r1 <- theta[2]
    r0 <- theta[3]
    beta <- theta[4:length(theta)]

    n <- length(y)

    # Estimating the linear model
    ee_reg <- ee_regression(theta = beta,       # Outcome model
                            X = X_out, y = y,   # ... observed data
                            model = "linear")   # ... linear model

    # Generating pseudo-outcomes using the model
    y1hat <- as.numeric(X1 %*% beta)  # Predicted Y when qsmk=1
    y0hat <- as.numeric(X0 %*% beta)  # Predicted Y when qsmk=0

    # Causal means
    ee_r1 <- matrix(y1hat - r1, nrow = 1)   # Causal mean for qsmk=1
    ee_r0 <- matrix(y0hat - r0, nrow = 1)   # Causal mean for qsmk=0

    # Average causal effect
    ee_rd <- matrix(rep((r1 - r0) - rd, n), nrow = 1)

    # Stacking the estimating equations and returning
    rbind(ee_rd, ee_r1, ee_r0, ee_reg)
  },
  init = c(rd = 0, r1 = 0, r0 = 0, init_reg),
  solver = "nleqslv",
  maxiter = 5000
)

gcomp <- cbind(estimate = coef(estr), confint(estr))[1:3, ]
rownames(gcomp) <- c("ACE", "E[Y^1]", "E[Y^0]")
round(gcomp, 2)
```

The first estimate is for the average causal effect (the second and third
are the causal means under all quit smoking and all don't quit smoking).
In the book, they report 3.5 (95% CI: 2.6, 4.5).

Rather than implement g-computation by-hand, we can also use the built-in
estimating equations.

```{r}
#| label: ch13-3-builtin
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Built-in g-formula estimating equation
    ee_gformula(theta, y = y, X = X_out, X1 = X1, X0 = X0)
  },
  init = c(rd = 0, r1 = 0, r0 = 0, init_reg),
  solver = "nleqslv",
  maxiter = 5000
)

gcomp <- cbind(estimate = coef(estr), confint(estr))[1:3, ]
rownames(gcomp) <- c("ACE", "E[Y^1]", "E[Y^0]")
round(gcomp, 2)
```

Which provides the same answers (as we would expect).

### Fine Point 13.2: A doubly robust estimator

The AIPW estimator combines IPW and the g-formula. Under the identifying
conditions, the AIPW estimator is doubly robust, meaning it is consistent
if either the propensity score model or the outcome model is correctly
specified (but not necessarily both). The canonical AIPW estimator is
$$ \hat{E}[Y^a] = n^{-1} \sum_{i=1}^n \frac{Y_i I(A_i = a)}{\pi_{A=a}(W_i)} + m(a, W_i; \beta) I(A_i) \frac{\pi_{A\ne a}(W_i)}{\pi_{A=a}(W_i)} $$
which can be implemented manually with the following estimating functions.

```{r}
#| label: ch13-fp-setup
# Refreshing design matrix for the propensity score model
W <- as.matrix(dc[, W_cols])
# Refreshing treatment array
a <- dc$qsmk
```

```{r}
#| label: ch13-fp-byhand
# The default solver converges reliably for the AIPW stacks in this section
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters into corresponding estimation equations
    ndim_X <- ncol(X_out)
    rd <- theta[1]
    r1 <- theta[2]
    r0 <- theta[3]
    beta <- theta[4:(3 + ndim_X)]
    alpha <- theta[(4 + ndim_X):length(theta)]

    # Estimating the propensity scores
    ee_ps <- ee_regression(theta = alpha,        # Estimate propensity scores
                           X = W, y = a,         # ... given observed A,W
                           model = "logistic")   # ... with logit model
    pi_val <- inverse_logit(as.numeric(W %*% alpha))

    # Estimating the outcome model
    ee_reg <- ee_regression(theta = beta,        # Outcome model
                            X = X_out, y = y,    # ... observed data
                            model = "linear")    # ... linear model
    y1_hat <- as.numeric(X1 %*% beta)            # Pseudo potential outcome
    y0_hat <- as.numeric(X0 %*% beta)            # Pseudo potential outcome

    n <- length(y)

    # Causal means
    ee_r1 <- matrix(
      (y * a / pi_val - y1_hat * (a - pi_val) / pi_val) - r1,
      nrow = 1
    )
    ee_r0 <- matrix(
      (y * (1 - a) / (1 - pi_val) + y0_hat * (a - pi_val) / (1 - pi_val)) - r0,
      nrow = 1
    )

    # Average causal effect
    ee_rd <- matrix(rep((r1 - r0) - rd, n), nrow = 1)

    # Stacking the estimating equations and returning
    rbind(ee_rd, ee_r1, ee_r0, ee_reg, ee_ps)
  },
  init = c(rd = 0, r1 = 0, r0 = 0, init_reg, init_ps),
  maxiter = 5000
)

# AIPW (by-hand)
aipw <- cbind(estimate = coef(estr), confint(estr))[1:3, ]
rownames(aipw) <- c("ACE", "E[Y^1]", "E[Y^0]")
round(aipw, 2)
```

Here, the AIPW results are similar to the previous g-computation and IPW
results.

We can also use the built-in functionality for the canonical AIPW
estimator (`ee_aipw`).

```{r}
#| label: ch13-fp-builtin
# The default solver again, as above
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_aipw(theta = theta, y = y, A = a, W = W,
            X = X_out, X1 = X1, X0 = X0)
  },
  init = c(rd = 0, r1 = 0, r0 = 0, init_ps, init_reg),
  maxiter = 5000
)

# AIPW (built-in)
aipw <- cbind(estimate = coef(estr), confint(estr))[1:3, ]
rownames(aipw) <- c("ACE", "E[Y^1]", "E[Y^0]")
round(aipw, 2)
```

These results match the by-hand implementation, as expected.

There is also an additional way to implement the AIPW estimator. This
approach fits a weighted outcome model, where the weights are from IPW.
This model is then used to predict values, like the standard g-computation
algorithm. This version is sometimes referred to as the weighted-regression
AIPW.

```{r}
#| label: ch13-fp-wregaipw
# The default solver again, as above
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters into corresponding estimation equations
    ndim_X <- ncol(X_out)
    rd <- theta[1]
    r1 <- theta[2]
    r0 <- theta[3]
    beta <- theta[4:(3 + ndim_X)]
    alpha <- theta[(4 + ndim_X):length(theta)]

    # Estimating the propensity scores
    ee_ps <- ee_regression(theta = alpha,        # Estimate propensity scores
                           X = W, y = a,         # ... given observed A,W
                           model = "logistic")   # ... with logit model
    pi_val <- inverse_logit(as.numeric(W %*% alpha))
    ipw <- 1 / ifelse(a == 1, pi_val, 1 - pi_val)

    # Estimating the outcome model
    ee_reg <- ee_regression(theta = beta,        # Outcome model
                            X = X_out, y = y,    # ... observed data
                            model = "linear",    # ... linear model
                            weights = ipw)       # ... weighted by IPW
    y1_hat <- as.numeric(X1 %*% beta)
    y0_hat <- as.numeric(X0 %*% beta)

    n <- length(y)

    # Causal means
    ee_r1 <- matrix(y1_hat - r1, nrow = 1)
    ee_r0 <- matrix(y0_hat - r0, nrow = 1)

    # Average causal effect
    ee_rd <- matrix(rep((r1 - r0) - rd, n), nrow = 1)

    # Stacking the estimating equations and returning
    rbind(ee_rd, ee_r1, ee_r0, ee_reg, ee_ps)
  },
  init = c(rd = 0, r1 = 0, r0 = 0, init_reg, init_ps),
  maxiter = 5000
)

# Weighted-regression AIPW
aipw <- cbind(estimate = coef(estr), confint(estr))[1:3, ]
rownames(aipw) <- c("ACE", "E[Y^1]", "E[Y^0]")
round(aipw, 2)
```

Note that these estimates differ slightly, but this is expected since it
solves a different equation. While not relevant in this setting, the
weighted-regression AIPW may be preferred over the canonical AIPW, since
the former is guaranteed to be bounded in the parameter space while the
latter is not.

As detailed in Shook-Sa et al. *Biometrics* 2025, the M-estimation setup
offers the additional feature that the AIPW estimators have doubly robust
point *and* variance estimation.

## Chapter 14: G-Estimation of Structural Nested Models

G-estimation differs from the previous approaches in what parameter it
targets. Rather than the average causal effect, we will estimate the
parameters of a structural nested model. The structural nested mean model
is
$$ E[Y^a - Y^{a=0} | A=a, W] = \varphi_0 a$$
Here, $\varphi_0$ represents the difference by $a$. However, we can also
study effect measure modification easily with structural nested models.
Consider
$$ E[Y^a - Y^{a=0} | A=a, W] = \varphi_0 a + \varphi_1 a V$$
where $V \in W$. Therefore, the structural nested model describes effect
measure modification by $V$.

### 14.5 G-estimation

We are going to use the estimating equations described in Technical
Point 14.2.

```{r}
#| label: ch14-5-setup
# Design matrix for propensity scores (using full df)
W_ge <- as.matrix(df[, W_cols])

# Design matrix for missing model
X_ge <- as.matrix(df[, X_cols])

# Design matrix for structural nested model
snm <- as.matrix(df[, "I", drop = FALSE])

# Treatment, outcome, and missing indicator
a_ge <- df$qsmk
y_ge <- df$wt82_71
r_ge <- ifelse(is.na(df$wt82_71), 0, 1)
```

```{r}
#| label: ch14-5-byhand
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters into corresponding estimation equations
    n_snm <- ncol(snm)
    n_W <- ncol(W_ge)
    alpha <- theta[1:n_snm]
    beta <- theta[(n_snm + 1):(n_snm + n_W)]
    gamma <- theta[(n_snm + n_W + 1):length(theta)]

    # Estimating equation for IPMW
    ee_ms <- ee_regression(theta = gamma,        # Missing score
                           X = X_ge, y = r_ge,   # ... observed data
                           model = "logistic")    # ... logit model
    pi_m <- inverse_logit(as.numeric(X_ge %*% gamma))
    ipmw <- r_ge / pi_m

    # Estimating equations for PS
    ee_log <- ee_regression(theta = beta,         # Propensity score
                            X = W_ge, y = a_ge,   # ... observed data
                            model = "logistic",   # ... logit model
                            weights = ipmw)       # ... weighted
    pi_val <- inverse_logit(as.numeric(W_ge %*% beta))

    # H(psi) equation for linear models
    h_psi <- y_ge - as.numeric((snm * a_ge) %*% alpha)

    # Estimating equation for the structural nested mean model
    ee_snm <- t(snm * (ipmw * h_psi * (a_ge - pi_val)))
    # Setting rows with missing Y's as zero (no contribution)
    ee_snm[is.na(ee_snm)] <- 0

    rbind(ee_snm, ee_log, ee_ms)
  },
  init = c(0, rep(0, ncol(W_ge)), rep(0, ncol(X_ge))),
  maxiter = 5000
)

snm_est <- cbind(estimate = coef(estr), confint(estr))[1, , drop = FALSE]
rownames(snm_est) <- "SNM phi_0"
round(snm_est, 2)
```

The book reported 3.4 (95% CI: 2.5, 4.5).

Now consider how we can use the built-in g-estimation functionality,
`ee_gestimation_snmm`.

```{r}
#| label: ch14-5-builtin
estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters into corresponding estimation equations
    n_snm <- ncol(snm)
    n_W <- ncol(W_ge)
    alpha <- theta[1:(n_snm + n_W)]
    gamma <- theta[(n_snm + n_W + 1):length(theta)]

    # Estimating equation for IPMW
    ee_ms <- ee_regression(theta = gamma,        # Missing score
                           X = X_ge, y = r_ge,   # ... observed data
                           model = "logistic")    # ... logit model
    pi_m <- inverse_logit(as.numeric(X_ge %*% gamma))
    ipmw <- r_ge / pi_m

    # Estimating equations for g-estimation
    ee_snm_val <- ee_gestimation_snmm(theta = alpha,
                                      y = y_ge, A = a_ge,
                                      W = W_ge, V = snm,
                                      weights = ipmw)
    # Setting rows with missing Y's as zero (no contribution)
    ee_snm_val[is.na(ee_snm_val)] <- 0

    rbind(ee_snm_val, ee_ms)
  },
  init = c(0, rep(0, ncol(W_ge)), rep(0, ncol(X_ge))),
  maxiter = 5000
)

snm_est <- cbind(estimate = coef(estr), confint(estr))[1, , drop = FALSE]
rownames(snm_est) <- "SNM phi_0 (built-in)"
round(snm_est, 2)
```

The structural nested model parameters match between implementations.

### 14.6: Structural nested models with two or more parameters

We now adapt the previous code to consider more than one parameter in the
structural nested model, adding `smokeintensity` as an effect modifier.

```{r}
#| label: ch14-6
# Update the structural nested model design matrix
snm2 <- as.matrix(df[, c("I", "smokeintensity")])

estr <- m_estimate(
  stacked_equations = function(theta) {
    # Dividing parameters into corresponding estimation equations
    n_snm <- ncol(snm2)
    n_W <- ncol(W_ge)
    alpha <- theta[1:n_snm]
    beta <- theta[(n_snm + 1):(n_snm + n_W)]
    gamma <- theta[(n_snm + n_W + 1):length(theta)]

    # Estimating equation for IPMW
    ee_ms <- ee_regression(theta = gamma,        # Missing score
                           X = X_ge, y = r_ge,   # ... observed data
                           model = "logistic")    # ... logit model
    pi_m <- inverse_logit(as.numeric(X_ge %*% gamma))
    ipmw <- r_ge / pi_m

    # Estimating equations for PS
    ee_log <- ee_regression(theta = beta,         # Propensity score
                            X = W_ge, y = a_ge,   # ... observed data
                            model = "logistic",   # ... logit model
                            weights = ipmw)       # ... weighted
    pi_val <- inverse_logit(as.numeric(W_ge %*% beta))

    # H(psi) equation for linear models
    h_psi <- y_ge - as.numeric((snm2 * a_ge) %*% alpha)

    # Estimating equation for the structural nested mean model
    ee_snm_val <- t(snm2 * (ipmw * h_psi * (a_ge - pi_val)))
    # Setting rows with missing Y's as zero
    ee_snm_val[is.na(ee_snm_val)] <- 0

    rbind(ee_snm_val, ee_log, ee_ms)
  },
  init = c(0, 0, rep(0, ncol(W_ge)), rep(0, ncol(X_ge))),
  maxiter = 5000
)

snm_est <- cbind(estimate = coef(estr), confint(estr))[1:2, ]
rownames(snm_est) <- c("SNM phi_0", "SNM phi_1")
round(snm_est, 2)
```

This provides the same results as the book (2.86 and 0.03). Unlike the
book, we also provide confidence intervals.

## Chapter 16: Instrumental Variable Analysis

Chapter 16 focuses on instrumental variable (IV) analysis. Here, we are
going to utilize an instrument (high state cigarette prices, variable name:
`highprice`) to estimate the effect of quitting smoking on weight gain.

Here, we will reload the data set, since the rows we need to drop due to
missing data differ from the previous examples.

```{r}
#| label: ch16-data
df_iv <- as.data.frame(causaldata::nhefs)
for (col in c("sex", "race", "education", "active", "exercise")) {
  df_iv[[col]] <- as.numeric(as.character(df_iv[[col]]))
}
df_iv$highprice <- as.integer(df_iv$price82 >= 1.5)
df_iv <- df_iv[complete.cases(df_iv[, c("wt82", "price82")]), ]
df_iv <- df_iv[, c("highprice", "qsmk", "wt82_71", "price82")]
df_iv$I <- 1

z_iv <- df_iv$highprice
a_iv <- df_iv$qsmk
y_iv <- df_iv$wt82_71
```

### 16.1: The three instrumental conditions

As described in the book, we will first examine the relationship between
our instrument $Z$ and the exposure $A$. Below we present estimating
equations for $\Pr(A=1 | Z=1) - \Pr(A=1 | Z=0)$.

```{r}
#| label: ch16-1
estr <- m_estimate(
  stacked_equations = function(theta) {
    n <- length(a_iv)
    ee_az1 <- matrix(z_iv * (a_iv - theta[1]), nrow = 1)
    ee_az0 <- matrix((1 - z_iv) * (a_iv - theta[2]), nrow = 1)
    ee_rd <- matrix(rep(theta[1] - theta[2] - theta[3], n), nrow = 1)
    rbind(ee_az1, ee_az0, ee_rd)
  },
  init = c(0.5, 0.5, 0)
)

iv_cond <- cbind(estimate = coef(estr), confint(estr))
rownames(iv_cond) <- c("Pr(A=1|Z=1)", "Pr(A=1|Z=0)", "Risk difference")
round(iv_cond, 3)
```

These are the same (point) estimates as those reported in the book. Here,
$Z$ is considered to be a *weak* instrument since the risk difference is
6.3%.

### 16.2: The usual IV estimand

The usual IV is defined as
$$ \beta = \frac{E[Y \mid Z=1] - E[Y \mid Z=0]}{E[A \mid Z=1] - E[A \mid Z=0]} $$

There are a few ways to implement this IV estimator with estimating
equations. The first is to estimate each of the pieces and combine them.

```{r}
#| label: ch16-2
estr <- m_estimate(
  stacked_equations = function(theta) {
    n <- length(y_iv)
    ee_uiv <- matrix(rep((theta[2] - theta[3]) - theta[1] * (theta[4] - theta[5]), n),
                     nrow = 1)
    ee_yz1 <- matrix(z_iv * (y_iv - theta[2]), nrow = 1)
    ee_yz0 <- matrix((1 - z_iv) * (y_iv - theta[3]), nrow = 1)
    ee_az1 <- matrix(z_iv * (a_iv - theta[4]), nrow = 1)
    ee_az0 <- matrix((1 - z_iv) * (a_iv - theta[5]), nrow = 1)
    rbind(ee_uiv, ee_yz1, ee_yz0, ee_az1, ee_az0)
  },
  init = c(1, 2, 2, 0.3, 0.2),
  solver = "nleqslv",
  maxiter = 5000
)

iv_est <- cbind(estimate = coef(estr), confint(estr))[1, , drop = FALSE]
rownames(iv_est) <- "IV estimate"
round(iv_est, 3)
```

The $Z$ conditional means match those reported in the book (2.686, 2.536,
0.2578, 0.1951), which gives us the same IV estimate as the book (2.4 kg).

Another way is shown in Boos & Stefanski (2013). The usual IV can be
expressed as a simpler pair of estimating equations.

```{r}
#| label: ch16-2-alt
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_uiv <- matrix((y_iv - a_iv * theta[1]) * (z_iv - theta[2]), nrow = 1)
    ee_z <- matrix(z_iv - theta[2], nrow = 1)
    rbind(ee_uiv, ee_z)
  },
  init = c(0, 0.5)
)

iv_est <- cbind(estimate = coef(estr), confint(estr))[1, , drop = FALSE]
rownames(iv_est) <- "IV estimate"
round(iv_est, 3)
```

There is also a built-in version of the usual IV estimator.

```{r}
#| label: ch16-2-builtin
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_iv_causal(theta, y = y_iv, A = a_iv, Z = z_iv)
  },
  init = c(0, 0.5)
)

iv_est <- cbind(estimate = coef(estr), confint(estr))[1, , drop = FALSE]
rownames(iv_est) <- "IV estimate (built-in)"
round(iv_est, 3)
```

As expected, the built-in implementation provides the same result.

In the book, another approach for estimation is also described: two-stage
least squares (2SLS). 2SLS works by estimating two linear regression
models. The first is for $A$ given $Z$. From that model, we get the
predicted value of $A$ given $Z$ and an intercept, denoted by $\hat{A}$.
Next, we fit a model for $Y$ given $\hat{A}$ and an intercept. The
coefficient for $\hat{A}$ from this model is the IV estimate.

```{r}
#| label: ch16-2-2sls-setup
Z_iv <- as.matrix(df_iv[, c("I", "highprice")])
```

```{r}
#| label: ch16-2-2sls
estr <- m_estimate(
  stacked_equations = function(theta) {
    beta <- theta[1:2]
    alpha <- theta[3:4]

    # First-stage regression
    ee_stage1 <- ee_regression(theta = alpha, y = a_iv, X = Z_iv, model = "linear")

    # Second-stage regression
    a_hat <- as.numeric(Z_iv %*% alpha)
    A_hat <- cbind(a_hat, rep(1, length(a_iv)))
    ee_stage2 <- ee_regression(theta = beta, y = y_iv, X = A_hat, model = "linear")

    # Returning stacked estimating equations
    rbind(ee_stage2, ee_stage1)
  },
  init = c(2, 2, 0.2, 0.1),
  solver = "nleqslv",
  maxiter = 5000
)

iv_est <- cbind(estimate = coef(estr), confint(estr))[1, , drop = FALSE]
rownames(iv_est) <- "2SLS IV estimate"
round(iv_est, 2)
```

The first element is the IV estimate. This matches the previous methods and
the book. The book reports their 95% CI as -36.5 to 41.3, which differs
slightly from what we have because we are using a different variance
estimator. The one reported in the book assumes homoskedasticity, whereas
the empirical sandwich variance estimator does not.

There is also a built-in procedure for 2SLS.

```{r}
#| label: ch16-2-2sls-builtin
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_2sls(theta, y = df_iv$wt82_71, A = df_iv$qsmk,
            Z = as.matrix(df_iv[, "highprice", drop = FALSE]),
            W = as.matrix(df_iv[, "I", drop = FALSE]))
  },
  init = c(2, 2, 0.2, 0.1),
  solver = "nleqslv",
  maxiter = 5000
)

iv_est <- cbind(estimate = coef(estr), confint(estr))[1, , drop = FALSE]
rownames(iv_est) <- "2SLS (built-in) IV estimate"
round(iv_est, 2)
```

The built-in procedure matches the by-hand implementation, as expected.
Note that the 2SLS implementation is more general and allows for one to
adjust for other variables or use multiple instruments.

### 16.5: The three instrument conditions revisited

Below is code to run the analysis at different thresholds for the 'high
price' cut-off, as described in the book.

```{r}
#| label: ch16-5
cutoffs <- c(1.60, 1.70, 1.80, 1.90)
by_cutoff <- data.frame(
  cutoff = cutoffs,
  iv = NA_real_,
  lower = NA_real_,
  upper = NA_real_
)

for (i in seq_along(cutoffs)) {
  z_new <- as.integer(df_iv$price82 >= cutoffs[i])
  estr <- m_estimate(
    stacked_equations = function(theta) {
      ee_2sls(theta, y = df_iv$wt82_71, A = df_iv$qsmk,
              Z = matrix(z_new, ncol = 1),
              W = as.matrix(df_iv[, "I", drop = FALSE]))
    },
    init = c(2, 2, 0.2, 0.1),
    solver = "nleqslv",
    maxiter = 5000
  )
  by_cutoff[i, c("iv", "lower", "upper")] <- c(coef(estr)[1], confint(estr)[1, ])
}

round(by_cutoff, 3)
```

These results are the same as those reported in the book (41.3, -40.9,
-21.1, -12.8).

## Chapter 17: Causal Survival Analysis

Replication of the following chapter 17 is not available yet.

## References

Hernan MA & Robins JM (2023). *Causal Inference: What If*. Boca Raton:
Chapman & Hall/CRC.

Shook-Sa BE, Zivich PN, Lee C, Xue K, Ross RK, Edwards JK, Stringer
JSA, & Cole SR. (2025). Double robust variance estimation with parametric
working models. *Biometrics*, 81(2), ujaf054.
