---
title: "Morris et al. (2022): Randomized Trials"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[Morris et al. (2022): Precision in Randomized Trials example](https://deli.readthedocs.io/en/latest/Examples/Morris-Trials-2022.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

The following is a replication of the analyses described in Morris et al. (2022),
which uses data from Wilson et al. (2017). Morris et al. examine three different
approaches to adjust for prognostic factors in randomized trials. The three ways
are direct adjustment, standardization, and inverse probability weighting. This
example highlights how causal inference methods (like those shown in Hernan &
Robins (2023)) can also be used with data from randomized trials.

Here, we are not going to consider direct adjustment. Attention is further
restricted to the any-test outcome. For finer details and comparisons between
the covariate adjustment approaches, see the original publication.

## Data

Data is available from Supplement S1 of Wilson et al. (2017) (referenced below).

```{r}
#| label: data
d <- get_tested
```

As done in the main paper, we conduct a complete-case analysis (i.e., discard
all the observations with missing outcomes).

```{r}
#| label: data-prep
ds <- d[complete.cases(d), ]

# Outcome
y_r <- ds$anytest

# Treatment
a_r <- ds$group

# Propensity score covariates: intercept + age + gender + msm + partners + ethnicgrp
# Create indicator columns for categorical variables
ps_mat <- model.matrix(
  ~ age + factor(gender) + factor(msm) + factor(partners) + factor(ethnicgrp),
  data = ds
)

# Outcome covariates: intercept + group + age + gender + msm + partners + ethnicgrp
oc_mat <- model.matrix(
  ~ group + age + factor(gender) + factor(msm) + factor(partners) + factor(ethnicgrp),
  data = ds
)

# Counterfactual design matrices
ds_a1 <- ds
ds_a1$group <- 1
oc_a1_mat <- model.matrix(
  ~ group + age + factor(gender) + factor(msm) + factor(partners) + factor(ethnicgrp),
  data = ds_a1
)

ds_a0 <- ds
ds_a0$group <- 0
oc_a0_mat <- model.matrix(
  ~ group + age + factor(gender) + factor(msm) + factor(partners) + factor(ethnicgrp),
  data = ds_a0
)
```

## Standardization

For standardization (or g-computation), we stack several estimating equations
together. Let $Y$ indicate the outcome, $A$ indicate the treatment, and $W$ be
the prognostic factors included in the model (age, gender, sexual orientation,
number of partners, ethnicity). The parameters are $\theta$, which we further
divide into the interest parameters ($\mu_0, \mu_1, \mu_2, \mu_3$) and the
nuisance parameters ($\beta$). As their name implies, nuisance parameters are
not of interest but are necessary for estimation of our interest parameters.
Importantly, by stacking the estimating equations, we can account for the
uncertainty in the estimation of the nuisance parameters into the uncertainty
in our interest parameters automatically.

The estimating functions are
$$\psi(Y_i, A_i, W_i; \theta) =
\begin{bmatrix}
    (\mu_1 - \mu_0) - \mu_2 \\
    \log(\mu_1 / \mu_0) - \mu_3 \\
    \hat{Y}_i^{a=1} - \mu_1 \\
    \hat{Y}_i^{a=0} - \mu_0 \\
    (Y_i - \text{expit}(g(A_i,W_i)^T \beta)) g(A_i,W_i)_i
\end{bmatrix}
$$

where $g(A_i,W_i)$ is a specific function to generate the design matrix, and
$\hat{Y}_i^{a} = \text{expit}(g(a,W_i)^T \beta)$. The first equation in the
stack is the risk difference (RD) and the second is the log-transformed risk
ratio (RR). Notice that these are defined in terms of other parameters.
Specifically, parameters from the third and fourth equations, which correspond
to the risk under all-1 and the risk under all-0. The predicted values of the
outcome are generated using the parameters from the final equation, which is
the logistic model used to estimate the probability of $Y_i$ conditional on
$A_i, W_i$.

```{r}
#| label: standardization
n_beta_oc <- ncol(oc_mat)

estr <- MEstimator(
  stacked_equations = function(theta) {
    # Extracting parameters from theta
    risk_diff <- theta[1]
    log_risk_ratio <- theta[2]
    risk_a1 <- theta[3]
    risk_a0 <- theta[4]
    beta <- theta[5:(4 + n_beta_oc)]
    n <- length(y_r)

    # Estimating nuisance model (outcome regression)
    ee_log <- ee_regression(
      theta = beta,
      X = oc_mat,
      y = y_r,
      model = "logistic"
    )

    # Generating predicted outcome values
    ee_ya1 <- matrix(
      inverse_logit(as.numeric(oc_a1_mat %*% beta)) - risk_a1,
      nrow = 1
    )
    ee_ya0 <- matrix(
      inverse_logit(as.numeric(oc_a0_mat %*% beta)) - risk_a0,
      nrow = 1
    )

    # Estimating interest parameters
    ee_risk_diff <- matrix(rep((risk_a1 - risk_a0) - risk_diff, n), nrow = 1)
    ee_risk_ratio <- matrix(rep(log(risk_a1 / risk_a0) - log_risk_ratio, n), nrow = 1)

    # Returning stacked estimating equations (order matters)
    rbind(
      ee_risk_diff,   # risk difference
      ee_risk_ratio,  # risk ratio
      ee_ya1,         # risk a=1
      ee_ya0,         # risk a=0
      ee_log          # logistic model
    )
  },
  init = c(0, 0, 0.5, 0.5, rep(0, n_beta_oc))
) |>
  estimate()

# Results
se <- sqrt(diag(estr@variance))
data.frame(
  Measure = c("Risk Difference", "Risk Ratio"),
  Estimate = round(estr@theta[1:2], 3),
  SE = round(se[1:2], 3)
)
```

## Inverse Probability Weighting

Now, we will consider the case of inverse probability weighting. Inverse
probability weighting relies on a separate nuisance model. To clarify this,
our nuisance parameters will be indicated by $\alpha$ here.

The estimating functions are
$$\psi(Y_i, A_i, W_i; \theta) =
\begin{bmatrix}
    (\mu_3 - \mu_4) - \mu_1 \\
    \log(\mu_3 / \mu_4) - \mu_2 \\
    \frac{Y_i A_i}{\text{expit}(g(W_i)^T \beta)} - \mu_3 \\
    \frac{Y_i (1-A_i)}{1 - \text{expit}(g(W_i)^T \beta)} - \mu_4 \\
    (A_i - \text{expit}(g(W_i)^T \beta)) g(W_i)
\end{bmatrix}
$$

where $g(W_i)$ is a specific function to generate the design matrix. As before,
the first and second equations are the risk difference and risk ratio,
respectively. Third and fourth equations correspond to the risk under all-1 and
the risk under all-0. The final estimating equation is the propensity score
model and is used to generate the propensity scores given $W_i$.

```{r}
#| label: ipw
n_beta_ps <- ncol(ps_mat)

estr <- MEstimator(
  stacked_equations = function(theta) {
    # Extracting parameters from theta
    risk_diff <- theta[1]
    log_risk_ratio <- theta[2]
    risk_a1 <- theta[3]
    risk_a0 <- theta[4]
    beta <- theta[5:(4 + n_beta_ps)]
    n <- length(a_r)

    # Estimating nuisance model (propensity score)
    ee_log <- ee_regression(
      theta = beta,
      X = ps_mat,
      y = a_r,
      model = "logistic"
    )

    # Generating predicted propensity score
    prop_score <- inverse_logit(as.numeric(ps_mat %*% beta))

    # Calculating weighted pieces
    ee_ya1 <- matrix((a_r * y_r) / prop_score - risk_a1, nrow = 1)
    ee_ya0 <- matrix(((1 - a_r) * y_r) / (1 - prop_score) - risk_a0, nrow = 1)

    # Estimating interest parameters
    ee_risk_diff <- matrix(rep((risk_a1 - risk_a0) - risk_diff, n), nrow = 1)
    ee_risk_ratio <- matrix(rep(log(risk_a1 / risk_a0) - log_risk_ratio, n), nrow = 1)

    # Returning stacked estimating equations (order matters)
    rbind(
      ee_risk_diff,   # risk difference
      ee_risk_ratio,  # risk ratio
      ee_ya1,         # risk a=1
      ee_ya0,         # risk a=0
      ee_log          # logistic model
    )
  },
  init = c(0, 0, 0.5, 0.5, rep(0, n_beta_ps))
) |>
  estimate()

# Results
se <- sqrt(diag(estr@variance))
data.frame(
  Measure = c("Risk Difference", "Risk Ratio"),
  Estimate = round(estr@theta[1:2], 3),
  SE = round(se[1:2], 3)
)
```

## Discussion

Results are the same as Morris et al. up to the third decimal place. We
replicated selected analyses in Morris et al. (2022) to showcase the basics of
stacking estimating equations and how they are implemented in `deli`. While the
package has built-in functionalities for standardization and inverse probability
weighting, we opted to present them by-hand here. By-hand takes some extra work
but can provide a better idea of how the package can be used to build estimating
equations. For example, each of the above approaches could further be extended
to account for missing outcomes under a weaker assumption.

**Differences in Results**

Why might results differ slightly from those in the original paper? The most
likely culprit is differences in the procedures for finding $\hat{\theta}$
across software.

`deli` operates by a root-finding procedure. Essentially, we are looking for
the values of $\hat{\theta}$ that lead to the estimating equations all being
approximately zero. Specifically, these examples use `deli`'s default solver,
the Newton-type `multiroot()` routine from the rootSolve package. A root-finding
procedure is used because it simplifies building stacked estimating equations.
However, this comes at the cost of speed and robustness of other approaches.

Other regression software tends to proceed under a different approach.
Generally, most software aims to maximize the log-likelihood. However, most
software for regression uses the iteratively reweighted least squares (IRLS)
algorithm. This algorithm is quite robust and fast, but is less general. While
these algorithms are expected to give very close answers, they can differ with
sparse data (e.g., small cells). As there is sparse data in this example (there
are many categorical variables for number of partners), this is the most likely
explanation for the small differences between results.

## References

Morris TP, Walker AS, Williamson EJ, & White IR. (2022). "Planning a method for
covariate adjustment in individually-randomised trials: a practical guide".
*Trials*, 23:328

Wilson E, Free C, Morris TP, Syred J, Ahamed I, Menon-Johansson AS et al.
(2017). "Internet-accessed sexually transmitted infection (e-STI) testing and
results service: a randomised, single-blind, controlled trial". *PLoS Medicine*,
14(12), e1002479.
