---
title: "Mroz (1987): Regression and Instrumental Variables"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[Mroz 1987: Regression and Instrumental Variables example](https://deli.readthedocs.io/en/latest/Examples/Mroz-1987.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

Mroz explored (mis)specification of statistical models using labor data on
married women from 1975. These data were used for a number of examples in the
book *Econometric Analysis of Cross Section and Panel Data* by Jeffrey
Wooldridge. Here some of these examples are shown further.

## Data

```{r}
#| label: data
d <- mroz
d <- d[complete.cases(d), ]
d$intercept <- 1
```

## Chapter 4: The Single-Equation Linear Model and OLS Estimation

Here, a simple model for the log-transformed wage (`lwage`) is fit as a
function of labor market experience (`exper`), years of schooling (`educ`),
age (`age`), number of kids 0-6 years old (`kidslt6`), and number of kids
6-18 years old (`kidsge6`). Fitting this model is easily done using the
built-in `ee_regression` functionality.

```{r}
#| label: ch4-ols
# Define the design matrix columns
design_cols <- c("intercept", "exper", "expersq", "educ",
                 "age", "kidslt6", "kidsge6")

# Define estimating equation and estimate
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_regression(theta,
                  X = as.matrix(d[, design_cols]),
                  y = d$lwage,
                  model = "linear")
  },
  init = rep(0, 7)
)

# Results
results <- data.frame(
  Param = design_cols,
  Coef  = estr@theta,
  SE    = sqrt(diag(estr@variance))
)
results[, -1] <- round(results[, -1], 3)
results
```

These results match those provided in the box. Note that the reported standard
error (SE) here corresponds to the heteroskedasticity-robust standard error
reported in the book.

## Chapter 5: Instrumental Variables Estimation of Single-Equation Linear Models

The next chapter considers instrumental variable estimation using 2-stage
least squares (2SLS). In particular we are interested in the effect of
education (`educ`) on log-transformed wages (`lwage`). Here, we will account
for labor market experience (`exper`) in both stages. The instruments in this
setting are mother's education (`motheduc`), father's education (`fatheduc`),
and husband's education (`huseduc`).

We will apply the 2SLS estimator using the built-in `ee_2sls` function.

```{r}
#| label: ch5-2sls
# Define estimating equation using built-in 2SLS. The 2SLS estimating
# functions are singular at an all-zero starting value, so we use the
# Levenberg-Marquardt solver ("lm"), which is also the default in Python
# delicatessen.
estr <- m_estimate(
  stacked_equations = function(theta) {
    ee_2sls(theta,
            y = d$lwage,
            A = d$educ,
            Z = as.matrix(d[, c("motheduc", "fatheduc", "huseduc")]),
            W = as.matrix(d[, c("intercept", "exper", "expersq")]))
  },
  init = rep(0, 1 + 3 + 3 * 2),
  solver = "lm"
)

# Results (first 4 parameters are the 2SLS coefficients)
labels_2sls <- c("educ", "intercept", "exper", "expersq")
results <- data.frame(
  Param = labels_2sls,
  Coef  = estr@theta[1:4],
  SE    = sqrt(diag(estr@variance))[1:4]
)
results[, -1] <- round(results[, -1], 3)
results
```

These results match the output provided in the book. Note that the order of
the output of `ee_2sls` is slightly different from the book.

## Example from `OneSampleMR`

As a final use of the Mroz data, we replicate the example with two action
variables (education and labor force experience) from the `OneSampleMR`
documentation, provided
[here](https://remlapmot.github.io/OneSampleMR/articles/f-statistic-comparison.html).
In this case, we will have age and the number of kids serve as the
instruments.

Currently, `ee_2sls` does not allow for multiple action variables. Therefore,
we instead code up the 2SLS estimator using the basic regression functions.
Briefly, we fit two models in the first stage (one for `educ` and one for
`exper`). Using the predicted values from these models, we then fit the second
stage model for `lwage`.

```{r}
#| label: onesamplemr
# Instrument design matrix
Z <- as.matrix(d[, c("intercept", "age", "kidslt6", "kidsge6")])

# Define the two-stage estimating equation by hand
estr <- m_estimate(
  stacked_equations = function(theta) {
    gamma <- theta[1:3]
    alpha <- theta[4:7]
    beta  <- theta[8:11]

    # First-stage regression for education
    ee_s1a <- ee_regression(alpha, X = Z, y = d$educ,
                            model = "linear")
    a_hat <- as.numeric(Z %*% alpha)

    # First-stage regression for experience
    ee_s1b <- ee_regression(beta, X = Z, y = d$exper,
                            model = "linear")
    b_hat <- as.numeric(Z %*% beta)

    # Second-stage regression for log(wage)
    Xhat <- cbind(d$intercept, a_hat, b_hat)
    ee_2s <- ee_regression(gamma, X = Xhat, y = d$lwage,
                           model = "linear")

    rbind(ee_2s, ee_s1a, ee_s1b)
  },
  init = rep(0, 3 + 4 * 2),
  solver = "lm"
)

# Results (first 3 parameters are second-stage coefficients)
labels_mr <- c("intercept", "educ", "exper")
results <- data.frame(
  Param = labels_mr,
  Coef  = estr@theta[1:3],
  SE    = sqrt(diag(estr@variance))[1:3]
)
results[, -1] <- round(results[, -1], 3)
results
```

These results mostly match those provided in the documentation. However, note
that the SE differs due to the use of a different variance estimator.

## References

Mroz, T. A. (1987). The sensitivity of an empirical model of married women's
hours of work to economic and statistical assumptions. *Econometrica* 55(4),
765-799.

Wooldridge, J. M. (2010). *Econometric analysis of cross section and panel
data*. MIT press.
