---
title: "Optimization advice"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

::: {.callout-note}
This article is translated from the
[Optimization Advice page](https://deli.readthedocs.io/en/latest/Optimization%20Advice.html)
in the documentation of [delicatessen](https://deli.readthedocs.io/), deli's
Python counterpart.
:::

```{r}
#| label: setup
library(deli)
```

This article offers guidance for when root-finding or minimization is difficult
for a given set of estimating functions. The flexibility of the generic
estimating-equation interface comes at a cost: deli does not provide the fastest
or most specialized routines for any one model, because it aims to solve them all.
When a fit stalls, diverges, or returns a variance full of `NaN`, the remedies
below usually help. Each is grounded in how the solve actually behaves.

The controls referenced throughout are `init`, `solver`, `maxiter`, `tolerance`,
`deriv_method`, and `dx`. `m_estimate()` takes all of them in a single call. In
the two-step form, `init` belongs to `MEstimator()` and travels with the object,
while the rest are arguments of `estimate()`.

## Choose good starting values

Optimization improves when the starting values in `init` are close to the true
parameter vector. We rarely know those values in advance, so the practical rule
is to start in a reasonable part of the parameter space. Three properties matter:

1. The starting values are numerically sensible.
2. They lie inside the range of possible parameter values. A proportion is bound
   to $[0, 1]$, so an initial value of $-10$ can break the search. A variance is
   non-negative, so it should not start below zero.
3. They are as close to the estimate as available information allows.

For most regression coefficients, a starting value of zero is a reasonable
default in the absence of other information. Some estimating equations are far
more sensitive to the starting point than a linear model is. Robust regression,
the tobit model, and accelerated failure time models can stall or diverge when
started far from the solution, because their estimating functions are flat or
steep in regions away from the optimum.

`init` is supplied when the estimator is constructed and travels with the
object. Constructing one without solving it and reading `init` back off the
object shows this directly:

```{r}
#| label: init-example
data(robust_regress)
robust_regress$height_c <- robust_regress$height - mean(robust_regress$height)

X <- cbind(1, robust_regress$height_c)
y <- robust_regress$weight

psi_robust <- function(theta) {
  ee_robust_regression(theta, X = X, y = y, model = "linear", k = 1.345)
}

# Seed the intercept at the mean of the outcome and the slope at zero
m <- MEstimator(psi_robust, init = c(mean(y), 0))

m@init
```

## Seed from a simpler fit

When an estimator stacks several estimating functions, some parameters can be
estimated on their own, outside the joint solve. Fitting those pieces first and
passing the results as `init` lets the joint optimization concentrate on the
parameters that cannot be solved separately. This is especially useful for
regression models, where dedicated and numerically stable fitting routines exist:
seeding `init` with the coefficients from `stats::lm()` or `stats::glm()` lets
deli borrow the strength of those routines.

A lighter version of the same idea does not require solving any coefficients.
For a regression model, set the intercept to the mean of the outcome (or its
link-transformed value for a generalized linear model) and leave the remaining
coefficients at zero. The seed in the previous section does exactly this.

`m_estimate()` and `gmm_estimate()` also accept a `subset` argument, which holds
some parameters fixed at their initial values and solves only for the rest. When
an estimator has hundreds of parameters and many can be pinned down outside the
procedure, restricting the search to the remaining `subset` reduces the problem
the solver has to face. The indices are one-based in deli.

## Increase the iteration budget

If better starting values are not enough, raise the iteration budget. `maxiter`
defaults to `5000`; increasing it lets the solver take more steps before giving
up.

```{r}
#| label: maxiter-example
#| eval: false
m <- m_estimate(psi_robust, init = c(mean(y), 0), maxiter = 20000)
```

## Try a different solver

deli offers three built-in solvers for `MEstimator`, selected through the
`solver` argument. Each has different operating characteristics, and a problem
that defeats one may yield to another.

| `solver` | Routine | Method | Notes |
|---|---|---|---|
| `"rootSolve"` | [rootSolve::multiroot()] | Newton | Default. No extra package required. |
| `"nleqslv"` | [nleqslv::nleqslv()] | Newton with line search and trust region | Alternative global strategies. |
| `"lm"` | [minpack.lm::nls.lm()] | Levenberg-Marquardt | Damped least squares; robust to a singular Jacobian at the start. |

`"rootSolve"` is the default. It applies a Newton method and converges on the
great majority of well-conditioned square systems without any additional
package. `"nleqslv"` is an alternative Newton solver with line-search and
trust-region strategies that can succeed where a plain Newton step overshoots.
`"lm"` is the damped least-squares Levenberg-Marquardt algorithm from MINPACK,
the same routine Python delicatessen calls by default; the damping makes it
tolerant of an ill-conditioned or singular Jacobian near the starting point,
where the Newton solvers stall.

The robust regression above is a concrete case. Started from the mean-of-outcome
seed, the Newton default makes almost no progress, because near that point the
Huber score clips most residuals and the Jacobian is nearly singular. The
Levenberg-Marquardt solver converges from the same seed.

```{r}
#| label: solver-comparison
m_newton <- MEstimator(psi_robust, init = c(mean(y), 0)) |>
  estimate(solver = "rootSolve")
m_lm <- MEstimator(psi_robust, init = c(mean(y), 0)) |>
  estimate(solver = "lm")

coef(m_newton)
coef(m_lm)
```

The Newton fit has barely moved from the seed, while the Levenberg-Marquardt fit
has reached the robust solution. The next section shows how to tell the two
outcomes apart without knowing the answer in advance.

For `GMMEstimator`, the solve is a minimization rather than a root-finding
problem, and `solver` accepts any method name understood by [stats::optim()].
The default is `"BFGS"`.

A custom solver is also accepted. Pass a function that takes `stacked_equations`
and `init` arguments and returns the solved parameter vector; some problems have
a bespoke algorithm that outperforms the general-purpose routines.

## Scale the predictors

When covariates sit on very different numeric scales, the bread matrix (the
derivative of the estimating functions) becomes ill-conditioned, which slows
convergence and degrades the numerical stability of the bread inversion that
produces the variance. Centering and scaling the predictors before fitting keeps
the bread well-conditioned.

The effect is easy to see in the condition number of the bread, where a larger
value means a matrix that is closer to singular and harder to invert reliably:

```{r}
#| label: scaling-example
set.seed(1)
n <- 200
x1 <- rnorm(n, mean = 0, sd = 1)
x2 <- rnorm(n, mean = 0, sd = 1e4)
y_sim <- 1 + 0.5 * x1 + 1e-4 * x2 + rnorm(n)

# Raw predictors, on wildly different scales
X_raw <- cbind(1, x1, x2)
fit_raw <- m_estimate(
  function(theta) ee_regression(theta, X = X_raw, y = y_sim, model = "linear"),
  init = c(0, 0, 0)
)
kappa(fit_raw@bread, exact = TRUE)

# Standardized predictors
X_scaled <- cbind(1, scale(x1), scale(x2))
fit_scaled <- m_estimate(
  function(theta) ee_regression(theta, X = X_scaled, y = y_sim, model = "linear"),
  init = c(0, 0, 0)
)
kappa(fit_scaled@bread, exact = TRUE)
```

Standardizing the predictors reduces the condition number by many orders of
magnitude. The point estimates and their standard errors are on the scaled
coefficients, so translate them back to the original units afterward if the raw
scale is the one you want to report.

## Choose a derivative method

The bread is a derivative, and deli computes it by the method named in
`deriv_method`.

| `deriv_method` | Meaning |
|---|---|
| `"capprox"` | Central finite difference (default). |
| `"fapprox"` | Forward finite difference. |
| `"bapprox"` | Backward finite difference. |
| `"exact"` | Forward-mode automatic differentiation. |

`"capprox"`, the default, is a central finite difference. It is accurate and
robust, and it is the recommended starting point. `"fapprox"` and `"bapprox"`
are one-sided differences, useful when the estimating function cannot be
evaluated on one side of the current parameter value. The `dx` argument sets the
finite-difference step for all three; it defaults to `1e-9` and is the one tuning
knob these methods expose.

`"exact"` is forward-mode automatic differentiation. It carries derivatives
through the computation exactly, so it is the most accurate option and has no
step size to choose. Exact differentiation now works for every built-in
estimating equation except the two that are not differentiable, `ee_percentile`
and `ee_positive_mean_deviation`. It requires that the estimating function be
built from operations the automatic differentiation supports; the pre-built
equations satisfy this, and custom equations do as long as they use supported
operations.

```{r}
#| label: deriv-example
m_exact <- m_estimate(
  psi_robust,
  init = c(mean(y), 0),
  solver = "lm",
  deriv_method = "exact"
)
sqrt(diag(vcov(m_exact)))
```

The derivative method affects only the bread, and therefore the variance; it
does not change the point estimate.

## Diagnose convergence

deli does not print a running log while it solves, so convergence is confirmed
after the fit. Three checks cover most cases.

**The estimating functions should be near zero at the solution.** A root of the
estimating equations makes the estimating functions sum to approximately zero
across observations. The estimating-function matrix is parameters-by-observations,
so this sum is the row sums, `rowSums(psi(...))`. Evaluating the stacked equations
at the fitted parameters and summing across observations reveals a solver that
stopped short:

```{r}
#| label: diagnostic-efunc
# Stalled Newton fit from the solver comparison above
rowSums(psi_robust(coef(m_newton)))

# Converged Levenberg-Marquardt fit
rowSums(psi_robust(coef(m_lm)))
```

The Newton values are far from zero, which flags the stalled fit; the
Levenberg-Marquardt values are zero to numerical precision.

**The built-in solvers warn when they stop without solving the equations.** Each
routine reports a status code, but the code on its own is not the test. Some of
nleqslv's codes mean only that the function values are left for the caller to
check, minpack.lm reports a stall as a convergence test that was met, and
rootSolve reports success even where its own convergence test failed. deli
therefore judges the returned point as well, and warns when the estimating
functions are not solved there. A warning of that kind is a signal to revisit the
starting values, the scaling, or the solver.

`gmm_estimate()` is judged the same way, on criteria suited to what it does. It
minimizes a quadratic in the summed moments rather than root-finding, and
`optim()` reports convergence on the flat tail of that objective, well before the
moments reach zero. Where the moment conditions are as many as the parameters
they have to vanish at a solution, so the returned point is judged twice.

The first check asks whether any single moment condition is in a state no root
can be in. For a condition whose per-observation contributions vary, that means
they barely cancel: contributions that all point the same way cannot sum to zero,
however small the sum happens to be. For a condition with the same value at every
observation, which is how a relation among the parameters is written, it means
that value does not vanish once measured against the scale of the problem. This is
the warning most fits that fail meet, and the message names the condition at
fault.

The second check measures the Newton step the returned point would still take
towards a root and warns when that step is larger than the estimates themselves.
It is what catches a fit built from a design matrix, where the contributions are
mixed-sign and so cancel well however wrong the parameters are.

An over-identified fit cannot drive every moment to zero and a `subset` fit holds
some parameters fixed, so neither is judged either way; for those, inspect
`rowSums()` of the estimating functions as above.

**The variance should be finite and of a reasonable magnitude.** A singular or
near-singular bread produces a variance full of `NaN`, `Inf`, or implausibly
large entries. Inspecting the standard errors catches this:

```{r}
#| label: diagnostic-variance
sqrt(diag(vcov(m_lm)))
```

If these are `NaN` or enormous, the bread could not be inverted reliably.
Scaling the predictors, improving the starting values, or checking the model
specification for a redundant parameter usually resolves it.

## A word on tolerance

The `tolerance` argument controls how close to a root the solver must get before
it stops; it defaults to `1e-9`. It is tempting to loosen `tolerance` to force a
fit to succeed, but this only permits a larger optimization error: the solver
stops further from the true root of the estimating equation, and the point
estimate and variance inherit that error. Do not loosen `tolerance` to get a fit
to converge unless you are certain the looser value is within acceptable error
for your problem. Better starting values, scaling, or a different solver address
the cause rather than the symptom.
