---
title: "Ross et al. (2024): Introduction to M-estimation"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(deli)
```

The following is a replication of the cases described in Ross et al. (2024).
The original paper provides a tutorial on M-estimation with introductory examples
in the context of regression, standardization, and measurement error.

## Data

The following data from Table 1 is used for the first two examples.

```{r}
#| label: data
# From Table 1
d <- data.frame(
  X = c(0, 0, 0, 0, 1, 1, 1, 1),
  W = c(0, 0, 1, 1, 0, 0, 1, 1),
  Y = c(0, 1, 0, 1, 0, 1, 0, 1),
  n = c(496, 74, 113, 25, 85, 15, 15, 3)
)

# Expand rows by count
d <- d[rep(seq_len(nrow(d)), d$n), ]
d$intercept <- 1
d$n <- NULL
rownames(d) <- NULL
```

## Example 1: Logistic Regression

Fit a logistic regression model for $Y$ given $X, W$.

```{r}
#| label: ex1
X <- as.matrix(d[, c("intercept", "X", "W")])
y <- d$Y

estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_regression(theta, X = X, y = y, model = "logistic")
  },
  init = c(0, 0, 0)
) |>
  estimate()

# Results
data.frame(
  Param = c("beta_0", "beta_1", "beta_2"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

## Example 2: Marginal risk difference

### G-computation

Using a logistic model, generate predictions to estimate the marginal risk
difference by $X$ via g-computation.

```{r}
#| label: ex2a
X <- as.matrix(d[, c("intercept", "X", "W")])
y <- d$Y

# Counterfactual design matrices
X1 <- X0 <- X
X1[, "X"] <- 1
X0[, "X"] <- 0

estr <- MEstimator(
  stacked_equations = function(theta) {
    beta <- theta[1:3]
    mu0 <- theta[4]
    mu1 <- theta[5]
    delta1 <- theta[6]
    n <- length(y)

    # Logistic regression for outcome
    ee_logit <- ee_regression(beta, X = X, y = y, model = "logistic")

    # Predicted risks under each treatment
    y1_hat <- inverse_logit(as.numeric(X1 %*% beta))
    y0_hat <- inverse_logit(as.numeric(X0 %*% beta))

    # Estimating equations for causal means and contrast
    ee_r0 <- matrix(y0_hat - mu0, nrow = 1)
    ee_r1 <- matrix(y1_hat - mu1, nrow = 1)
    ee_rd <- matrix(rep((mu1 - mu0) - delta1, n), nrow = 1)

    rbind(ee_logit, ee_r0, ee_r1, ee_rd)
  },
  init = c(0, 0, 0, 0.5, 0.5, 0)
) |>
  estimate()

data.frame(
  Param = c("beta_0", "beta_1", "beta_2", "mu_0", "mu_1", "delta"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

### Inverse probability weighting

Now estimate the marginal risk difference using inverse probability weighting.

```{r}
#| label: ex2b
a <- d$X
W <- as.matrix(d[, c("intercept", "W")])
y <- d$Y

estr <- MEstimator(
  stacked_equations = function(theta) {
    alpha <- theta[1:2]
    mu0 <- theta[3]
    mu1 <- theta[4]
    delta1 <- theta[5]
    n <- length(y)

    # Logistic regression for propensity score
    ee_logit <- ee_regression(alpha, X = W, y = a, model = "logistic")

    # Propensity score and weights
    pscore <- inverse_logit(as.numeric(W %*% alpha))
    wt <- a / pscore + (1 - a) / (1 - pscore)

    # Weighted estimating equations for causal means
    ee_r1 <- matrix(a * y * wt - mu1, nrow = 1)
    ee_r0 <- matrix((1 - a) * y * wt - mu0, nrow = 1)
    ee_rd <- matrix(rep((mu1 - mu0) - delta1, n), nrow = 1)

    rbind(ee_logit, ee_r0, ee_r1, ee_rd)
  },
  init = c(0, 0, 0.5, 0.5, 0)
) |>
  estimate()

data.frame(
  Param = c("alpha_0", "alpha_1", "mu_0", "mu_1", "delta"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

## Example 3: Outcome misclassification

Correct for mismeasurement of the outcome variable with the Rogan-Gladen estimator.

```{r}
#| label: ex3-data
# Misclassification data
d3 <- data.frame(
  R = c(1, 1, 0, 0, 0, 0),
  Y = c(0, 0, 1, 1, 0, 0),
  W = c(1, 0, 1, 0, 1, 0),
  n = c(680, 270, 204, 38, 18, 71)
)

d3 <- d3[rep(seq_len(nrow(d3)), d3$n), ]
d3$n <- NULL
rownames(d3) <- NULL
```

```{r}
#| label: ex3
estr <- MEstimator(
  stacked_equations = function(theta) {
    ee_rogan_gladen(theta, y = d3$Y, y_star = d3$W, r = d3$R)
  },
  init = c(0.75, 0.75, 0.75, 0.75)
) |>
  estimate()

data.frame(
  Param = c("Corrected", "Mismeasured", "Sensitivity", "Specificity"),
  Coef = round(estr@theta, 2),
  LCL = round(confint(estr)[, 1], 2),
  UCL = round(confint(estr)[, 2], 2)
)
```

## References

Ross RK, Zivich PN, Stringer JS, & Cole SR. (2024). M-estimation for common
epidemiological measures: introduction and applied examples. *International
Journal of Epidemiology*, 53(2), dyae030.
