---
title: "Translating from Python delicatessen"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(deli)

# The Python blocks below run against a project-local virtual environment
# containing delicatessen and NumPy.
reticulate::use_virtualenv("../../.venv", required = TRUE)
```

deli is a faithful R port of the Python
[delicatessen](https://deli.readthedocs.io/) library for M-estimation and the
empirical sandwich variance estimator. Every capability in delicatessen has a
counterpart in deli, and the two produce the same numbers when the same settings
are used. The surface is idiomatic R rather than a line-for-line transcription:
some functions are renamed to avoid clashing with base R, a few arguments carry
different names, and a small number of defaults differ. This article maps the
two libraries onto each other and gives the exact recipes for reproducing a
Python result in R.

Python users are not expected to paste Python code into R and have it run. The
goal is functional parity: the same models, the same options, and matching point
estimates and variances under matched settings.

## How the libraries line up

Both libraries are built on the same idea. You supply a function that returns a
stacked matrix of estimating-equation contributions, one column per observation
and one row per parameter. An estimator object solves that system for the point
estimate and assembles the sandwich variance from the bread (the derivative of
the estimating functions) and the meat (their outer product).

The central objects correspond directly:

- `delicatessen.MEstimator` becomes the `MEstimator` S7 class in deli, solved
  with the `estimate()` generic. deli additionally offers `m_estimate()`, a
  formula and one-step convenience wrapper with no Python equivalent.
- `delicatessen.GMMEstimator` becomes `GMMEstimator`, with the same
  `estimate()` generic and an optional `gmm_estimate()` wrapper.

In Python you call methods on the estimator (`m.estimate()`,
`m.confidence_intervals()`). In deli these are generic functions that take the
estimator as their first argument (`estimate(m)`, `confidence_intervals(m)`).

## Class and function correspondence

| Python (delicatessen) | R (deli) | Notes |
|---|---|---|
| `MEstimator(stacked_equations, init, subset, finite_correction)` | `MEstimator(stacked_equations, init, subset, finite_correction)` | `subset` uses 1-based parameter indices in R (see below). |
| `m.estimate(solver, maxiter, tolerance, deriv_method, dx, allow_pinv)` | `estimate(m, solver, maxiter, tolerance, deriv_method, dx, allow_pinv)` | Different `solver` and `deriv_method` defaults (see below). |
| `GMMEstimator(...)` | `GMMEstimator(...)` | R adds a `weight_matrix` property. |
| `g.estimate(solver='bfgs', ...)` | `estimate(g, solver = "BFGS", ...)` | R accepts any `stats::optim()` method name. |
| `m.confidence_intervals(alpha)` | `confidence_intervals(m, alpha)` | Both switch to the t-distribution when `finite_correction` is set. |
| `m.z_scores(null)` | `z_scores(m, null)` | |
| `m.p_values(null)` | `p_values(m, null)` | |
| `m.s_values(null)` | `s_values(m, null)` | |
| `m.influence_functions()` | `influence_functions(m)` | |
| `m.confidence_bands(subset, alpha, method, n_draws, seed)` | `confidence_bands(m, alpha, method, n_draws, seed)` | Different `n_draws` default (see below). |
| `m.print_results(...)` | `print(m)` / `summary(m)` | Idiomatic replacement; `summary(m, alpha = ...)` supported. |
| `compute_sandwich(...)` | `compute_sandwich(...)` | Exported in both. |
| `delta_method(theta, g, covariance, deriv_method, dx)` | `delta_method(object, transform, covariance, deriv_method, dx)` | `g` renamed `transform`; different `deriv_method` default. |
| `compute_confidence_bands(...)` | `compute_confidence_bands(...)` | |
| `logit`, `inverse_logit` | `logit`, `inverse_logit` | |
| `identity(value)` | `identity_transform(value)` | Renamed to avoid masking base R `identity()`. |
| `polygamma(n, x)` | `deli_polygamma(n, x)` | Renamed to avoid a namespace clash. |
| `digamma(z)` | `deli_digamma(z)` | Renamed to avoid masking base R `digamma()`. At the poles (non-positive integers) `deli_digamma()` returns `NaN`, matching base R `digamma()`; SciPy returns `-Inf`. `NA` and `NaN` inputs propagate unchanged. |
| `standard_normal_cdf`, `standard_normal_pdf` | `standard_normal_cdf`, `standard_normal_pdf` | |
| `robust_loss_functions(residual, loss, k, a, b)` | `robust_loss_functions(residuals, loss, k)` | Hampel thresholds passed as `k = c(a, b, c)` (see below). |
| `aggregate_efuncs(est_funcs, group)` | `aggregate_efuncs(est_funcs, group)` | |
| `regression_predictions(...)` | `regression_predictions(...)` | Returns a data frame. |
| `aft_predictions_individual(...)` | `aft_predictions_individual(...)` | |
| `aft_predictions_function(...)` | `aft_predictions_function(...)` | |
| `spline(...)` | `deli_spline(...)` | Renamed to avoid masking `stats::spline()`. |
| `additive_design_matrix(...)` | `additive_design_matrix(...)` | |

The forward-difference and autodiff helpers `approx_differentiation()` and
`auto_differentiation()` are public in Python. In deli they back the
`deriv_method` argument of `estimate()`, `compute_sandwich()`, and
`delta_method()` and are not exported as user-facing functions. Choose the
derivative method through `deriv_method` rather than calling them directly.

## Estimating-equation correspondence

All 36 pre-built estimating equations in delicatessen are available in deli
under the same `ee_*` names. Most take identical argument names. The exceptions
are listed in the final column.

| Python | R | Argument differences |
|---|---|---|
| `ee_mean` | `ee_mean` | |
| `ee_mean_variance` | `ee_mean_variance` | |
| `ee_mean_robust` | `ee_mean_robust` | Hampel thresholds via `k = c(a, b, c)`. |
| `ee_mean_geometric` | `ee_mean_geometric` | |
| `ee_percentile` | `ee_percentile` | |
| `ee_positive_mean_deviation` | `ee_positive_mean_deviation` | |
| `ee_regression` | `ee_regression` | |
| `ee_glm` | `ee_glm` | |
| `ee_mlogit` | `ee_mlogit` | |
| `ee_robust_regression` | `ee_robust_regression` | Hampel thresholds via `k = c(a, b, c)`. |
| `ee_ridge_regression` | `ee_ridge_regression` | |
| `ee_lasso_regression` | `ee_lasso_regression` | |
| `ee_dlasso_regression` | `ee_dlasso_regression` | |
| `ee_elasticnet_regression` | `ee_elasticnet_regression` | |
| `ee_bridge_regression` | `ee_bridge_regression` | |
| `ee_beta_regression` | `ee_beta_regression` | |
| `ee_tobit` | `ee_tobit` | |
| `ee_additive_regression` | `ee_additive_regression` | |
| `ee_ipw` | `ee_ipw` | |
| `ee_ipw_msm` | `ee_ipw_msm` | |
| `ee_gformula` | `ee_gformula` | |
| `ee_aipw` | `ee_aipw` | |
| `ee_gestimation_snmm` | `ee_gestimation_snmm` | |
| `ee_2sls` | `ee_2sls` | |
| `ee_iv_causal` | `ee_iv_causal` | |
| `ee_mean_sensitivity_analysis` | `ee_mean_sensitivity_analysis` | |
| `ee_survival_model(theta, t, delta, distribution)` | `ee_survival_model(theta, time, event, distribution)` | `t` becomes `time`, `delta` becomes `event`. |
| `ee_aft(theta, X, t, delta, distribution, weights)` | `ee_aft(theta, X, time, event, distribution, ...)` | `t` becomes `time`, `delta` becomes `event`; R adds `offset`. |
| `ee_plogit(theta, X, t, delta, S, ...)` | `ee_plogit(theta, X, time, event, S, ...)` | `t` becomes `time`, `delta` becomes `event`; R adds `offset`. |
| `ee_rogan_gladen` | `ee_rogan_gladen` | |
| `ee_rogan_gladen_extended` | `ee_rogan_gladen_extended` | |
| `ee_regression_calibration` | `ee_regression_calibration` | |
| `ee_emax(theta, dose, response, robust, k)` | `ee_emax(theta, dose, response, loss, k)` | `robust` becomes `loss`. |
| `ee_emax_ed` | `ee_emax_ed` | |
| `ee_loglogistic(theta, dose, response, robust, k)` | `ee_loglogistic(theta, dose, response, loss, k)` | `robust` becomes `loss`. |
| `ee_loglogistic_ed` | `ee_loglogistic_ed` | |

The generalized linear model `ee_glm` carries the full set of Python options,
including the `hyperparameter` argument, the `tweedie` and `negative_binomial`
distributions, and the gamma distribution's nuisance shape equation. As in
Python, the gamma and negative binomial fits append an extra parameter to
`theta` (the log shape or log dispersion), so the parameter vector is one
element longer than the number of regression coefficients.

## Argument renames and default differences

This section documents every difference that affects the numbers or the call,
with the exact recipe for reproducing a default Python run in R. Unless noted,
each recipe reproduces Python to machine precision.

### Solver default

Python's `MEstimator.estimate()` defaults to `solver='lm'`, the
Levenberg-Marquardt routine from MINPACK exposed through
`scipy.optimize.root(method='lm')`. deli defaults to `solver = "rootSolve"`
([rootSolve::multiroot()]), which requires no additional package and converges
reliably on the great majority of problems.

deli also provides the identical Levenberg-Marquardt solver under
`solver = "lm"` (backed by `minpack.lm`, the same MINPACK code Python calls). To
reproduce a default Python `MEstimator` fit, set the solver explicitly:

```r
m <- MEstimator(psi, init = c(0, 0)) |>
  estimate(solver = "lm")
```

The `"lm"` solver is the recommended choice for problems where a Newton-type
solver stalls, in particular Huber robust regression started far from the
solution: when every residual is clipped, the bread is identically zero and
`rootSolve` and `nleqslv` cannot make progress, while the damped
Levenberg-Marquardt step still converges.

Python's `GMMEstimator.estimate()` defaults to `solver='bfgs'`. deli defaults to
`solver = "BFGS"` for `GMMEstimator`, so the GMM solver defaults already match;
deli accepts any `stats::optim()` method name.

### Derivative method default

The bread matrix is the derivative of the estimating functions. Python's
estimators default to `deriv_method='approx'`, SciPy's `approx_fprime`, a
forward finite difference. deli defaults to `deriv_method = "capprox"`, a central
finite difference, which is more accurate for a negligible cost.

Both libraries offer the same delicatessen-implemented finite differences under
the same names, plus forward-mode automatic differentiation:

| Method | Meaning | Python name | R name |
|---|---|---|---|
| SciPy forward difference | `scipy.optimize.approx_fprime` | `"approx"` | none |
| Forward difference | delicatessen implementation | `"fapprox"` | `"fapprox"` |
| Backward difference | delicatessen implementation | `"bapprox"` | `"bapprox"` |
| Central difference | delicatessen implementation | `"capprox"` | `"capprox"` |
| Automatic differentiation | forward mode | `"exact"` | `"exact"` |

deli does not expose SciPy's `"approx"`. It is a forward difference at the same
step, so its closest match is deli's `"fapprox"`.

To reproduce a default Python fit, match Python's defaults from R:

```r
# Reproduce a default Python MEstimator run
m <- MEstimator(psi, init = c(0, 0)) |>
  estimate(solver = "lm", deriv_method = "fapprox")
```

Because `"fapprox"` and SciPy's `"approx"` are both forward differences at the
same step `dx`, they agree to numerical-approximation error rather than
bit-for-bit. For a linear model the finite difference of a linear function is
exact, so the two coincide exactly; for nonlinear models expect agreement to the
finite-difference tolerance.

If instead you control both sides of the comparison and want exact agreement,
set both libraries to the same delicatessen-implemented method. For example,
leave deli at its `"capprox"` default and pass `deriv_method='capprox'` in
Python, with a matched solver. That combination reproduces the sandwich
components across languages to machine precision.

### `delta_method` transform argument and default

Python's `delta_method(theta, g, covariance, deriv_method='exact', dx)` takes
the transformation as `g` and differentiates it with automatic differentiation
by default. deli's `delta_method(object, transform, covariance,
deriv_method = "capprox", dx)` renames `g` to `transform` and defaults to a
central finite difference. deli also accepts a fitted estimator as its first
argument, in which case the covariance is taken from the fit. To reproduce
Python's default, pass `deriv_method = "exact"`:

```r
delta_method(m, transform = g, deriv_method = "exact")
```

### `subset` uses 1-based indices

The `subset` argument selects which parameters to solve for while holding the
rest at their initial values. Python indexes parameters from zero; R indexes
from one. A Python `subset=[0, 2]` becomes `subset = c(1, 3)` in R. Add one to
every Python index.

### `confidence_bands` draw count

The supremum-t confidence bands are computed by simulating from the estimated
sampling distribution. Python's estimator method uses `n_draws=1e6`; deli uses
`n_draws = 1e5` at every entry point. The smaller default runs about an order of
magnitude faster, and the Monte Carlo error it introduces is negligible relative
to the band width. To reproduce Python's estimator default exactly, set the
draw count:

```r
confidence_bands(m, n_draws = 1e6)
```

Both draw counts target the same population quantile and agree to within Monte
Carlo error. Note that seeded band values are still not bit-identical across
languages: `MASS::mvrnorm` and NumPy draw multivariate normals with different
samplers, an inherent difference in the random-number stream rather than in the
method.

### Robust loss Hampel thresholds

The Hampel loss has three thresholds. Python passes them as separate arguments:
`robust_loss_functions(residual, loss='hampel', k=8, a=2, b=4)`, where `a` is the
inner threshold, `b` the middle, and `k` the outer. deli collects them into a
single ordered vector `k = c(a, b, c)`:

```r
robust_loss_functions(residuals, loss = "hampel", k = c(2, 4, 8))
```

The same convention applies to `ee_mean_robust` and `ee_robust_regression`. For
the other losses, `k` is a single tuning constant in both libraries.

### Pharmacokinetic loss argument

`ee_emax` and `ee_loglogistic` accept an optional robust loss. Python names the
argument `robust`; deli names it `loss` for consistency with the other robust
estimating equations. As in Python, supplying a loss also requires its tuning
constant `k`.

### Survival time and event arguments

`ee_survival_model`, `ee_aft`, and `ee_plogit` take the observed time and the
event indicator. Python names them `t` and `delta`; deli names them `time` and
`event`, which reads more clearly and is consistent across the survival family
and the `plogit_predict` helper. deli additionally accepts an `offset` on
`ee_aft` and `ee_plogit`.

### Gompertz survival predictions

Python's `survival_predictions` branches only on the exponential distribution
and otherwise applies the Weibull survival and hazard formulas, so requesting
`distribution="gompertz"` there silently returns Weibull values. deli's
`ee_survival_model` supports the Gompertz distribution, so deli's
`survival_predictions()` implements the matching Gompertz measures,

$$S(t) = \exp\!\left(-\frac{\lambda}{\gamma}\left(e^{\gamma t} - 1\right)\right),
\qquad h(t) = \lambda e^{\gamma t},$$

and rejects any distribution outside `"exponential"`, `"weibull"`, and
`"gompertz"` rather than falling through to the Weibull branch.

### Finite-sample correction and inference

When `finite_correction` is set on the fit (for example `"HC1"`), both libraries
switch the confidence intervals and p-values from the standard normal to the
t-distribution with `n_obs - n_params` degrees of freedom. This behavior already
matches; no recipe is needed.

## Worked example

The following fits the same linear regression in both libraries and shows that
the R code reproduces the Python numbers exactly. Both blocks run when this
article is built, so the outputs below can be compared directly.

The Python fit, using default settings (`solver='lm'`, `deriv_method='approx'`):

```{python}
#| label: example-fit-python
import numpy as np
from delicatessen import MEstimator
from delicatessen.estimating_equations import ee_regression

d_x = [1.2, 0.3, 2.1, 1.7, 0.9, 2.5, 1.1, 0.4, 1.9, 2.2]
d_y = [3.4, 1.1, 5.2, 4.3, 2.6, 6.1, 3.0, 1.4, 4.8, 5.5]
X = np.asarray([[1, xi] for xi in d_x], dtype=float)
y = np.asarray(d_y, dtype=float)

def psi(theta):
    return ee_regression(theta, X=X, y=y, model="linear")

m = MEstimator(psi, init=[0.0, 0.0])
m.estimate()

print(m.theta)
print(np.sqrt(np.diag(m.variance)))
print(m.confidence_intervals())
```

The equivalent deli fit, matching Python's default solver and forward-difference
derivative:

```{r}
#| label: example-fit
d <- data.frame(
  x = c(1.2, 0.3, 2.1, 1.7, 0.9, 2.5, 1.1, 0.4, 1.9, 2.2),
  y = c(3.4, 1.1, 5.2, 4.3, 2.6, 6.1, 3.0, 1.4, 4.8, 5.5)
)
X <- cbind(1, d$x)

psi <- function(theta) {
  ee_regression(theta, X = X, y = d$y, model = "linear")
}

m <- MEstimator(psi, init = c(0, 0)) |>
  estimate(solver = "lm", deriv_method = "fapprox")

m@theta
sqrt(diag(m@variance))
confidence_intervals(m)
```

The point estimates, standard errors, and confidence intervals match the Python
output to the printed precision.

deli also offers a formula interface that builds the design matrix and initial
values for you. It is convenient R sugar with no Python counterpart, and it
produces the same fit:

```{r}
#| label: example-formula
m2 <- m_estimate(
  y ~ x,
  data = d,
  .ee = ee_regression,
  model = "linear",
  solver = "lm",
  deriv_method = "fapprox"
)
coef(m2)
```

## Reproducibility summary

- Point estimates, variances, and interval bounds reproduce Python to machine
  precision when you match the solver and set both libraries to the same
  delicatessen-implemented derivative method.
- Matching Python's default `"approx"` derivative with deli's `"fapprox"` agrees
  to finite-difference tolerance, and exactly for linear models.
- Simulation-based confidence bands agree to Monte Carlo error but are not
  bit-identical, because the two libraries use different multivariate normal
  samplers.
